---
title: "Deliberation and Learning: Evidence From Deliberative Polls"
author: "Robert C. Luskin, Gaurav Sood, and James S. Fishkin"
date: "Working draft"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    fig_caption: true
bibliography: references.bib
link-citations: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H")
options(knitr.graphics.rel_path = FALSE)
source("R/paper.R")
figure <- \(name) knitr::include_graphics(normalizePath(file.path("figs", paste0(name, ".pdf"))))
count <- \(x) prettyNum(x, big.mark = ",")
gains <- read_out("poll_gains.csv")
main_gains <- read_out("poll_gains_main.csv")
guessing <- read_out("guessing_gains.csv")
meta <- read_out("meta.csv")
meta_c <- read_out("meta_control.csv")
models <- read_out("models.csv")
effects <- read_out("control_effects.csv")
hetero <- read_out("control_heterogeneity.csv")
peers <- read_out("peer_effects_pooled.csv")
selected <- read_out("selection.csv")
assignment <- read_out("assignment_check.csv")
polls <- read_out("polls.csv")
item_catalog <- read_item_catalog()

a1r <- control_row(effects, "America in One Room 2019", "Attended vs uninvited control")
climate <- control_row(effects, "America in One Room: Climate 2021", "Attended vs uninvited control")
climate_t3 <- control_row(effects, "America in One Room: Climate 2021", "Attended vs control, one year later")
amr <- control_row(effects, "Antimicrobial Resistance 2024", "Attended vs randomized control")
ni <- control_row(effects, "Northern Ireland 2007", "T3 attendee vs control, no baseline")
amr_country <- \(country) control_row(effects, "Antimicrobial Resistance 2024", paste0("Attended vs randomized control: ", country))
mu <- meta_c[meta_c$parameter == "mu", ]
persist <- round(100 * climate_t3$estimate / climate$estimate)
het <- \(study, moderator) hetero[hetero$study == study & hetero$moderator == moderator, ]
peer <- \(p) peers[peers$peer == p, ]
term <- \(model, t) model_term(models, model, t)
size <- \(model, what) models[[what]][models$model == model][1]
```

\begin{abstract}
Among the great virtues ascribed to deliberation is learning. We ask how much participants in Deliberative Polls learn, what control groups tell us about those gains, and who learns. Respondent-level answers to `r nrow(item_catalog)` knowledge questions cover `r nrow(polls)` polls held from `r min(polls$year)` to `r max(polls$year)` on elections, energy, health, crime, climate, and other policy issues across Europe, Australia, the Americas, Africa, and Asia. In the `r size("core", "polls")` polls with verified discussion-group identifiers, `r count(size("core", "n"))` attendees in `r size("core", "groups")` groups supply paired item answers. Across these polls, the share of items answered correctly rose by `r meta_row(meta, "raw, pooled", "mu")`, or `r meta_row(meta, "guessing adjusted, pooled", "mu")` after adjustment for guessing. A broader `r nrow(gains)`-poll analysis appears in the appendix. Four polls also have control-group answers, although Northern Ireland's controls were interviewed only at T3. In the three control polls with common baseline interviews, held in seven countries, attendees scored `r coef_text(mu$median, 2)` standard deviations above controls after adjustment for initial knowledge (95\% interval `r coef_text(mu$lower, 2)` to `r coef_text(mu$upper, 2)`); in one poll followed for a year, `r persist`\% of that adjusted gap remained. Some control comparisons show larger adjusted gaps among those starting with less knowledge or without degrees. Groupmates' prior knowledge has no detectable adjusted association with participants' T2 scores, though group composition is not fully balanced on T1 knowledge.
\end{abstract}

# Introduction

Among the great virtues commonly ascribed to deliberation is learning. The participants learn about the topics under discussion and, to a lesser degree, incidentally, about related ones. They learn not only relevant facts but the arguments for and against particular alternatives and what other people unlike themselves, their circumstances, and their ways of thinking are like. This is helpful, both at the individual level, for discerning what policies most closely accord with one's values and interests and voting accordingly, and at the aggregate level, for forming "authentic" majorities---majorities that "know what they are doing" [@luskin2003]. How much citizens know shapes what they prefer and how they vote [@converse1964; @bartels1996; @althaus1998; @dellicarpini1996], even on foreign policy [@aldrich1989; @foyle2004], though information shortcuts can partly stand in for knowledge [@popkin1991; @lupia1994].

The expectation of learning follows directly from the nature of "deliberation," conventionally construed as discussion, but not just any discussion [@fishkin2005]. While definitions vary, deliberation is generally viewed as normatively elevated discussion---in which, among other things, the participants clarify and refine their preferences by open-mindedly weighing competing arguments. As part of that weighing, they absorb factual information and sift out misinformation.

Or so the theory runs. But how much do participants actually learn, is it the deliberation that teaches them, and who learns the most? We examine learning of relevant facts. This is not the only sort of learning involved, but it is the easiest to measure, and it is both important in itself and positively correlated with the rest.

Participants gain knowledge during Deliberative Polls. In three polls that surveyed a control group at baseline and follow-up---two in the United States and one across six middle-income countries---attendees scored `r coef_text(mu$median, 2)` standard deviations above controls after adjustment for initial knowledge. In the one poll re-interviewed a year later, `r persist`\% of that adjusted gap remained. Northern Ireland also interviewed a T3 control group, without baseline knowledge questions. Earlier studies of Deliberative Polls and other forums found that participants learn a lot [@luskin2002b; @barabas2004; @sturgis2005; @esterling2011; @theuwis2025], but most compared participants before and after, without a comparison group. We also examine how gains vary with starting knowledge and schooling. We also examine whether groupmates' initial knowledge is associated with learning and check how balanced group composition is at baseline.

# Deliberative Polling and Learning

Deliberative Polling [@fishkin1991; @fishkin1995; @luskin2002b] aims to capture fuller-information, more thoughtful opinions than those expressed in ordinary polls. In very brief, a Deliberative Poll interviews and assembles, as many as are willing of, a random sample to deliberate some policy or electoral choice, in alternating small group discussions and plenary question-and-answer sessions with panels of competing experts, and then re-measures the sample's opinions at the end. These latter are frequently different, both at the individual level and in the aggregate, from what they were before deliberating. The absolute net change is statistically significant nearly three-quarters of the time and averages nearly .1 on a 0--1 scale [@luskin2007b].

This design, admittedly a particular embodiment of deliberation, affords additional opportunities and incentives to learn. The participants are provided with carefully balanced briefing documents, rehearsing the arguments each side would make. There is usually a section of basic background information, and the arguments embed each side's pet facts---those most congenial to its arguments. In addition, the participants know from the moment they are first recruited that they will be attending the deliberation and that the deliberation (typically) will be an "Event," capital "e"---away from home, usually televised or taped for later broadcast, often attended by high-profile policy experts or policy-makers in the role of observers or panelists. Thus, motivated by a desire to "do well," many participants use the weeks between their recruitment and the event itself to prepare---not only by reading the briefing materials but by paying heightened attention to stories about the topic in the media, by talking more with friends, family, and coworkers about it, and even, in some cases, by researching it in the library or online.

So if knowledge, like its close relative "sophistication," depends on the confluence of cognitive ability, motivation, and exposure to relevant information [@luskin1990; @luskin2002a], the deliberation in Deliberative Polling stands to increase two of those three. There is not much to be done, in adulthood, about ability---and in fact we should want the participants to be representative in this, as in all other respects. But the treatment can and does affect motivation and exposure. Judging from past polls, that is enough to produce considerable, sometimes dramatic learning [@luskin2002b; @list2013].

The online design is intended to be as similar as possible to the face-to-face one, given the difference in mode. The deliberation is by voice rather than text, so as to capture as much as possible of deliberation's social elements. The participants still "meet" in small groups, still led by trained moderators, and still question panels of policy experts and policy makers. They still receive briefing materials. They are still a representative sample---sometimes randomly sampled offline, with "nonusers" receiving free access, including free equipment, to get them online, sometimes drawn from online panels and matched to approximate a random sample. On the other hand, deliberating by voice, even by video, online can never capture the full richness and complexity of in-person interactions; the discussions are interrupted, spread over a period of weeks, with much greater possibility of real-world intrusion, and the assignment to small groups is perforce constrained by possible meeting times. All these differences may act to weaken the online treatment's effects [@luskin2006; @iyengar2005].

# Data

We include polls with respondent-level knowledge-item answers at both T1 and T2. Each poll appears once in the upstream registry and in Appendix \ref{app:polls}. The accompanying dp-data repository holds the item definitions, scored responses, respondent links, group assignments, and poll metadata. We join its participant and score tables into one paired-attendee panel, selecting one source for each poll where releases overlap. We average correctness over the questions fielded to a respondent to obtain the proportion correct before (T1) and after (T2) deliberating. The main attendee analyses use the `r size("core", "polls")` polls with verified small-group identifiers; `r nrow(gains) - size("core", "polls")` others enter the broader appendix estimate. Control comparisons use the same canonical participant and score tables, restricting to the studies with observed control answers. Appendix \ref{app:items} gives the questions and scoring rules.

The control groups differ in how they were formed. America in One Room and its 2021 climate successor interviewed a separate random sample that was never invited [@fishkin2021]. The poll on antimicrobial resistance, held online in Brazil, Colombia, India, Indonesia, Nigeria and Tanzania, randomized panel members after the first interview; the release contains the controls and the invitees who attended, but not the invitees who stayed home [@mendelson2026]. Northern Ireland interviewed 150 controls at T3, when 93 participants also returned; the controls have no T1 knowledge answers, so only a T3 cross-sectional comparison is possible. These are comparisons of attendees with controls, so selection into attendance remains possible. Poll metadata and reconstructed respondent answers come from dp-data, whose exports resolve duplicate Primaries rows and mark implausible ages missing.

For the three polls with common baselines, we compare attendees with controls using both a T2 regression adjusted for T1 and a difference in T2-minus-T1 gains. The Northern Ireland contrast compares T3 scores; its standard error clusters attendees by discussion group and treats controls as separate clusters.

# Measuring Learning

We measure knowledge by closed-ended knowledge questions, mostly multiple choice, in some cases extracted from party or candidate placements, and in a few cases true-false [@luskin1987; @luskin2005; @price1999; @mondak1999; @mondak2001]. Appendix \ref{app:items} lists the items. Correct responses are scored as 1, incorrect responses and "don't knows" as 0 [@luskin2011], and these scores are then averaged across items to give the proportion answered correctly.

Table \ref{tab:models} models T2 knowledge conditional on T1 knowledge. Writing the outcome as $K_2-K_1$ while retaining $K_1$ gives the same fit and the same coefficients for every other regressor; only the coefficient on $K_1$ falls by one. An unadjusted change-score model instead fixes the coefficient on $K_1$ in the T2 equation at one. It asks whether covariates predict the raw score difference, whereas Table \ref{tab:models} compares participants with the same measured starting score. The approaches can disagree when groups differ at baseline [@allison1990; @vanbreukelen2006]. The observed score has a ceiling, and guessing and measurement error complicate its relationship to learning of the underlying facts [@luskin2016]. We therefore treat the T1 coefficient as an association between measured scores, not an estimate of how prior knowledge causes learning.

Closed-ended items invite guessing, which can hide learning in the difference between observed scores. We fit the item-level latent class model of @cor2016 to all `r nrow(gains)` polls. It estimates the share who learned each question by allowing a correct answer to come from either knowledge or guessing, assuming knowledge is not lost between waves. We average its estimates across each poll's questions and bootstrap respondents to measure sampling error. In the main `r size("core", "polls")`-poll sample, the pooled guessing-adjusted gain is `r meta_row(meta, "guessing adjusted, pooled", "mu")`, compared with the observed gain of `r meta_row(meta, "raw, pooled", "mu")`. The two pooled results appear in Figure \ref{fig:learning}; Appendix Table \ref{tab:meta} gives the broader `r nrow(gains)`-poll estimates. Table \ref{tab:models} retains respondent-level observed scores. The available joint person-level model in `guess` assigns one knowledge trajectory to each respondent across all questions, ruling out knowing one fact while learning another; its posterior is not a suitable outcome for this regression.

# How Much Do Participants Learn, and Does Participating Cause It?

Figure \ref{fig:learning} shows the observed attendee gain in the `r size("core", "polls")` polls with verified group identifiers and both pooled observed and guessing-adjusted gains. Its second panel shows control comparisons: T1-adjusted T2 differences in three polls, their pooled result, the climate poll's T3 follow-up, and Northern Ireland's T3-only contrast. The antimicrobial-resistance control study has no group identifier, so it enters that comparison but not the main attendee-gain pool. Northern Ireland uses the control group's T3 standard deviation because its controls have no baseline knowledge answers.

Every poll shows learning. In the main sample, raw gains range from `r coef_text(min(main_gains$raw))` to `r coef_text(max(main_gains$raw))` of the items; pooled, the proportion correct rose by `r meta_row(meta, "raw, pooled", "mu")`, or `r meta_row(meta, "raw SD, pooled", "mu", 2)` standard deviations. The guessing-adjusted pooled estimate is `r meta_row(meta, "guessing adjusted SD, pooled", "mu", 2)` on the same T1 standard-deviation scale. The pooling is a Bayesian random-effects meta-analysis: each poll's mean gain is its true gain plus sampling error, true gains are normally distributed around an overall mean, the prior on the mean is flat and that on the between-poll standard deviation half-normal with scale .1 [@rover2020]. Face-to-face polls produced about twice the observed gain of online ones (`r meta_row(meta, "raw, by mode", "face_to_face")` against `r meta_row(meta, "raw, by mode", "online")`; Appendix Table \ref{tab:meta}).

A pre-post gain alone does not show that deliberation taught anything: participants take the same quiz twice and live through weeks of news. The control groups provide a benchmark. Adjusting for T1 knowledge, attendees scored `r est_se(a1r$estimate, a1r$std_error)` higher in proportion correct in America in One Room, `r est_se(climate$estimate, climate$std_error)` higher in the 2021 climate poll, and `r est_se(amr$estimate, amr$std_error)` higher in the antimicrobial-resistance poll, where the country estimates range from `r coef_text(amr_country("Tanzania")$estimate, 2)` in Tanzania to `r coef_text(amr_country("Nigeria")$estimate, 2)` in Nigeria. Pooled, the adjusted difference is `r interval(mu$median, mu$lower, mu$upper, 2)` standard deviations. A year after the climate poll, attendees still scored `r coef_text(climate_t3$estimate)` higher than controls (s.e. `r coef_text(climate_t3$std_error)`), `r persist`\% of the immediate adjusted difference. The controls gained `r coef_text(a1r$control_gain, 2)`, `r coef_text(climate$control_gain, 2)` and `r coef_text(amr$control_gain, 2)` of the items in the three studies.

The antimicrobial-resistance poll randomized invitations, but its treated group here is the attendees, who were 23 to 92\% of invitees depending on the country. Attendees gained `r coef_text(amr$treated_gain)` of the items and randomized controls `r coef_text(amr$control_gain)`; the difference in changes is `r est_se(amr$gain_difference, amr$gain_difference_se)`. This comparison includes selection into attendance. The two American polls likewise compare attendees with separately drawn, uninvited controls. In America in One Room, attendees answered `r coef_text(selected$k1[1])` of T1 items correctly against `r coef_text(selected$k1[2])` for invitees who stayed home; in the climate poll the figures were `r coef_text(selected$k1[3])` and `r coef_text(selected$k1[4])`. Their differences in changes are `r est_se(a1r$gain_difference, a1r$gain_difference_se)` and `r est_se(climate$gain_difference, climate$gain_difference_se)`, respectively. Adjusting for T1 and comparing changes impose different restrictions on baseline differences; neither resolves selection into attendance. In Northern Ireland, returning attendees scored `r est_se(ni$estimate, ni$std_error)` more of the T3 items correctly than controls. This is an unadjusted cross-sectional difference: the controls had no T1 knowledge battery, and event attendance was self-selected.

```{r learning, fig.cap = "Knowledge gains and attendee-control differences, with 95\\% intervals. A: observed attendee T2 minus T1 gain in 28 polls with verified group identifiers and pooled observed and guessing-adjusted learning, scaled by each poll's T1 standard deviation. B: T1-adjusted T2 attendee-control differences in three polls and their pooled estimate, plus the climate T3 follow-up; these use the control group's T1 standard deviation. Northern Ireland is an unadjusted T3 contrast, scaled by the control group's T3 standard deviation and excluded from the pooled control estimate. Antimicrobial Resistance has no group identifier and enters only panel B here. Pooled estimates use Bayesian random-effects meta-analysis.", fig.pos = "tbp", out.width = "100%"}
figure("learning")
```

# Who Learns?

## Regressors and hypotheses

The main model uses variables built consistently from the paired item answers and group identifiers: initial knowledge, groupmates' initial knowledge, group size, mode, and poll mean initial knowledge. Appendix \ref{app:further} develops additional hypotheses for education, demographics, attitudes, and briefing reading and reports the corresponding historical-poll models.

*Initial knowledge* ($K_1$). The information-rich tend to get information-richer. The more one knows going in, the more one tends to learn and thus know going out. Absent ceiling effects, the coefficient should be $> 1$; given them, it may be $< 1$. Prior knowledge on a particular topic improves people's ability to comprehend new information, enabling them to extract the central elements of a message and draw appropriate inferences efficiently [@eckhardt1991; @recht1988]. Prior knowledge also enhances people's ability to store new information on that topic and retrieve it later [@cooke1993; @fiske1990; @hambrick2003; @mcgraw1990; @recht1988; @schneider1993]. So the less political information individuals have stored in memory, the more difficult it is for them to acquire new information.

*Mode*. Online deliberation, with its thinner exposure to other participants, may be, either intrinsically or at least in its current instantiations, a somewhat etiolated version of the face-to-face treatment [@luskin2006]. It is therefore plausible that people learn less from online deliberations.

*Observed group size*. We count group members with paired item answers. This is a lower bound on actual group size when a member lacks one of the interviews. Larger groups may offer more perspectives but give each participant less time to speak [@latane1979].

*Group knowledge*. Participants may learn more from the other members of their small group when the latter know more. The other members' mean T1 score measures their overall knowledge. Their mean correctness on the particular items a participant missed at T1 measures knowledge that could fill that participant's gaps. We add this targeted measure alongside the overall group mean in a supplementary model. Participants who missed no T1 items receive zero on the targeted measure and a separate indicator for having no missed items.

*Issue salience*. More salient issues, to which many participants have already given some attention, exhibit decidedly less net attitude change [@luskin2007b] and fall shorter of single-peakedness [@list2013]. We suspect they also, and relatedly, show less learning: there is less left to learn. We proxy salience by the participants' mean T1 knowledge.

## Model

The nested structure of the data, with individuals clustered in small groups clustered in polls, suggests a hierarchical linear model [@raudenbush2002] with random intercepts by small group and poll:
$$K_{igd2} = \beta_0 + \alpha_g + \gamma_d + \mathbf{x}_{igd}'\boldsymbol{\beta} + u_{igd},$$
where $i$, $g$ and $d$ index individuals, small groups and polls; $\alpha_g \sim N(0, \sigma^2_\alpha)$ and $\gamma_d \sim N(0, \sigma^2_\gamma)$ are unmodeled parts of the intercept; $\mathbf{x}_{igd}$ holds initial knowledge, the other group members' mean T1 knowledge, group size, online mode, and the poll mean T1 score; and $u_{igd} \sim N(0, \sigma^2_u)$ is a disturbance. We estimate it by maximum likelihood [@bates2015]. The main model uses `r count(size("core", "n"))` participants in `r size("core", "groups")` groups in `r size("core", "polls")` polls. Of the `r nrow(gains)` polls with paired item answers, Denmark and Vermont lack verified group rosters, while the antimicrobial-resistance release lacks small-group IDs. Appendix Table \ref{tab:historical-models} reports a richer specification on the `r size("historical", "polls")` historical polls where additional covariates can be measured.

## Who knows more after deliberation?

```{r models}
table_1 <- model_table(models, "core", "Group-linked sample")
separators <- rep("", nrow(table_1))
summary_start <- match("Participants", table_1[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(table_1) - 1L)] <- "\\addlinespace[0.25em]"
knitr::kable(
  table_1, format = "latex", booktabs = TRUE, position = "H", align = "lr",
  linesep = separators,
  caption = sprintf("T2 proportion correct conditional on T1 in %s polls with verified small groups. Maximum likelihood coefficients (standard errors).", size("core", "polls"))
)
```

\noindent{\footnotesize\emph{Notes.} Scores come from respondent-level item answers in the canonical dp-data panel. Groupmates' mean T1 excludes the respondent herself; both that mean and group size use members with paired item answers. The model has poll and small-group random intercepts. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts. The richer historical-poll and reading-report models appear in Appendix Table \ref{tab:historical-models}.\par}

Table \ref{tab:models} shows that participants with higher T1 scores also score higher at T2 (`r term("core", "k1")`), but the coefficient below one corresponds to smaller observed gains at higher T1 scores. The score ceiling and measurement error complicate an interpretation as differences in actual learning. The coefficient on groupmates' mean T1 is `r term("core", "group_k1")`; its interval includes zero. Participants in online polls score lower at T2 (`r term("core", "online")`), conditional on T1, and those in polls whose participants start out knowing more score higher (`r term("core", "poll_k1")`). In the historical-poll model, a bachelor's degree is associated with a T2 advantage at the lowest baseline score (`r term("historical", "educationBA or more")`), which narrows as T1 knowledge rises (`r term("historical", "k1:educationBA or more")` per unit). In the nine polls with a reading report, those who said they read more of the briefing materials scored higher at T2 (`r term("briefing", "read_briefing")`), though reading is itself a choice (Appendix Table \ref{tab:historical-models}).

## Whom participation helps most

The control-group studies also let us ask whether the adjusted attendee-control difference varies across participants. Adding the product of attendance and centred T1 knowledge, the difference is smaller for those who start knowing more, by `r est_se(het("America in One Room 2019", "k1")$interaction, het("America in One Room 2019", "k1")$interaction_se, 2)` per unit of T1 knowledge in America in One Room and `r est_se(het("America in One Room: Climate 2021", "k1")$interaction, het("America in One Room: Climate 2021", "k1")$interaction_se, 2)` in the climate poll; the antimicrobial-resistance estimate is smaller and less precise (Appendix Table \ref{tab:hetero}). Part of this pattern may be the ceiling: those who start high have less room to show a gain. The adjusted difference is also larger for those without a bachelor's degree in the climate and antimicrobial-resistance polls (`r est_se(het("America in One Room: Climate 2021", "ba")$interaction, het("America in One Room: Climate 2021", "ba")$interaction_se)` and `r est_se(het("Antimicrobial Resistance 2024", "ba")$interaction, het("Antimicrobial Resistance 2024", "ba")$interaction_se)`), though not in America in One Room. These comparisons can also reflect selection into attendance.

## Do small groups matter?

If participants learn from each other, those placed with more knowledgeable groupmates should learn more. A causal interpretation requires groupmates' characteristics to be unrelated to one's own, as random assignment would imply. We test this by regressing each participant's characteristic on the mean among the other members of the group, controlling for the mean among the other members of the poll, without which the test is biased toward finding negative sorting [@guryan2009]. Across the 28 group-linked polls, the T1 knowledge balance coefficient is `r est_se(assignment$estimate[1], assignment$std_error[1])` and the gender coefficient is `r est_se(assignment$estimate[2], assignment$std_error[2])`. In the 21 historical polls with the needed covariates, the age coefficient is `r est_se(assignment$estimate[3], assignment$std_error[3])` and the bachelor's-degree coefficient is `r est_se(assignment$estimate[4], assignment$std_error[4])`. The positive T1 balance coefficient in the expanded sample indicates some sorting of more knowledgeable participants into more knowledgeable groups. We therefore describe the peer estimates as associations, without a causal interpretation.

The group comparisons remain informative descriptively. Within each of `r peer("k1")$polls` polls with group identifiers, we regress T2 knowledge on the other members' mean T1 knowledge, controlling for one's own and the poll's, and pool the estimates (Figure \ref{fig:peers}). The pooled association is `r interval(peer("k1")$median, peer("k1")$lower, peer("k1")$upper)`: its upper interval endpoint corresponds to `r round(10 * peer("k1")$upper, 1)` points more on T2 for a ten-point difference in groupmates' T1 scores. The share of women in the group has no detectable association either (`r interval(peer("female")$median, peer("female")$lower, peer("female")$upper)`). In the historical-poll sample, groupmates' knowledge on items each participant missed also has no detectable association when included alongside the overall peer mean (Appendix Table \ref{tab:historical-models}). Group knowledge measured after the deliberation could show a spurious association: groupmates' T2 answers reflect the experience they shared with the participant [@manski1993].

```{r peers, fig.cap = "Association between the other group members' mean T1 knowledge and own T2 knowledge, by poll, with 95\\% intervals. Each estimate controls for own T1 knowledge and the poll's leave-one-out mean, with standard errors clustered by small group; the pooled estimate is from a Bayesian random-effects meta-analysis. Baseline knowledge sorting makes these associations descriptive.", fig.pos = "tbp", out.width = "100%"}
figure("peer_effects")
```

# Conclusion

Participants in Deliberative Polls gain factual knowledge, and attendees in the three polls with common baseline interviews score higher than controls after adjusting for T1. Northern Ireland's T3-only comparison points in the same direction but lacks baseline knowledge for controls. Much of the adjusted gap lasts in the one poll with a year-later interview. Within the participant samples, observed gains are larger among those who start out knowing less. Groupmates' initial knowledge has no detectable adjusted association with T2 scores in the expanded group-linked sample; baseline sorting limits a causal reading.

The attendee-control comparisons can reflect selection into attendance, even where invitations were randomized. The knowledge items are few and chosen by the organizers, so the gains concern facts the event emphasized.

# References {-}

<div id="refs"></div>

\clearpage
\appendix
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

# The Deliberative Polls {#app:polls}

All `r nrow(polls)` polls with respondent-level knowledge items enter the broader before-and-after analysis. The `r sum(polls$group_model)` polls with verified discussion-group identifiers enter the main attendee analysis and Table \ref{tab:models}. Four provide control comparisons, including Northern Ireland's T3-only contrast; Antimicrobial Resistance has no group identifier.

```{r polls, results = "asis"}
polls |>
  dplyr::transmute(
    poll, year,
    mode = dplyr::if_else(mode == "Online", "Online", "In person"),
    group_model = dplyr::if_else(group_model, "Yes", "No"),
    control = dplyr::if_else(control_group, "Yes", "No")
  ) |>
  purrr::set_names(c("Poll", "Year", "Mode", "Group model", "Control")) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lclcc",
    caption = "Polls with respondent-level knowledge-item answers at both waves."
  ) |>
  print()
```

\clearpage
# Further results {#app:further}

```{r broad-gains}
gains |>
  dplyr::left_join(
    dplyr::select(guessing, "poll_id", "adjusted", "adjusted_se"),
    by = "poll_id", relationship = "one-to-one"
  ) |>
  dplyr::arrange(pollname) |>
  dplyr::transmute(
    Poll = pollname,
    `Observed gain` = est_se(raw, raw_se),
    `Guessing adjusted` = est_se(adjusted, adjusted_se)
  ) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lcc",
    caption = "T2 minus T1 attendee knowledge gains across all 31 item-linked polls, in proportions correct. Estimates with bootstrap standard errors for guessing adjustment."
  )
```

```{r meta}
meta |>
  dplyr::filter(model %in% c(
    "raw, pooled", "raw, by mode", "guessing adjusted, pooled",
    "raw, all", "guessing adjusted, all"
  )) |>
  dplyr::mutate(
    parameter = dplyr::recode(parameter, mu = "Mean gain", face_to_face = "Face-to-face", online = "Online", tau = "Between-poll SD"),
    model = dplyr::recode(
      model, "raw, pooled" = "Observed pooled",
      "raw, by mode" = "Observed by mode",
      "guessing adjusted, pooled" = "Guessing adjusted pooled",
      "raw, all" = "Observed all 31",
      "guessing adjusted, all" = "Guessing adjusted all 31"
    ),
    dplyr::across(c(median, lower, upper), \(x) coef_text(x))
  ) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, position = "H", align = "llrrr", linesep = "",
    col.names = c("Model", "Parameter", "Median", "2.5%", "97.5%"),
    caption = "Bayesian random-effects meta-analysis of observed and guessing-adjusted gains in the 28-poll main sample and all 31 item-linked polls, in proportions of knowledge items"
  )
```

```{r hetero}
hetero |>
  dplyr::mutate(
    moderator = dplyr::recode(moderator, k1 = "T1 knowledge (centred)", ba = "Bachelor's degree"),
    interaction = est_se(interaction, interaction_se),
    n = count(n)
  ) |>
  dplyr::select(study, moderator, interaction, n) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, position = "H", linesep = "",
    col.names = c("Study", "Moderator", "Treatment x moderator", "n"),
    caption = "Heterogeneous effects of participation. Each row adds the product of treatment and the moderator to the regression of T2 on treatment and T1 knowledge."
  )
```

## Additional historical-poll covariates

*Education*. Education, qua years of schooling, is distinct from, and only moderately correlated with, political knowledge. Controlling for cognitive ability, with which it is more strongly correlated, its effect on knowledge appears negligible [@luskin1990]. But absent a separate measure of cognitive ability, it can roughly proxy the latter. Naturally, we expect it to have a positive effect. There may, however, be some mutual substitution with initial knowledge. People who know enough already may not need to be as bright to learn a lot, and people who are bright enough may not need to know as much to do so. We measure education as comparably as possible across polls in three categories: less than high school graduation, high school graduation, and a bachelor's degree or more.

*Gender and age*. These demographic variables exhibit real-world political knowledge gaps, with men knowing more than women and older people more than younger ones [@dellicarpini1996]. Age is measured in decades and gender as a dummy scored 1 for women.

*Proportion of women*. Whether from greater comfort, more widely shared cognitive or communication style, or lower stereotype threat [@steele1997], women may learn more in small groups containing more women. We calculate the proportion excluding the participant herself.

*Heterogeneity of opinion*. In the real world, people often know congenial but not uncongenial facts---those commonly cited by those sharing their views but not those commonly cited by those holding opposing views. Part of deliberation's effect on learning may consist in getting people to note and accept (if, in many cases, still argue around) uncongenial facts. This can be expected to occur mainly on site. Participants learn a great deal in anticipation of the event, but the learning at that stage is mainly of congenial facts [@hansen2004]. On site, the participants find it harder to avoid confronting the uncongenial ones. This may happen more, the more opposing points of view are represented and argued in the small group---the more attitudinally heterogeneous the group is. We gauge this as the generalized variance, the determinant of the covariance matrix, of the group's policy attitudes [@anderson2003].

*Attitude extremity*. Participants starting with more extreme views may have shuttered out more uncongenial facts, leaving them with more ground for deliberation to make up. We measure extremity as the mean, across attitude indices $j$, of $|A_{ij} - .5|$, where $A_{ij}$ is the $i$th respondent's score on the $j$th index.

*Reading the briefing materials*. One of the main information vectors is the briefing document sent to participants shortly after recruitment, normally well in advance of the on-site deliberations. It provides some background facts and lays out the major arguments for and against given policy options, which embed additional facts congenial to each argument. Respondent-linked reading reports are available for nine polls: National Issues Convention, National Issues Convention 2, Tomorrow's Europe, San Mateo, Australia, Central Power & Light, Southwestern Electric Power, West Texas Utilities, and Europolis. We score their ordered categories from zero to one using the source-specific response scales and analyze them in a separate model. Comparable reports have not been recovered for the other polls in the group-linked analysis.

## Historical-poll models

```{r historical-models}
historical_models <- model_table(
  models, c("historical", "briefing"),
  c("Historical polls", "Reading reports")
)
knitr::kable(
  historical_models, format = "latex", booktabs = TRUE,
  longtable = TRUE, align = "lrr", linesep = "",
  caption = "Supplementary historical-poll models of T2 proportion correct conditional on T1. Maximum likelihood coefficients (standard errors)."
)
```

\noindent{\footnotesize\emph{Notes.} The historical-poll model includes poll and small-group random intercepts; the reading-reports model uses poll fixed effects and small-group random intercepts. Adding groupmates' T1 correctness on questions the respondent missed to the historical-poll model gives `r term("items", "group_k1_items")` among `r count(size("items", "n"))` participants. Respondents with no missed T1 items receive zero on this measure and a separate indicator (`r term("items", "no_missed_items")`); none is removed for lack of an opportunity to learn. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts.\par}

# Knowledge items {#app:items}

The catalog contains `r nrow(item_catalog)` distinct baseline questions in `r dplyr::n_distinct(item_catalog$poll_id)` polls. The item definitions and source references are maintained in `dp-data/metadata/items.csv`; this appendix is generated from that file. Wording is paraphrased where only source labels survive. Choices are transcribed where recoverable; Northern Ireland's archived choice labels are truncated, while the retained climate and antimicrobial-resistance reports omit them. A code in brackets identifies the scored baseline answer, and notes describe open-response coding.

```{r items, results='asis'}
cat(item_appendix_markdown())
```
