---
title: 'Deliberation and Learning: Evidence From Deliberative Polls\thanks{Data and replication materials: \url{https://github.com/soodoku/dp-learning}.}'
author: "Robert C. Luskin, Gaurav Sood, and James S. Fishkin"
date: "Working draft"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    fig_caption: true
bibliography: [references.bib, data-version.bib]
link-citations: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{placeins}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H")
options(knitr.graphics.rel_path = FALSE)
source("R/paper.R")
figure <- \(name) knitr::include_graphics(normalizePath(file.path("figs", paste0(name, ".pdf"))))
count <- \(x) prettyNum(x, big.mark = ",")
gains <- read_out("poll_gains.csv")
main_gains <- read_out("poll_gains_main.csv")
guessing <- read_out("guessing_gains.csv")
meta <- read_out("meta.csv")
controlled_learning <- read_out("control_learning.csv")
provenance <- jsonlite::read_json("tabs/provenance.json", simplifyVector = TRUE)
models <- read_out("models.csv")
effects <- read_out("control_effects.csv")
hetero <- read_out("control_heterogeneity.csv")
peers <- read_out("peer_effects_pooled.csv")
selected <- read_out("selection.csv")
assignment <- read_out("assignment_check.csv")
polls <- read_out("polls.csv")
item_catalog <- read_item_catalog()

a1r <- control_row(effects, "America in One Room 2019", "Attended vs uninvited control")
climate <- control_row(effects, "America in One Room: Climate 2021", "Attended vs uninvited control")
climate_t3 <- control_row(effects, "America in One Room: Climate 2021", "Attended vs control, one year later")
amr <- control_row(effects, "Antimicrobial Resistance 2024", "Attended vs randomized control")
ni <- control_row(effects, "Northern Ireland 2007", "T3 attendee vs control, no baseline")
amr_country <- \(country) control_row(effects, "Antimicrobial Resistance 2024", paste0("Attended vs randomized control: ", country))
main_polls <- polls[polls$group_model, ]
pp <- function(x) num(100 * x, 1)
pooled <- function(model) meta[meta$model == model & meta$parameter == "mu", ]
control_gain <- function(id, metric) controlled_learning[controlled_learning$poll_id == id & controlled_learning$term == metric, ]
persist <- round(100 * climate_t3$estimate / climate$estimate)
het <- \(study, moderator) hetero[hetero$study == study & hetero$moderator == moderator, ]
peer <- \(p) peers[peers$peer == p, ]
term <- \(model, t) model_term(models, model, t)
size <- \(model, what) models[[what]][models$model == model][1]
```

\begin{abstract}
We study factual learning among `r count(size("core", "n"))` participants in `r size("core", "groups")` discussion groups across `r size("core", "polls")` Deliberative Polls held between `r min(main_polls$year)` and `r max(main_polls$year)`. The polls address elections, energy, health, crime, climate, and other public issues in Europe, Australia, China, and the United States. Averaging equally across polls, knowledge scores increased by `r pp(pooled("raw, pooled")$estimate)` percentage points, and the average increase relative to each poll's initial score was `r pp(pooled("relative, pooled")$estimate)`\%. Adjustment for guessing yields estimated learning of `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points. In two polls with before-and-after control interviews, attendees' gains exceeded controls' gains by `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` points; guessing-adjusted differences were `r pp(control_gain("america-in-one-room-2019", "adjusted")$estimate)` and `r pp(control_gain("a1r-climate-2021", "adjusted")$estimate)` points. We assess uncertainty by resampling polls and their discussion groups. Peer associations and comparisons with controls are descriptive: group assignment and attendance need not be independent of prior knowledge.
\end{abstract}

# Introduction

Among the great virtues commonly ascribed to deliberation is learning. The participants learn about the topics under discussion and, to a lesser degree, incidentally, about related ones. They learn not only relevant facts but the arguments for and against particular alternatives and what other people unlike themselves, their circumstances, and their ways of thinking are like. This is helpful, both at the individual level, for discerning what policies most closely accord with one's values and interests and voting accordingly, and at the aggregate level, for forming "authentic" majorities---majorities that "know what they are doing" [@luskin2003]. How much citizens know shapes what they prefer and how they vote [@converse1964; @bartels1996; @althaus1998; @dellicarpini1996], even on foreign policy [@aldrich1989; @foyle2004], though information shortcuts can partly stand in for knowledge [@popkin1991; @lupia1994].

The expectation of learning follows directly from the nature of "deliberation," conventionally construed as discussion, but not just any discussion [@fishkin2005]. While definitions vary, deliberation is generally viewed as normatively elevated discussion---in which, among other things, the participants clarify and refine their preferences by open-mindedly weighing competing arguments. As part of that weighing, they absorb factual information and sift out misinformation.

Or so the theory runs. But how much do participants actually learn, is it the deliberation that teaches them, and who learns the most? We examine learning of relevant facts. This is not the only sort of learning involved, but it is the easiest to measure, and it is both important in itself and positively correlated with the rest.

We compare the knowledge participants bring to deliberation with what they know afterward, using one sample throughout the main analysis. Two of these polls also interviewed people who were not invited, allowing us to compare changes among attendees and controls. Earlier studies document learning in Deliberative Polls and other forums [@luskin2002b; @barabas2004; @sturgis2005; @esterling2011; @theuwis2025]. We report gains in both absolute and relative terms, adjust for guessing, and examine whether participants learn more alongside more knowledgeable groupmates.

# Deliberative Polling and Learning

Deliberative Polling [@fishkin1991; @fishkin1995; @luskin2002b] aims to capture fuller-information, more thoughtful opinions than those expressed in ordinary polls. In very brief, a Deliberative Poll interviews and assembles, as many as are willing of, a random sample to deliberate some policy or electoral choice, in alternating small group discussions and plenary question-and-answer sessions with panels of competing experts, and then re-measures the sample's opinions at the end. These latter are frequently different, both at the individual level and in the aggregate, from what they were before deliberating. The absolute net change is statistically significant nearly three-quarters of the time and averages nearly .1 on a 0--1 scale [@luskin2007b].

This design, admittedly a particular embodiment of deliberation, affords additional opportunities and incentives to learn. The participants are provided with carefully balanced briefing documents, rehearsing the arguments each side would make. There is usually a section of basic background information, and the arguments embed each side's pet facts---those most congenial to its arguments. In addition, the participants know from the moment they are first recruited that they will be attending the deliberation and that the deliberation (typically) will be an "Event," capital "e"---away from home, usually televised or taped for later broadcast, often attended by high-profile policy experts or policy-makers in the role of observers or panelists. Thus, motivated by a desire to "do well," many participants use the weeks between their recruitment and the event itself to prepare---not only by reading the briefing materials but by paying heightened attention to stories about the topic in the media, by talking more with friends, family, and coworkers about it, and even, in some cases, by researching it in the library or online.

So if knowledge, like its close relative "sophistication," depends on the confluence of cognitive ability, motivation, and exposure to relevant information [@luskin1990; @luskin2002a], the deliberation in Deliberative Polling stands to increase two of those three. There is not much to be done, in adulthood, about ability---and in fact we should want the participants to be representative in this, as in all other respects. But the treatment can and does affect motivation and exposure. Judging from past polls, that is enough to produce considerable, sometimes dramatic learning [@luskin2002b; @list2013].

The online design is intended to be as similar as possible to the face-to-face one, given the difference in mode. The deliberation is by voice rather than text, so as to capture as much as possible of deliberation's social elements. The participants still "meet" in small groups, still led by trained moderators, and still question panels of policy experts and policy makers. They still receive briefing materials. They are still a representative sample---sometimes randomly sampled offline, with "nonusers" receiving free access, including free equipment, to get them online, sometimes drawn from online panels and matched to approximate a random sample. On the other hand, deliberating by voice, even by video, online can never capture the full richness and complexity of in-person interactions; the discussions are interrupted, spread over a period of weeks, with much greater possibility of real-world intrusion, and the assignment to small groups is perforce constrained by possible meeting times. All these differences may act to weaken the online treatment's effects [@luskin2006; @iyengar2005].

# Data

The main sample comprises `r size("core", "polls")` polls and `r count(size("core", "n"))` attendees with knowledge answers before and after deliberation. It supports both the learning estimates and the analysis of discussion groups. We use the versioned Deliberative Poll dataset [@dpdata], selecting one respondent record per poll and averaging item correctness into knowledge scores. Appendix \ref{app:items} gives the questions and scoring rules.^[Appendix \ref{app:polls} includes three additional polls with paired item answers but no usable discussion-group membership: Denmark, Vermont, and Antimicrobial Resistance. It also retains `r gains$respondents[gains$poll_id == "btp-online-primaries-2004"] - main_gains$respondents[main_gains$poll_id == "btp-online-primaries-2004"]` additional online-primary respondents without group assignments. These observations enter the broader appendix analysis.]

America in One Room and the climate poll interviewed separate samples who were never invited to deliberate [@fishkin2021]. We compare changes in their scores with changes among attendees in the main sample, using the same questions in both arms. These comparisons may reflect selection into attendance as well as deliberation. Additional control comparisons and follow-ups appear in Appendix \ref{app:controls}.

# Measuring Learning

We measure knowledge by closed-ended knowledge questions, mostly multiple choice, in some cases extracted from party or candidate placements, and in a few cases true-false [@luskin1987; @luskin2005; @price1999; @mondak1999; @mondak2001]. Appendix \ref{app:items} lists the items. Correct responses are scored as 1, incorrect responses and "don't knows" as 0 [@luskin2011], and these scores are then averaged across items to give the proportion answered correctly.

Table \ref{tab:models} models T2 knowledge conditional on T1 knowledge. Writing the outcome as $K_2-K_1$ while retaining $K_1$ gives the same fit and the same coefficients for every other regressor; only the coefficient on $K_1$ falls by one. An unadjusted change-score model instead fixes the coefficient on $K_1$ in the T2 equation at one. It asks whether covariates predict the raw score difference, whereas Table \ref{tab:models} compares participants with the same measured starting score. The approaches can disagree when groups differ at baseline [@allison1990; @vanbreukelen2006]. The observed score has a ceiling, and guessing and measurement error complicate its relationship to learning of the underlying facts [@luskin2016]. We therefore treat the T1 coefficient as an association between measured scores, not an estimate of how prior knowledge causes learning.

A correct answer can reflect knowledge or guessing. We apply the item model of @cor2016 separately to each poll's paired answers. Under the assumption that knowledge is not lost between the two interviews, it estimates the share who learned each fact. We average across questions to obtain learning on a 0--1 scale. In each control study, we fit this model separately to attendees and controls and subtract their estimated learning. Both arms use the same items; their guessing probabilities are estimated separately. Table \ref{tab:models} continues to use observed individual scores, since the item model estimates learning rates rather than an individual latent outcome.

We use one hierarchical bootstrap for uncertainty throughout the paper. For estimates across polls, we sample polls with replacement separately within online and in-person modes, retaining their observed counts, and then sample whole discussion groups within each selected poll. For an individual poll, we hold the poll fixed and sample its groups. Independently sampled controls are single-person units, sampled separately from attendees. Each sampled participant retains all items and interviews. We refit the models and recompute denominators in each of `r provenance$bootstrap_replicates` draws, using the 2.5th and 97.5th percentiles for 95\% confidence intervals. Thus the same resampling scheme propagates sampling and guessing-model estimation uncertainty. Stratifying by mode preserves support for the online coefficient in every draw. Inference treats each poll's question battery as fixed.

# How Much Do Participants Learn?

An absolute gain is the change in the percentage of items answered correctly. A relative gain divides that change by the poll's initial mean score: an increase from 40\% to 50\% correct is 10 percentage points, or 25\% of the initial score. We calculate relative gains from poll means rather than averaging individual ratios, which are undefined for people with zero initial scores. Pooled results give each poll equal weight; the relative result is the average of the poll-specific ratios, not the ratio of two pooled means.

Across the main sample, the average observed gain is `r pp(pooled("raw, pooled")$estimate)` percentage points (95\% interval `r pp(pooled("raw, pooled")$lower)` to `r pp(pooled("raw, pooled")$upper)`). The average relative increase is `r pp(pooled("relative, pooled")$estimate)`\% (95\% interval `r pp(pooled("relative, pooled")$lower)` to `r pp(pooled("relative, pooled")$upper)`\%). Estimated learning after adjustment for guessing averages `r pp(pooled("guessing adjusted, pooled")$estimate)` points (`r pp(pooled("guessing adjusted, pooled")$lower)` to `r pp(pooled("guessing adjusted, pooled")$upper)`). Appendix Table \ref{tab:meta} gives these summaries and the comparisons by mode.

A before-and-after gain can include learning from repeated questions or events outside deliberation. Figure \ref{fig:learning} therefore also compares attendees' gains with controls' gains. In America in One Room, the observed difference in gains is `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` points; adjustment for guessing gives `r pp(control_gain("america-in-one-room-2019", "adjusted")$estimate)` points. The climate poll's corresponding differences are `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` and `r pp(control_gain("a1r-climate-2021", "adjusted")$estimate)` points. These comparisons use precisely the same attendees and controls for observed and guessing-adjusted learning. They do not eliminate selection into attendance.

```{r learning, fig.cap = "Learning in the main sample, in percentage points, with 95\\% hierarchical-bootstrap confidence intervals. A: observed attendee gains in the main-sample polls, followed by equally weighted averages of observed and guessing-adjusted learning. B: attendee gain minus control gain in the two polls with common baseline and follow-up interviews, shown before and after guessing adjustment. Whole discussion groups are resampled within polls; independently sampled controls are resampled individually within their arm. Pooled intervals also resample polls within mode. Each draw refits the guessing model. The question batteries are treated as fixed.", fig.pos = "tbp", out.width = "100%"}
figure("learning")
```

# Who Learns?

## Regressors and hypotheses

The main model uses variables built consistently from the paired item answers and discussion groups: initial knowledge, groupmates' initial knowledge, group size, mode, and poll mean initial knowledge. Appendix \ref{app:further} develops additional hypotheses for education, demographics, attitudes, and briefing reading and reports the models with those additional controls.

*Initial knowledge* ($K_1$). The information-rich tend to get information-richer. The more one knows going in, the more one tends to learn and thus know going out. Absent ceiling effects, the coefficient should be $> 1$; given them, it may be $< 1$. Prior knowledge on a particular topic improves people's ability to comprehend new information, enabling them to extract the central elements of a message and draw appropriate inferences efficiently [@eckhardt1991; @recht1988]. Prior knowledge also enhances people's ability to store new information on that topic and retrieve it later [@cooke1993; @fiske1990; @hambrick2003; @mcgraw1990; @recht1988; @schneider1993]. So the less political information individuals have stored in memory, the more difficult it is for them to acquire new information.

*Mode*. Online deliberation, with its thinner exposure to other participants, may be, either intrinsically or at least in its current instantiations, a somewhat etiolated version of the face-to-face treatment [@luskin2006]. It is therefore plausible that people learn less from online deliberations.

*Observed group size*. We count group members with paired item answers. This is a lower bound on actual group size when a member lacks one of the interviews. Larger groups may offer more perspectives but give each participant less time to speak [@latane1979].

*Group knowledge*. Participants may learn more from the other members of their small group when the latter know more. The other members' mean T1 score measures their overall knowledge. Their mean correctness on the particular items a participant missed at T1 measures knowledge that could fill that participant's gaps. We add this targeted measure alongside the overall group mean in a supplementary model. Participants who missed no T1 items receive zero on the targeted measure and a separate indicator for having no missed items.

*Issue salience*. More salient issues, to which many participants have already given some attention, exhibit decidedly less net attitude change [@luskin2007b] and fall shorter of single-peakedness [@list2013]. We suspect they also, and relatedly, show less learning: there is less left to learn. We proxy salience by the participants' mean T1 knowledge.

## Model

The nested structure of the data, with individuals clustered in small groups clustered in polls, suggests a hierarchical linear model [@raudenbush2002] with random intercepts by small group and poll:
$$K_{igd2} = \beta_0 + \alpha_g + \gamma_d + \mathbf{x}_{igd}'\boldsymbol{\beta} + u_{igd},$$
where $i$, $g$ and $d$ index individuals, discussion groups, and polls. The model contains initial knowledge, groupmates' mean initial knowledge, group size, mode, and the poll's mean initial score. Poll and group random intercepts describe the nested outcome structure [@bates2015]. We fit by maximum likelihood and obtain uncertainty from the same hierarchical bootstrap used for the learning estimates. The analysis includes `r count(size("core", "n"))` participants in `r size("core", "groups")` groups and `r size("core", "polls")` polls. Appendix Table \ref{tab:historical-models} adds participant characteristics available in fewer polls.

## Who knows more after deliberation?

```{r models, results="asis"}
table_1 <- model_table(models, "core", "Main sample")
separators <- rep("", nrow(table_1))
summary_start <- match("Participants", table_1[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(table_1) - 1L)] <- "\\addlinespace[0.25em]"
table_tex <- knitr::kable(
  table_1, format = "latex", booktabs = TRUE, position = "tbp", align = "lr",
  linesep = separators,
  caption = sprintf("T2 proportion correct conditional on T1 in %s polls. Coefficients with 95 percent hierarchical-bootstrap confidence intervals.", size("core", "polls"))
)
cat(sub("\\end{table}", paste0("\\par\\medskip\\noindent{\\footnotesize\\raggedright\\emph{Notes.} Groupmates' mean T1 excludes the respondent herself; both that mean and group size use members with paired item answers. The model has poll and small-group random intercepts. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts. The models with additional participant characteristics and reading reports appear in Appendix Table \\ref{tab:historical-models}.\\par}", "\n\\end{table}"), table_tex, fixed = TRUE))
```

Table \ref{tab:models} compares T2 scores among participants with the same measured starting knowledge. The T1 coefficient is `r term("core", "k1")`; the coefficient on groupmates' initial knowledge is `r term("core", "group_k1")`. Because scores have a ceiling and contain measurement error, the relationship between T1 and T2 should not be read as a causal effect of prior knowledge. Appendix Table \ref{tab:historical-models} adds education, demographics, attitudes, and briefing-material reading. Appendix Table \ref{tab:hetero} asks whether attendee--control differences vary with initial knowledge or education and expresses the comparisons in percentage points.

## Do small groups matter?

If participants learn from each other, those placed with more knowledgeable groupmates may learn more. A causal interpretation requires groupmates' characteristics to be unrelated to one's own. We assess this by regressing each participant's initial knowledge on the mean among the other members of the group, controlling for the mean among the other members of the poll [@guryan2009]. The coefficient is `r interval(assignment$estimate[1], assignment$lower[1], assignment$upper[1])`. This check does not establish random assignment, so we describe the peer estimates as associations.

Within each poll, we regress T2 knowledge on groupmates' mean T1 knowledge, controlling for one's own and the poll's. Figure \ref{fig:peers} shows these coefficients and their equally weighted average. The average association is `r interval(peer("k1")$estimate, peer("k1")$lower, peer("k1")$upper)`. The upper endpoint corresponds to `r round(10 * peer("k1")$upper, 1)` percentage points more on T2 for a ten-point difference in groupmates' T1 scores. Appendix Table \ref{tab:historical-models} also considers knowledge of the questions each participant initially missed. Group knowledge after deliberation would be harder to interpret because groupmates share the same experience [@manski1993].

```{r peers, fig.cap = "Association between the other group members' mean T1 knowledge and own T2 knowledge, by poll, with 95\\% intervals. Each estimate controls for own T1 knowledge and the poll's leave-one-out mean, with 95 percent hierarchical-bootstrap intervals. The pooled estimate weights polls equally and resamples polls within mode and then their groups. These are descriptive associations.", fig.pos = "tbp", out.width = "100%"}
figure("peer_effects")
```

# Conclusion

Participants' factual knowledge increases across the polls we study. Reporting both percentage-point changes and increases relative to initial scores makes the magnitude interpretable, while adjustment for guessing addresses one source of measurement error. In the two polls with common before-and-after control interviews, attendees show larger gains than controls. The comparisons remain subject to selection into attendance, and the peer regressions are descriptive. The knowledge questions concern facts selected by organizers, so these estimates describe learning about the topics the events emphasized.

\FloatBarrier
# References {-}

<div id="refs"></div>

\clearpage
\appendix
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

# The Deliberative Polls {#app:polls}

All `r nrow(polls)` polls with respondent-level knowledge items enter the broader before-and-after analysis. The `r sum(polls$group_model)` polls marked as the main sample enter every main attendee analysis and Table \ref{tab:models}. Four provide control comparisons, including Northern Ireland's T3-only contrast; Antimicrobial Resistance has no group identifier.

```{r polls, results = "asis"}
polls |>
  dplyr::transmute(
    poll, year,
    mode = dplyr::if_else(mode == "Online", "Online", "In person"),
    group_model = dplyr::if_else(group_model, "Yes", "No"),
    control = dplyr::if_else(control_group, "Yes", "No")
  ) |>
  purrr::set_names(c("Poll", "Year", "Mode", "Main sample", "Control")) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lclcc",
    caption = "Polls with respondent-level knowledge-item answers at both waves."
  ) |>
  print()
```

\clearpage
# Further results {#app:further}

```{r broad-gains}
gains |>
  dplyr::arrange(pollname) |>
  dplyr::transmute(
    Poll = pollname, `Initial (%)` = pp(k1_mean),
    `Gain (pp)` = pp(raw), `Relative (%)` = pp(relative),
    `Adjusted (pp)` = pp(adjusted)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = c(">{\\raggedright\\arraybackslash}p{2.35in}", ">{\\raggedleft\\arraybackslash}p{0.8in}",
      ">{\\raggedleft\\arraybackslash}p{0.65in}", ">{\\raggedleft\\arraybackslash}p{0.8in}",
      ">{\\raggedleft\\arraybackslash}p{0.8in}"),
    caption = "Initial knowledge and gains in the broader sample. Relative gain divides observed gain by the initial score; adjusted gain accounts for guessing.")
```

```{r meta}
meta |>
  dplyr::filter(!grepl("SD", model)) |>
  dplyr::mutate(
    Sample = dplyr::if_else(grepl(", all$", model), "Broader sample", "Main sample"),
    Quantity = dplyr::case_when(
      parameter == "face_to_face" ~ "Observed gain: in person (pp)",
      parameter == "online" ~ "Observed gain: online (pp)",
      grepl("relative", model) ~ "Increase from initial score (%)",
      grepl("guessing", model) ~ "Guessing-adjusted learning (pp)",
      TRUE ~ "Observed gain (pp)"
    )
  ) |>
  dplyr::transmute(Sample, Quantity, Polls = polls, Estimate = pp(estimate),
                   `95% interval` = paste(pp(lower), pp(upper), sep = " to ")) |>
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "",
    position = "H", caption = "Average learning across polls, with hierarchical-bootstrap confidence intervals.")
```

Each row in Table \ref{tab:meta} averages the indicated quantity equally across the listed polls. Observed gain is the change in percentage correct; guessing-adjusted learning estimates the percentage who learned an item, averaged across items. Relative increase expresses each poll's observed gain as a percentage of its initial mean, then averages those ratios. The mode rows average only in-person or online polls. The broader sample adds Denmark, Vermont, and Antimicrobial Resistance, plus respondents without a group assignment. For those observations, individuals are the available resampling units, so their intervals cannot account for unobserved discussion-group dependence. Intervals use the same hierarchical bootstrap as the main analysis.

```{r hetero}
hetero |>
  dplyr::mutate(
    comparison = dplyr::if_else(moderator == "k1", "Initial score +10 pp", "Bachelor’s degree"),
    multiplier = dplyr::if_else(moderator == "k1", 10, 100)
  ) |>
  dplyr::transmute(Study = study, Comparison = comparison,
    `Difference (pp)` = num(multiplier * interaction, 1),
    `95% interval` = paste(num(multiplier * lower, 1), num(multiplier * upper, 1), sep = " to "),
    N = count(n)) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.8in}", ">{\\raggedright\\arraybackslash}p{1.35in}", ">{\\raggedleft\\arraybackslash}p{0.75in}",
      ">{\\raggedleft\\arraybackslash}p{0.85in}", ">{\\raggedleft\\arraybackslash}p{0.7in}"),
    caption = "How the baseline-adjusted attendee--control difference varies with initial knowledge and education.")
```

Table \ref{tab:hetero} reports changes in the attendee--control difference, not changes in everyone's knowledge. N is the regression sample size. Degree compares a bachelor’s degree with no degree. A negative entry means the attendee advantage is smaller for people with higher initial knowledge or a bachelor’s degree. For example, the America in One Room initial-knowledge estimate of `r num(10 * het("America in One Room 2019", "k1")$interaction, 1)` points compares two people whose starting scores differ by ten percentage points. Each regression includes attendance, initial score, the listed characteristic, and its interaction with attendance. Centering initial scores at their sample mean does not change the reported interaction. All intervals use the hierarchical bootstrap. Selection into attendance and score ceilings limit causal interpretation.

## Additional participant characteristics


*Education*. Education, qua years of schooling, is distinct from, and only moderately correlated with, political knowledge. Controlling for cognitive ability, with which it is more strongly correlated, its effect on knowledge appears negligible [@luskin1990]. But absent a separate measure of cognitive ability, it can roughly proxy the latter. Naturally, we expect it to have a positive effect. There may, however, be some mutual substitution with initial knowledge. People who know enough already may not need to be as bright to learn a lot, and people who are bright enough may not need to know as much to do so. We measure education as comparably as possible across polls in three categories: less than high school graduation, high school graduation, and a bachelor's degree or more.

*Gender and age*. These demographic variables exhibit real-world political knowledge gaps, with men knowing more than women and older people more than younger ones [@dellicarpini1996]. Age is measured in decades and gender as a dummy scored 1 for women.

*Proportion of women*. Whether from greater comfort, more widely shared cognitive or communication style, or lower stereotype threat [@steele1997], women may learn more in small groups containing more women. We calculate the proportion excluding the participant herself.

*Heterogeneity of opinion*. In the real world, people often know congenial but not uncongenial facts---those commonly cited by those sharing their views but not those commonly cited by those holding opposing views. Part of deliberation's effect on learning may consist in getting people to note and accept (if, in many cases, still argue around) uncongenial facts. This can be expected to occur mainly on site. Participants learn a great deal in anticipation of the event, but the learning at that stage is mainly of congenial facts [@hansen2004]. On site, the participants find it harder to avoid confronting the uncongenial ones. This may happen more, the more opposing points of view are represented and argued in the small group---the more attitudinally heterogeneous the group is. We gauge this as the generalized variance, the determinant of the covariance matrix, of the group's policy attitudes [@anderson2003].

*Attitude extremity*. Participants starting with more extreme views may have shuttered out more uncongenial facts, leaving them with more ground for deliberation to make up. We measure extremity as the mean, across attitude indices $j$, of $|A_{ij} - .5|$, where $A_{ij}$ is the $i$th respondent's score on the $j$th index.

*Reading the briefing materials*. One of the main information vectors is the briefing document sent to participants shortly after recruitment, normally well in advance of the on-site deliberations. It provides some background facts and lays out the major arguments for and against given policy options, which embed additional facts congenial to each argument. Respondent-linked reading reports are available for nine polls: National Issues Convention, National Issues Convention 2, Tomorrow's Europe, San Mateo, Australia, Central Power & Light, Southwestern Electric Power, West Texas Utilities, and Europolis. We score their ordered categories from zero to one using the source-specific response scales and analyze them in a separate model. Comparable reports have not been recovered for the other polls in the group-linked analysis.

\clearpage
## Models with additional controls

```{r historical-models}
historical_models <- model_table(
  models, c("historical", "briefing"),
  c("Additional controls", "Reading reports")
)
separators <- rep("", nrow(historical_models))
summary_start <- match("Participants", historical_models[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(historical_models) - 1L)] <- "\\addlinespace[0.25em]"
knitr::kable(
  historical_models, format = "latex", booktabs = TRUE,
  longtable = TRUE, align = "lrr",
  linesep = separators,
  caption = "T2 knowledge with additional controls. Coefficients with 95 percent hierarchical-bootstrap confidence intervals."
)
```

\noindent{\footnotesize\raggedright\emph{Notes.} The model with additional controls includes poll and small-group random intercepts; the reading-reports model uses poll fixed effects and small-group random intercepts. Adding groupmates' T1 correctness on questions the respondent missed to the model with additional controls gives `r term("items", "group_k1_items")` among `r count(size("items", "n"))` participants. Respondents with no missed T1 items receive zero on this measure and a separate indicator (`r term("items", "no_missed_items")`); none is removed for lack of an opportunity to learn. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts.\par}

## Additional control comparisons {#app:controls}

The main figure compares gains in the two control studies in the main sample. The additional antimicrobial-resistance study randomized invitations but released attendees and controls, so its attendee comparison includes selection into attendance [@mendelson2026]. Its observed difference in gains is `r pp(amr$gain_difference)` percentage points (95\% interval `r pp(amr$gain_lower)` to `r pp(amr$gain_upper)`). Discussion-group assignments are unavailable for that study; the bootstrap can preserve the recorded sampling units but cannot represent unidentified group dependence.

Regressing T2 scores on attendance and T1 scores answers a different question from comparing gains. The baseline-adjusted attendee--control differences are `r pp(a1r$estimate)` points in America in One Room (`r pp(a1r$lower)` to `r pp(a1r$upper)`), `r pp(climate$estimate)` in the climate poll (`r pp(climate$lower)` to `r pp(climate$upper)`), and `r pp(amr$estimate)` in the antimicrobial-resistance study (`r pp(amr$lower)` to `r pp(amr$upper)`). These intervals use the same bootstrap as the main comparisons.

A year after the climate poll, the baseline-adjusted gap is `r pp(climate_t3$estimate)` points (`r pp(climate_t3$lower)` to `r pp(climate_t3$upper)`). Northern Ireland has only a follow-up control interview: `r ni$n_treated` returning attendees scored `r pp(ni$estimate)` points above `r ni$n_control` controls (`r pp(ni$lower)` to `r pp(ni$upper)`). This is a cross-sectional comparison without a control baseline. We do not apply the immediate-learning model's no-forgetting assumption to these later outcomes.

# Knowledge items {#app:items}

The catalog lists `r nrow(item_catalog)` questions across the main and additional polls [@dpdata]. Wording is paraphrased where only source labels survive; capitalization and unambiguous abbreviations are normalized for readability. Original labels and scoring rules are preserved in the cited dataset. Choices are transcribed where recoverable; truncated or unavailable options are identified. Notes explain open-response coding.

```{r items, results='asis'}
cat(item_appendix_markdown())
```
