---
title: 'Deliberation and Learning: Evidence From Deliberative Polls\thanks{Data and replication materials: \url{https://github.com/soodoku/dp-learning}.}'
author: "Robert C. Luskin, Gaurav Sood, and James S. Fishkin"
date: "Working draft"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    fig_caption: true
bibliography: [references.bib, data-version.bib]
link-citations: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{placeins}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H")
options(knitr.graphics.rel_path = FALSE)
source("R/paper.R")
figure <- \(name) knitr::include_graphics(normalizePath(file.path("figs", paste0(name, ".pdf"))))
count <- \(x) prettyNum(x, big.mark = ",")
gains <- read_out("poll_gains.csv")
meta <- read_out("meta.csv")
controlled_learning <- read_out("control_learning.csv")
controlled_pooled <- read_out("control_learning_pooled.csv")
provenance <- jsonlite::read_json("tabs/provenance.json", simplifyVector = TRUE)
models <- read_out("models.csv")
effects <- read_out("control_effects.csv")
hetero <- read_out("control_heterogeneity.csv")
peers <- read_out("peer_effects_pooled.csv")
selected <- read_out("selection.csv")
assignment <- read_out("assignment_check.csv")
coverage <- read_out("covariate_coverage.csv")
phase_changes <- read_out("phase_contrasts.csv")
phase_coverage <- read_out("phase_coverage.csv")
phase_selection <- read_out("pre_arrival_selection.csv")
phase_provenance <- jsonlite::read_json("tabs/phase_provenance.json", simplifyVector = TRUE)
phase_balanced <- dplyr::filter(phase_changes, sample == "all_three_observed_attendees", n_people > 0)
phase_pairs <- dplyr::filter(phase_changes, sample == "available_paired_attendees", n_people > 0)
phase_baseline_exit <- dplyr::filter(phase_pairs, contrast == "post_minus_pre_arrival")
phase_row <- function(poll, contrast) {
  row <- dplyr::filter(phase_balanced, poll_id == poll, .data$contrast == .env$contrast)
  stopifnot(nrow(row) == 1L)
  row
}
phase_selection_row <- function(poll, reference) {
  row <- dplyr::filter(phase_selection, source_dataset == "control", poll_id == poll,
                      analysis == "baseline_selection_difference", category == reference)
  stopifnot(nrow(row) == 1L)
  row
}
marousi_exit <- phase_row("marousi-2006", "post_minus_arrival")
marousi_before <- phase_row("marousi-2006", "arrival_minus_pre_arrival")
marousi_total <- phase_row("marousi-2006", "post_minus_pre_arrival")
europe_exit <- phase_row("tomorrows-europe-2007", "post_minus_arrival")
europe_before <- phase_row("tomorrows-europe-2007", "arrival_minus_pre_arrival")
europe_total <- phase_row("tomorrows-europe-2007", "post_minus_pre_arrival")
a1r_invited <- phase_selection_row("america-in-one-room-2019", "attended_minus_invited_nonattender")
climate_invited <- phase_selection_row("a1r-climate-2021", "attended_minus_invited_nonattender")
a1r_baseline_control <- phase_selection_row("america-in-one-room-2019", "attended_minus_control")
climate_baseline_control <- phase_selection_row("a1r-climate-2021", "attended_minus_control")
balance_knowledge <- assignment[assignment$covariate == "k1", ]
polls <- read_out("polls.csv")
item_catalog <- read_item_catalog() |> dplyr::filter(poll_id %in% gains$poll_id)

a1r <- control_row(effects, "America in One Room 2019", "Attended vs uninvited control")
climate <- control_row(effects, "America in One Room: Climate 2021", "Attended vs uninvited control")
climate_t3 <- control_row(effects, "America in One Room: Climate 2021", "Attended vs control, one year later")
ni <- control_row(effects, "Northern Ireland 2007", "T3 attendee vs control, no baseline")
pp <- function(x) num(100 * x, 1)
pooled <- function(model) meta[meta$model == model & meta$parameter == "mu", ]
control_gain <- function(id, metric) controlled_learning[controlled_learning$poll_id == id & controlled_learning$term == metric, ]
control_average <- function(metric) controlled_pooled[controlled_pooled$term == metric, ]
persist <- round(100 * climate_t3$estimate / climate$estimate)
het <- \(study, moderator) hetero[hetero$study == study & hetero$moderator == moderator, ]
peer <- \(p) peers[peers$peer == p, ]
term <- \(model, t) model_term(models, model, t)
size <- \(model, what) models[[what]][models$model == model][1]
reading <- models[models$model == "briefing" & models$term == "read_briefing", ]
```

\begin{abstract}
We study factual learning among `r count(size("core", "n"))` participants in `r size("core", "groups")` discussion groups across `r size("core", "polls")` Deliberative Polls held between `r min(polls$year)` and `r max(polls$year)`. The polls address elections, energy, health, crime, climate, and other public issues in Europe, Australia, China, and the United States. Across each study's selected initial and later interviews, and averaging equally across polls, knowledge scores increased by `r pp(pooled("raw, pooled")$estimate)` percentage points, and the average increase relative to each poll's initial score was `r pp(pooled("relative, pooled")$estimate)`\%. An item model assuming no forgetting yields estimated learning of `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points. In two polls with before-and-after control interviews, attendees' gains exceeded controls' gains by `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` percentage points; their average difference was `r pp(control_average("raw")$estimate)` percentage points, or `r pp(control_average("adjusted")$estimate)` percentage points after guessing adjustment. In the `r dplyr::n_distinct(phase_balanced$poll_id)` studies with verified pre-arrival, arrival, and exit measurements, substantial gains precede arrival. Attendees also score higher than invited nonattenders before the event in the two control studies. Separating interview timing and attendance shows why the pooled gains cannot all be attributed to event-period deliberation. We assess uncertainty by resampling polls and their discussion groups.
\end{abstract}

# Introduction

Among the great virtues commonly ascribed to deliberation is learning. The participants learn about the topics under discussion and, to a lesser degree, incidentally, about related ones. They learn not only relevant facts but the arguments for and against particular alternatives and what other people unlike themselves, their circumstances, and their ways of thinking are like. This is helpful, both at the individual level, for discerning what policies most closely accord with one's values and interests and voting accordingly, and at the aggregate level, for forming "authentic" majorities---majorities that "know what they are doing" [@luskin2003]. How much citizens know shapes what they prefer and how they vote [@converse1964; @bartels1996; @althaus1998; @dellicarpini1996], even on foreign policy [@aldrich1989; @foyle2004], though information shortcuts can partly stand in for knowledge [@popkin1991; @lupia1994].

The expectation of learning follows directly from the nature of "deliberation," conventionally construed as discussion, but not just any discussion [@fishkin2005]. While definitions vary, deliberation is generally viewed as normatively elevated discussion---in which, among other things, the participants clarify and refine their preferences by open-mindedly weighing competing arguments. As part of that weighing, they absorb factual information and sift out misinformation.

Or so the theory runs. But how much do participants actually learn, is it the deliberation that teaches them, and who learns the most? We examine learning of relevant facts. This is not the only sort of learning involved, but it is the easiest to measure, and it is both important in itself and positively correlated with the rest.

We compare participants' knowledge across each study's retained interviews, using one sample for the main learning estimates and discussion-group models. Those interviews span different parts of the deliberative process. Where verified event timing permits, we separately measure changes before arrival and during the event on a common sample. We also compare pre-arrival knowledge by invitation and attendance to assess selection into the observed participant sample. Two of these polls also interviewed people who were not invited, allowing us to compare changes among attendees and controls. Earlier studies document learning in Deliberative Polls and other forums [@luskin2002b; @barabas2004; @sturgis2005; @esterling2011; @theuwis2025]. We report gains in both absolute and relative terms, adjust for guessing, and examine whether participants learn more alongside more knowledgeable groupmates.

# Deliberative Polling and Learning

Deliberative Polling [@fishkin1991; @fishkin1995; @luskin2002b] aims to capture fuller-information, more thoughtful opinions than those expressed in ordinary polls. In very brief, a Deliberative Poll interviews and assembles, as many as are willing of, a random sample to deliberate some policy or electoral choice, in alternating small group discussions and plenary question-and-answer sessions with panels of competing experts, and then re-measures the sample's opinions at the end. These latter are frequently different, both at the individual level and in the aggregate, from what they were before deliberating. The absolute net change is statistically significant nearly three-quarters of the time and averages nearly .1 on a 0--1 scale [@luskin2007b].

This design, admittedly a particular embodiment of deliberation, affords additional opportunities and incentives to learn. The participants are provided with carefully balanced briefing documents, rehearsing the arguments each side would make. There is usually a section of basic background information, and the arguments embed each side's pet facts---those most congenial to its arguments. In addition, the participants know from the moment they are first recruited that they will be attending the deliberation and that the deliberation (typically) will be an "Event," capital "e"---away from home, usually televised or taped for later broadcast, often attended by high-profile policy experts or policy-makers in the role of observers or panelists. Thus, motivated by a desire to "do well," many participants use the weeks between their recruitment and the event itself to prepare---not only by reading the briefing materials but by paying heightened attention to stories about the topic in the media, by talking more with friends, family, and coworkers about it, and even, in some cases, by researching it in the library or online.

So if knowledge, like its close relative "sophistication," depends on the confluence of cognitive ability, motivation, and exposure to relevant information [@luskin1990; @luskin2002a], the deliberation in Deliberative Polling stands to increase two of those three. There is not much to be done, in adulthood, about ability---and in fact we should want the participants to be representative in this, as in all other respects. But the treatment can and does affect motivation and exposure. Judging from past polls, that is enough to produce considerable, sometimes dramatic learning [@luskin2002b; @list2013].

The online design is intended to be as similar as possible to the face-to-face one, given the difference in mode. The deliberation is by voice rather than text, so as to capture as much as possible of deliberation's social elements. The participants still "meet" in small groups, still led by trained moderators, and still question panels of policy experts and policy makers. They still receive briefing materials. They are still a representative sample---sometimes randomly sampled offline, with "nonusers" receiving free access, including free equipment, to get them online, sometimes drawn from online panels and matched to approximate a random sample. On the other hand, deliberating by voice, even by video, online can never capture the full richness and complexity of in-person interactions; the discussions are interrupted, spread over a period of weeks, with much greater possibility of real-world intrusion, and the assignment to small groups is perforce constrained by possible meeting times. All these differences may act to weaken the online treatment's effects [@luskin2006; @iyengar2005].

# Theory and hypotheses

Learning requires exposure to information and the ability and motivation to absorb it. These mechanisms motivate the individual and group characteristics in our regressions.

*Initial knowledge* ($K_1$). The information-rich tend to get information-richer. The more one knows going in, the more one tends to learn and thus know going out. Absent ceiling effects, this predicts a positive association with gain; bounded scores can instead produce a negative association. Prior knowledge on a particular topic improves people's ability to comprehend new information, enabling them to extract the central elements of a message and draw appropriate inferences efficiently [@eckhardt1991; @recht1988]. Prior knowledge also enhances people's ability to store new information on that topic and retrieve it later [@cooke1993; @fiske1990; @hambrick2003; @mcgraw1990; @recht1988; @schneider1993]. So the less political information individuals have stored in memory, the more difficult it is for them to acquire new information.

*Education*. Education, qua years of schooling, is distinct from, and only moderately correlated with, political knowledge. Controlling for cognitive ability, with which it is more strongly correlated, its effect on knowledge appears negligible [@luskin1990]. But absent a separate measure of cognitive ability, it can roughly proxy the latter. Naturally, we expect it to have a positive effect. There may, however, be some mutual substitution with initial knowledge. People who know enough already may not need to be as bright to learn a lot, and people who are bright enough may not need to know as much to do so.

*Gender and age*. These demographic variables exhibit real-world political knowledge gaps, with men knowing more than women and older people more than younger ones [@dellicarpini1996]. These differences can reflect prior opportunities to acquire information. Conditional on initial knowledge, their association with learning during deliberation need not have the same direction. We include both as participant characteristics without imposing a directional hypothesis.

*Gender composition*. Whether from greater comfort, more widely shared cognitive or communication style, or lower stereotype threat [@steele1997], women may learn more in groups containing more women. We examine this possibility by interacting gender with the proportion of female groupmates.

*Reading the briefing materials*. Reading provides exposure to the facts and competing arguments discussed at the event. We therefore expect participants who read more of the materials to learn more, although motivation may influence both reading and learning.

*Group knowledge*. More knowledgeable groupmates can supply facts, correct errors, and explain unfamiliar arguments. We therefore expect their initial knowledge to be positively associated with a participant's gain, conditional on that participant's own initial knowledge.

*Group size and mode*. Larger groups may offer more perspectives but give each participant less time to speak [@latane1979], leaving the expected association with learning ambiguous. Online deliberation may offer thinner exposure to other participants and weaker opportunities to learn than in-person discussion [@luskin2006].

*Attitudes*. Strong policy preferences may motivate attention to relevant information or make conflicting information harder to accept. Disagreement within a group may expose participants to unfamiliar arguments, but it can also impede communication. We examine attitude extremity and group disagreement without imposing a directional prediction.

# Data and measures

The main sample comprises `r size("core", "polls")` polls and `r count(size("core", "n"))` attendees with retained knowledge scores at the study's selected initial and later interviews and recorded discussion-group membership. It supports both the learning estimates and the analysis of discussion groups. We use the versioned Deliberative Poll dataset [@dpdata], selecting one respondent record per poll and averaging item correctness into knowledge scores. Appendix \ref{app:items} gives the questions and scoring rules.

The selected interviews do not have a uniform time window. In the main models, $K_1$ and $K_2$ denote the retained initial and later scores, not necessarily arrival and immediate exit. A recruitment telephone interview can precede briefing and preparation; some later interviews are follow-ups. For example, the National Issues Convention's selected later score is from a survey about ten months after its event. Its immediate-exit measurement is retained separately. The broad main-sample estimates therefore summarize changes across study-specific interviews and cannot all be interpreted as learning during the event.

For comparisons with verified timing, $t_0$ denotes pre-arrival, $t_1$ arrival or the start of discussion, $t_2$ immediate exit, and $t_3$ a later follow-up. Missing phases are not renumbered. New Haven's questionnaire after its first deliberative session is an interim measurement, not arrival. The phase analysis requires recorded attendance, positively observed questionnaires, and matching knowledge batteries and denominators. The broader selected-wave benchmarks retain historically zero-filled scores even where the questionnaire's presence is uncertain; their inclusion rules differ from the verified-phase analysis. The phase analysis includes people outside the discussion-group model sample when their attendance and scores are verified. The current exports support paired pre-arrival-to-exit estimates for `r dplyr::n_distinct(phase_baseline_exit$poll_id)` polls and all three comparisons for `r dplyr::n_distinct(phase_balanced$poll_id)`; this describes verified export coverage, not the number of polls that ever collected each questionnaire. Appendix \ref{app:phases} reports the available pairs and their sample sizes.

America in One Room and the climate poll interviewed separate samples who were never invited to deliberate [@fishkin2021]. We compare changes in their scores with changes among attendees in the main sample, using the same questions in both arms. These comparisons may reflect selection into attendance as well as deliberation. Additional control comparisons and follow-ups appear in Appendix \ref{app:controls}.

## Participant and group characteristics

*Education*. We map the original response categories to below secondary completion, secondary/some college, and degree or higher. The first category is the regression reference. We interact education with initial knowledge to allow their associations with gain to depend on one another.

*Gender and age*. Gender is coded 1 for women and 0 for men. Age enters in decades. Missing or refused responses remain missing.

*Group composition and mode*. Groupmates' initial knowledge and the proportion of women exclude the participant herself. Group size counts members with both baseline and follow-up answers, so it is a lower bound on actual attendance when members lack an interview. Mode distinguishes online from in-person deliberation.

*Reading the briefing materials*. Respondent-linked reading reports are available for `r sum(coverage$reading > 0)` polls. They describe reading before discussion, recalled in the post-deliberation questionnaire. We score their ordered categories from zero to one using the source-specific response scales.

*Attitude extremity and disagreement*. We scale baseline policy measures to 0--1 using their response endpoints. The earlier polls retain their existing policy indices; the additional polls use individual policy preferences. Extremity is the participant's mean absolute distance from the midpoint, $E_i=J_i^{-1}\sum_j |A_{ij}-.5|$, over observed measures. Group disagreement is the average absolute difference between group members' responses:
$$H_g=\frac{1}{J_g}\sum_{j=1}^{J_g}\frac{2}{n_{gj}(n_{gj}-1)}\sum_{i<k}|A_{ij}-A_{kj}|.$$
Here $n_{gj}$ counts group members with an observed response to measure $j$, and $J_g$ counts measures with at least two such responses. A difference of .10 in $H_g$ means the average pairwise gap is ten percent of the response scale larger. Reversing a scale leaves both measures unchanged. Missing responses are excluded. Policy batteries differ in content and aggregation across polls, so these measures summarize attitudes toward each poll's own issues.

# Measuring Learning

We measure knowledge by closed-ended knowledge questions, mostly multiple choice, in some cases extracted from party or candidate placements, and in a few cases true-false [@luskin1987; @luskin2005; @price1999; @mondak1999; @mondak2001]. Appendix \ref{app:items} lists the items. Correct responses are scored as 1, incorrect responses and "don't knows" as 0 [@luskin2011], and these scores are then averaged across items to give the proportion answered correctly.

Table \ref{tab:models} models the change in proportion correct, $K_2-K_1$, conditional on initial knowledge $K_1$. It compares observed gains among participants with the same measured starting score. Including $K_1$ makes this algebraically equivalent to modeling $K_2$ conditional on $K_1$, as Section \ref{sec:model} shows. An unadjusted change-score model instead constrains the baseline coefficient to zero in the gain equation. These specifications can disagree when groups differ at baseline [@allison1990; @vanbreukelen2006]. Score ceilings, guessing, and measurement error complicate the relationship between observed gains and learning of the underlying facts [@luskin2016]. We therefore treat the initial-knowledge coefficient as an association, not an estimate of how prior knowledge causes learning.

A correct answer can reflect knowledge or guessing. We apply the item model of @cor2016 separately to each poll's paired answers. Under the assumption that knowledge is not lost between the selected interviews, it estimates the share who learned each fact. This assumption is stronger when the later interview is a delayed follow-up; the adjusted estimates are not uniformly measures of immediate event learning. We average across questions to obtain learning on a 0--1 scale. In each control study, we fit this model separately to attendees and controls and subtract their estimated learning. Both arms use the same items; their guessing probabilities are estimated separately. Table \ref{tab:models} uses observed individual gains, since the item model estimates learning rates rather than an individual latent outcome.

We estimate uncertainty using a hierarchical bootstrap. For estimates across polls, we sample polls with replacement separately within online and in-person modes, retaining their observed counts, and then sample whole discussion groups within each selected poll. For an individual poll, we hold the poll fixed and sample its groups. Independently sampled controls are single-person units, sampled separately from attendees. Each sampled participant retains all items and interviews. We refit the models and recompute denominators in each of `r provenance$bootstrap_replicates` draws, using the 2.5th and 97.5th percentiles for 95\% confidence intervals. Thus the same resampling scheme propagates sampling and guessing-model estimation uncertainty. Stratifying by mode preserves support for the online coefficient in every draw. Inference treats each poll's question battery as fixed.

# How Much Do Participants Learn?

An absolute gain is the change in the percentage of items answered correctly. A relative gain divides that change by the poll's initial mean score: an increase from 40\% to 50\% correct is 10 percentage points, or 25\% of the initial score. We calculate relative gains from poll means rather than averaging individual ratios, which are undefined for people with zero initial scores. The main pooled estimates give each poll equal weight; the relative result is the average of the poll-specific ratios, not the ratio of two pooled means.

Across the main sample, the average selected-wave gain is `r pp(pooled("raw, pooled")$estimate)` percentage points (95\% interval `r pp(pooled("raw, pooled")$lower)` to `r pp(pooled("raw, pooled")$upper)`). The average relative increase is `r pp(pooled("relative, pooled")$estimate)`\% (95\% interval `r pp(pooled("relative, pooled")$lower)` to `r pp(pooled("relative, pooled")$upper)`\%). Estimated learning after adjustment for guessing averages `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points (`r pp(pooled("guessing adjusted, pooled")$lower)` to `r pp(pooled("guessing adjusted, pooled")$upper)`). Appendix Table \ref{tab:meta} also reports precision-weighted averages and separate estimates by mode.

A before-and-after gain can include learning from repeated questions or events outside deliberation. Figure \ref{fig:learning} therefore also compares attendees' gains with controls' gains. The observed differences are `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` percentage points in America in One Room and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` percentage points in the climate poll. Their equally weighted average is `r pp(control_average("raw")$estimate)` percentage points (95\% interval `r pp(control_average("raw")$lower)` to `r pp(control_average("raw")$upper)`), rising to `r pp(control_average("adjusted")$estimate)` percentage points after guessing adjustment (`r pp(control_average("adjusted")$lower)` to `r pp(control_average("adjusted")$upper)`). Both estimates use the same attendees, controls, and questions. These comparisons do not eliminate selection into attendance.

```{r learning, fig.cap = "Knowledge-score changes in percentage points, with 95\\% hierarchical-bootstrap confidence intervals. A: attendee changes across each main-sample poll's selected initial and later interviews. Timing and questionnaire-presence rules vary across these study-specific windows. B: observed attendee gain minus control gain in the two polls with common baseline and follow-up interviews. Each panel ends with equally weighted observed and guessing-adjusted averages. Whole discussion groups are resampled within polls; independent controls are resampled individually. Pooled intervals resample polls within mode. With one control study per mode, the pooled control interval is conditional on these two studies. Each draw refits the guessing model, treating question batteries as fixed.", fig.pos = "tbp", out.width = "100%"}
figure("learning")
```

\FloatBarrier

## Changes before arrival and during the event {#sec:phases}

Both studies with complete three-wave samples show knowledge gains before arrival and additional gains during the event. Among `r marousi_exit$n_people` Marousi attendees with all three observed questionnaires, scores rose by `r pp(marousi_before$estimate)` percentage points from pre-arrival to arrival and `r pp(marousi_exit$estimate)` from arrival to exit, for a total of `r pp(marousi_total$estimate)` points. Among `r europe_exit$n_people` Tomorrow's Europe attendees, the corresponding changes were `r pp(europe_before$estimate)`, `r pp(europe_exit$estimate)`, and `r pp(europe_total$estimate)` points. Table \ref{tab:phase-changes} uses the same respondents for all three changes within each poll, so its total equals the sum of its two component changes before rounding. Pair-specific samples, reported in Appendix \ref{app:phases}, answer a different question when questionnaire availability differs.

```{r phase-changes, results="asis"}
phase_comparison_table(phase_balanced) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, escape = FALSE, position = "H",
    align = c(">{\\raggedright\\arraybackslash}p{1.65in}", "r", "r", "r", "r"),
    caption = "Knowledge changes on common three-wave attendee samples. Entries are percentage points; brackets contain 95 percent group-bootstrap intervals where supported."
  ) |>
  print()
```

The phase estimates are unweighted changes in proportion correct, with no guessing adjustment. Tomorrow's Europe intervals resample discussion groups in `r phase_provenance$bootstrap_replicates` draws, retaining each respondent's three scores together. Marousi has `r marousi_exit$n_unknown_group` respondents without verified discussion groups in its complete sample, so we report descriptive estimates without a confidence interval. Its original merge also contains conflicting telephone and exit identifiers; these paired estimates remain conditional on the authored row associations. We preserve that unresolved linkage rather than silently relinking respondents.

Changes before arrival could reflect briefing, preparation, repeated testing, or other events between interviews. Arrival-to-exit changes isolate the event's time window, but do not identify deliberation's causal effect without a credible comparison. Even on a common sample, these estimates describe the attendees with observed questionnaires at all three phases; they do not describe all recruited citizens or all attendees.

## Selection into attendance and exit measurement {#sec:selection}

Attendees enter the event with higher measured knowledge than either invited nonattenders or uninvited controls in both control studies. In America in One Room, the pre-arrival gaps are `r pp(a1r_invited$estimate)` and `r pp(a1r_baseline_control$estimate)` percentage points, respectively. In the climate poll they are `r pp(climate_invited$estimate)` and `r pp(climate_baseline_control$estimate)` points (Table \ref{tab:pre-arrival-selection}). These summaries use everyone with verified attendance status and an observed baseline in each source frame, including attendees without an observed exit. They therefore have different denominators from the paired gain and group-model samples.

```{r pre-arrival-selection, results="asis"}
phase_selection_table(phase_selection) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, position = "H", linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.65in}",
              ">{\\raggedright\\arraybackslash}p{1.2in}", "r", "r", "r"),
    caption = "Pre-arrival knowledge differences: attendees minus each comparison group. N counts observed baseline scores; gaps are percentage points. Descriptive summaries have no estimated sampling intervals."
  ) |>
  print()
```

Invited nonattenders describe selection after invitation; uninvited controls provide a separate untreated comparison. A baseline gap does not by itself establish how much attendance selection biases the gain comparison: the groups can also differ in unmeasured motivation, preparation, and their knowledge trajectories. Conditioning on measured baseline knowledge or subtracting control gains does not eliminate these differences. We therefore report the comparisons as descriptive and do not label them selection-adjusted causal effects.

Questionnaire completion creates an additional selection stage. Appendix Table \ref{tab:phase-attrition} compares baseline knowledge among known attendees with observed, absent, or uncertain exit questionnaires. A blank quiz within an observed questionnaire scores zero; a wholly absent questionnaire remains unmeasured. Records with unknown questionnaire presence do not enter the paired phase estimates. Records with unknown attendance remain in coverage and baseline summaries, rather than being counted as nonattenders. Current Cor-Sood source cohorts have unverified attendance in the canonical export and supply coverage rather than attendee phase estimates; their historical counterparts supply estimates where available. These are limits of the current verified data, not evidence that the original polls lacked attendees or arrival surveys.

# Who Learns?

## Model {#sec:model}

Let $G_{igd}=K_{igd2}-K_{igd1}$ denote the change in proportion correct for participant $i$ in discussion group $g$ and poll $d$. We fit a hierarchical linear model [@raudenbush2002]:
$$G_{igd}=\beta_0+\beta_1K_{igd1}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd},$$
where $\mathbf{z}_{igd}$ contains groupmates' mean initial knowledge, group size, and mode. The second specification adds education, age, gender, the proportion of female groupmates, and the education-by-initial-knowledge and gender-by-group-composition interactions. The third and fourth specifications separately extend the demographic model: the third adds reported briefing reading, and the fourth adds attitude extremity and group disagreement. The $\beta$ coefficients are fixed regression parameters. For fitting, we assume
$$\alpha_{gd}\sim N(0,\sigma_g^2),\qquad \gamma_d\sim N(0,\sigma_p^2),\qquad \varepsilon_{igd}\sim N(0,\sigma_e^2),$$
with these components mutually independent conditional on the regressors and independent across distinct groups, polls, and individuals, respectively. The group and poll random intercepts allow outcomes to covary within discussion groups and across groups in the same poll. Their variances, $\sigma_g^2$ and $\sigma_p^2$, measure residual variation at those levels; $\sigma_e^2$ measures individual residual variation. We estimate the coefficients and variance components by maximum likelihood [@bates2015] and obtain confidence intervals from the hierarchical bootstrap.

Adding $K_{igd1}$ to both sides gives the equivalent equation for the selected later score:
$$K_{igd2}=\beta_0+(1+\beta_1)K_{igd1}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd}.$$
Thus using gain instead of the later score shifts the coefficient on initial knowledge down by one and leaves the other coefficients and fitted residuals unchanged. We report gain because it directly expresses the change the paper seeks to explain. Conditioning on initial knowledge compares that change among participants with the same measured starting knowledge.

The analysis includes `r count(size("core", "n"))` participants in `r size("core", "groups")` groups and `r size("core", "polls")` polls. The demographic model retains `r count(size("demographic", "n"))` participants and all `r size("demographic", "polls")` polls; `r count(size("core", "n") - size("demographic", "n"))` participants have missing covariates. The reading model includes `r count(size("briefing", "n"))` participants in `r size("briefing", "polls")` polls. The attitude model retains `r count(size("attitudes", "n"))` participants in all `r size("attitudes", "polls")` polls; `r count(size("demographic", "n") - size("attitudes", "n"))` additional participants lack attitude measures.

## Who gains more?

```{r models, results="asis"}
table_1 <- model_table(models, c("core", "demographic", "briefing", "attitudes"), c("Baseline", "Demographics", "+ Reading", "+ Attitudes"))
table_1 <- stack_intervals(table_1)
separators <- rep("\\addlinespace[0.15em]", nrow(table_1))
summary_start <- match("Participants", table_1[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(table_1) - 1L)] <- "\\addlinespace[0.25em]"
table_tex <- knitr::kable(
  table_1, format = "latex", booktabs = TRUE, position = "!htb", align = "lrrrr", escape = FALSE,
  linesep = separators,
  caption = "Change in proportion correct (later minus initial), conditional on initial knowledge. Coefficients with 95 percent hierarchical-bootstrap confidence intervals below."
)
table_tex <- sub("\\centering", "\\centering\\small\\setlength{\\tabcolsep}{3pt}", table_tex, fixed = TRUE)
cat(sub("\\end{table}", paste0("\\par\\medskip\\noindent{\\footnotesize\\raggedright\\emph{Notes.} Groupmates' mean initial knowledge excludes the respondent herself; both that mean and group size use members with retained initial and later scores. All models have poll and small-group random intercepts. Education is coded as below secondary completion, secondary/some college, or degree or higher; the first category is the reference. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts. Columns 3 and 4 separately add reading (0--1) and attitudes to column 2. Reading reports are available for fewer polls. Replacing pairwise disagreement with the mean within-group SD gives a dispersion coefficient of ", term("attitude_sd", "attitude_sd"), "; the peer-knowledge coefficient is ", term("attitude_sd", "group_k1"), ".\\par}", "\n\\end{table}"), table_tex, fixed = TRUE))
```

Table \ref{tab:models} compares gains conditional on measured initial knowledge. The initial-knowledge coefficient in the baseline model is `r term("core", "k1")`. Because scores have a ceiling and contain measurement error, this should not be read as a causal effect of prior knowledge. The second column brings the education, age, and gender hypotheses into the same set of polls. The third column adds briefing reading, and the fourth adds attitudes. Appendix Table \ref{tab:hetero} examines whether attendee--control differences vary with initial knowledge or education.

At the reference initial score of zero, both education categories have larger estimated gains than the below-secondary group; the education-by-initial-knowledge interactions allow these gaps to vary with starting knowledge. The age estimate is `r term("demographic", "age_decades")` per decade. The intervals for age, gender, and the gender-by-group-composition interaction include zero. The gender coefficient compares women with men when the share of female groupmates is zero; the interaction shows how that difference changes as the share increases. The fourth column adds baseline attitude extremity and policy disagreement. On the same participants, adding these measures changes the peer-knowledge coefficient from `r coef_text(models$estimate[models$model == "demographic_attitude_sample" & models$term == "group_k1"])` to `r coef_text(models$estimate[models$model == "attitudes" & models$term == "group_k1"])`; the intervals for both added attitude measures include zero.

Reported reading of the briefing materials is also associated with larger gains. Across `r size("briefing", "polls")` polls, the difference between the lowest and highest reading categories is `r pp(reading$estimate)` percentage points (95\% interval `r pp(reading$lower)` to `r pp(reading$upper)`), conditional on initial knowledge, demographics, group characteristics, and mode, with poll and small-group random intercepts. These reports were collected after deliberation, so the association need not reflect the effect of reading alone.

## Group composition and learning

Table \ref{tab:models} addresses the peer-knowledge hypothesis directly. A ten-percentage-point difference in groupmates' mean initial knowledge is associated with `r num(10 * models$estimate[models$model == "core" & models$term == "group_k1"], 2)` percentage points of additional gain in the baseline model and `r num(10 * models$estimate[models$model == "demographic" & models$term == "group_k1"], 2)` after demographic adjustment. The intervals in the table quantify uncertainty about these pooled associations.

Appendix Figure \ref{fig:peers} estimates the association separately within each poll. It shows whether the pooled result conceals different patterns across settings. Its average weights those separate slopes equally; Table \ref{tab:models} instead fits a common slope in a hierarchical model. Its contribution is to display variation across polls.

Appendix Table \ref{tab:baseline-composition} examines whether participants with similar baseline characteristics were grouped together. For initial knowledge, the association with groupmates' initial knowledge is `r interval(balance_knowledge$estimate, balance_knowledge$lower, balance_knowledge$upper)`. Its interval includes zero. This diagnostic concerns characteristics measured before discussion; later knowledge and gain can themselves respond to group composition. A causal interpretation of peer coefficients also requires evidence about the assignment process. Group knowledge after deliberation is difficult to interpret because groupmates share the same experience [@manski1993].

# Conclusion

Participants' factual knowledge increases across the study-specific interview windows we examine. Reporting absolute gains in percentage points and increases relative to initial scores makes the magnitude interpretable, while adjustment for guessing addresses one source of measurement error. In the two polls with common before-and-after control interviews, attendees show larger gains than controls. The comparisons remain subject to selection into attendance, and the peer regressions are descriptive. The three-wave comparisons show that knowledge can increase before arrival as well as during the event. Baseline gaps by attendance and differential questionnaire completion limit what the gains establish about recruited citizens or deliberation's causal effect. The knowledge questions concern facts selected by organizers, so these estimates describe learning about the topics the events emphasized.

\FloatBarrier
# References {-}

\begingroup
\small

<div id="refs"></div>

\endgroup

\clearpage
\appendix
\counterwithin{table}{section}
\counterwithin{figure}{section}
\renewcommand{\thetable}{\thesection\arabic{table}}
\renewcommand{\thefigure}{\thesection\arabic{figure}}

# The Deliberative Polls {#app:polls}

These `r nrow(polls)` polls supply the main learning and discussion-group sample. The verified-phase comparisons use their own samples, described in Appendix \ref{app:phases}. Three also provide control comparisons, including Northern Ireland's T3-only contrast. Tables \ref{tab:polls} and \ref{tab:poll-gains} list polls in the same chronological order.

```{r polls, results = "asis"}
polls |>
  dplyr::transmute(
    poll, year,
    mode = dplyr::if_else(mode == "Online", "Online", "In person"),
    control = dplyr::if_else(control_group, "Yes", "No")
  ) |>
  purrr::set_names(c("Poll", "Year", "Mode", "Control")) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lclc",
    caption = "Polls with respondent-level knowledge-item answers at both waves."
  ) |>
  print()
```

\clearpage
# Further results {#app:further}

```{r poll-gains}
gains |>
  dplyr::arrange(match(poll_id, polls$poll_id)) |>
  dplyr::transmute(
    Poll = pollname, `Initial (%)` = pp(k1_mean),
    `Gain (pp)` = pp(raw),
    `Adjusted (pp)` = pp(adjusted)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = c(">{\\raggedright\\arraybackslash}p{2.8in}", "r", "r", "r"),
    caption = "Selected initial knowledge (percent correct), selected-wave gains, and guessing-adjusted learning (percentage points) in the main sample. Interview windows and questionnaire-presence rules vary by poll.")
```

\begin{minipage}{\textwidth}

```{r meta}
meta |>
  dplyr::filter(!grepl("SD|relative", model)) |>
  dplyr::mutate(
    Weighting = dplyr::if_else(grepl("precision weighted", model), "Precision", "Equal"),
    Quantity = dplyr::case_when(
      parameter == "face_to_face" ~ "Observed gain: in person (pp)",
      parameter == "online" ~ "Observed gain: online (pp)",
      grepl("guessing", model) ~ "Guessing-adjusted learning (pp)",
      TRUE ~ "Observed gain (pp)"
    )
  ) |>
  dplyr::transmute(Quantity, Weighting, Polls = polls, Estimate = pp(estimate),
                   `95% interval` = paste(pp(lower), pp(upper), sep = " to ")) |>
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "",
    position = "H", caption = "Average learning across polls, with hierarchical-bootstrap confidence intervals.")
```

Observed gain is the selected later score minus the selected initial score; guessing-adjusted learning estimates the percentage who learned an item, averaged across questions. Equal weighting gives each poll the same weight. Precision weighting uses $w_d=1/\widehat{\mathrm{Var}}(\hat\theta_d)$, with sampling variances estimated by resampling discussion groups within each poll. The weighted average is $\sum_d w_d\hat\theta_d/\sum_d w_d$. These weights are held fixed when resampling polls and groups for its interval. The mode rows split the equal-poll average into in-person and online polls.

\end{minipage}

```{r hetero}
hetero |>
  dplyr::mutate(
    comparison = dplyr::if_else(moderator == "k1", "Initial score +10 pp", "Bachelor’s degree"),
    multiplier = dplyr::if_else(moderator == "k1", 10, 100)
  ) |>
  dplyr::transmute(Study = study, Comparison = comparison,
    `Difference (pp)` = num(multiplier * interaction, 1),
    `95% interval` = paste(num(multiplier * lower, 1), num(multiplier * upper, 1), sep = " to "),
    N = count(n)) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.8in}", ">{\\raggedright\\arraybackslash}p{1.35in}", ">{\\raggedleft\\arraybackslash}p{0.75in}",
      ">{\\raggedleft\\arraybackslash}p{0.85in}", ">{\\raggedleft\\arraybackslash}p{0.7in}"),
    caption = "How the baseline-adjusted attendee--control difference varies with initial knowledge and education.")
```

Table \ref{tab:hetero} reports changes in the attendee--control difference, not changes in everyone's knowledge. N is the regression sample size. Degree compares a bachelor’s degree with no degree. A negative entry means the attendee advantage is smaller for people with higher initial knowledge or a bachelor’s degree. For example, the America in One Room initial-knowledge estimate of `r num(10 * het("America in One Room 2019", "k1")$interaction, 1)` percentage points compares two people whose starting scores differ by ten percentage points. Each regression includes attendance, initial score, the listed characteristic, and its interaction with attendance. Centering initial scores at their sample mean does not change the reported interaction. All intervals use the hierarchical bootstrap. Selection into attendance and score ceilings limit causal interpretation.

## Groupmates' knowledge of missed questions {#app:item-peers}

The demographic specification can also include groupmates' initial correctness on the particular questions the participant missed at the initial interview. For each missed question, we calculate the proportion of other group members who answered correctly, then average across those questions. This targeted measure enters alongside groupmates' overall initial knowledge. Respondents who answered every initial question correctly receive zero on the targeted measure and a separate indicator for having no missed questions.

The targeted peer coefficient is `r term("items", "group_k1_items")` among `r count(size("items", "n"))` participants in `r size("items", "polls")` polls. The coefficient on overall peer knowledge is `r term("items", "group_k1")`, and the no-missed-questions indicator is `r term("items", "no_missed_items")`. All other regressors, random intercepts, and the bootstrap follow Table \ref{tab:models}.

## Additional control comparisons {#app:controls}

Regressing later scores on attendance and initial scores answers a different question from comparing gains. The baseline-adjusted attendee--control differences are `r pp(a1r$estimate)` percentage points in America in One Room (`r pp(a1r$lower)` to `r pp(a1r$upper)`), and `r pp(climate$estimate)` in the climate poll (`r pp(climate$lower)` to `r pp(climate$upper)`). These intervals use the same bootstrap as the main comparisons.

A year after the climate poll, the baseline-adjusted gap is `r pp(climate_t3$estimate)` percentage points (`r pp(climate_t3$lower)` to `r pp(climate_t3$upper)`). Northern Ireland has only a follow-up control interview: `r ni$n_treated` returning attendees scored `r pp(ni$estimate)` percentage points above `r ni$n_control` controls (`r pp(ni$lower)` to `r pp(ni$upper)`). This is a cross-sectional comparison without a control baseline. We do not apply the main guessing model's no-forgetting assumption to these later outcomes.

## Poll-specific peer associations

The following estimates fit a separate slope in each poll, conditional on own initial knowledge. Their equal-poll average is `r interval(peer("k1")$estimate, peer("k1")$lower, peer("k1")$upper)`.

```{r peers, fig.cap = "Association between the other group members' mean initial knowledge and own knowledge gain, by poll, with 95\\% intervals. Each estimate controls for own initial knowledge. Intervals use the hierarchical bootstrap. The pooled estimate weights polls equally and resamples polls within mode and then their groups. These are descriptive associations.", fig.pos = "tbp", out.width = "100%"}
figure("peer_effects")
```

## Baseline composition of discussion groups

We regress each participant's baseline characteristic on its mean among the other group members, controlling for the mean among other members of the poll and poll fixed effects [@guryan2009]. This is a descriptive sorting diagnostic; it does not reconstruct the original assignment mechanism. A randomization test would require the original assignment pools and restrictions, including any scheduling or language constraints. Balance on observed covariates alone does not establish random assignment.

```{r baseline-composition}
assignment |>
  dplyr::transmute(
    Characteristic = dplyr::recode(covariate, k1 = "Initial knowledge", female = "Female", age = "Age", education_ba = "Degree or higher"),
    Association = interval(estimate, lower, upper), Participants = count(n)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "", caption = "Baseline composition of discussion groups. Associations with 95 percent hierarchical-bootstrap intervals.")
```

\clearpage
# Verified-phase comparisons and sample coverage {#app:phases}

Table \ref{tab:phase-pairs} reports available paired attendees for each verified comparison. These estimates are unweighted changes in proportion correct without guessing adjustment. Each row requires observed questionnaires and the same battery and denominator at both phases; it does not require the third phase or discussion-group membership. Thus the three rows for a poll need not use the same people and need not add up. Table \ref{tab:phase-changes} instead restricts all three changes to the common three-wave sample. Estimates without intervals retain their respondents rather than inventing a clustering unit.

```{r phase-pairs, results="asis"}
phase_pair_table(phase_pairs) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.95in}",
              ">{\\raggedright\\arraybackslash}p{1.55in}", "r", "r", "r"),
    caption = "Verified-phase changes among available paired attendees. Changes and 95 percent group-bootstrap intervals are percentage points. A dash means no supported interval because groups are unknown or too few."
  ) |>
  print()
```

Table \ref{tab:phase-attrition} reports exit-questionnaire availability and baseline knowledge for the two three-wave studies and the two control studies. An absent exit requires evidence of a wholly absent questionnaire; an uncertain exit is not recoded as absent. Baseline means use the scored records counted in the last-but-one column. The groups shown exhaust known attendees in these source frames, including people excluded from paired-change estimates.

```{r phase-attrition, results="asis"}
phase_attrition_table(phase_selection) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{2.0in}",
              ">{\\raggedright\\arraybackslash}p{1.05in}", "r", "r", "r"),
    caption = "Exit availability among known attendees: sample counts and pre-arrival knowledge (percent correct)."
  ) |>
  print()
```

Marousi's full telephone frame contains `r count(phase_coverage$n_people[phase_coverage$poll_id == "marousi-2006" & phase_coverage$phase == "t0"])` records, of which `r count(phase_coverage$n_unknown_attendance[phase_coverage$poll_id == "marousi-2006" & phase_coverage$phase == "t0"])` have unknown attendance. Their baseline scores are retained in the source export, but absence of an arrival or exit form does not establish that they were invited or that they declined. The `r dplyr::n_distinct(phase_coverage$poll_id[phase_coverage$source_dataset == "cor_sood"])` Cor-Sood cohorts likewise retain unknown attendance. No cross-deposit linkage is inferred from an identical respondent number. Detailed coverage and baseline summaries remain in the replication tables; the phase input revision is \texttt{`r substr(phase_provenance$data_commit, 1, 7)`}, and the main-analysis revision is \texttt{`r substr(provenance$data_commit, 1, 7)`} [@dpdata].

\clearpage
# Knowledge items {#app:items}

The catalog lists `r nrow(item_catalog)` questions for the same `r nrow(gains)` polls used in the main learning and group-model analysis [@dpdata]. Wording is paraphrased where only source labels survive; capitalization and unambiguous abbreviations are normalized for readability. Original labels and scoring rules are preserved in the cited dataset. Choices are transcribed where recoverable; truncated or unavailable options are identified. Notes explain open-response coding.

```{r items, results='asis'}
cat(item_appendix_markdown(items = item_catalog))
```
