---
title: 'Deliberation and Learning: Evidence From Deliberative Polls\thanks{Data and replication materials: \url{https://github.com/soodoku/dp-learning}.}'
author: "Robert C. Luskin, Gaurav Sood, and James S. Fishkin"
date: "Working draft"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    fig_caption: true
bibliography: [references.bib, data-version.bib]
link-citations: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{placeins}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H")
options(knitr.graphics.rel_path = FALSE)
source("R/paper.R")
figure <- \(name) knitr::include_graphics(normalizePath(file.path("figs", paste0(name, ".pdf"))))
count <- \(x) prettyNum(x, big.mark = ",")
gains <- read_out("poll_gains.csv")
meta <- read_out("meta.csv")
controlled_learning <- read_out("control_learning.csv")
controlled_pooled <- read_out("control_learning_pooled.csv")
provenance <- jsonlite::read_json("tabs/provenance.json", simplifyVector = TRUE)
models <- read_out("models.csv")
effects <- read_out("control_effects.csv")
hetero <- read_out("control_heterogeneity.csv")
peers <- read_out("peer_effects_pooled.csv")
selected <- read_out("selection.csv")
assignment <- read_out("assignment_check.csv")
coverage <- read_out("covariate_coverage.csv")
balance_knowledge <- assignment[assignment$covariate == "k1", ]
polls <- read_out("polls.csv")
item_catalog <- read_item_catalog() |> dplyr::filter(poll_id %in% gains$poll_id)

a1r <- control_row(effects, "America in One Room 2019", "Attended vs uninvited control")
climate <- control_row(effects, "America in One Room: Climate 2021", "Attended vs uninvited control")
climate_t3 <- control_row(effects, "America in One Room: Climate 2021", "Attended vs control, one year later")
ni <- control_row(effects, "Northern Ireland 2007", "T3 attendee vs control, no baseline")
pp <- function(x) num(100 * x, 1)
pooled <- function(model) meta[meta$model == model & meta$parameter == "mu", ]
control_gain <- function(id, metric) controlled_learning[controlled_learning$poll_id == id & controlled_learning$term == metric, ]
control_average <- function(metric) controlled_pooled[controlled_pooled$term == metric, ]
persist <- round(100 * climate_t3$estimate / climate$estimate)
het <- \(study, moderator) hetero[hetero$study == study & hetero$moderator == moderator, ]
peer <- \(p) peers[peers$peer == p, ]
term <- \(model, t) model_term(models, model, t)
size <- \(model, what) models[[what]][models$model == model][1]
reading <- models[models$model == "briefing" & models$term == "read_briefing", ]
```

\begin{abstract}
We study factual learning among `r count(size("core", "n"))` participants in `r size("core", "groups")` discussion groups across `r size("core", "polls")` Deliberative Polls held between `r min(polls$year)` and `r max(polls$year)`. The polls address elections, energy, health, crime, climate, and other public issues in Europe, Australia, China, and the United States. Averaging equally across polls, knowledge scores increased by `r pp(pooled("raw, pooled")$estimate)` percentage points, and the average increase relative to each poll's initial score was `r pp(pooled("relative, pooled")$estimate)`\%. Adjustment for guessing yields estimated learning of `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points. In two polls with before-and-after control interviews, attendees' gains exceeded controls' gains by `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` percentage points; their average difference was `r pp(control_average("raw")$estimate)` percentage points, or `r pp(control_average("adjusted")$estimate)` percentage points after guessing adjustment. We assess uncertainty by resampling polls and their discussion groups.
\end{abstract}

# Introduction

Among the great virtues commonly ascribed to deliberation is learning. The participants learn about the topics under discussion and, to a lesser degree, incidentally, about related ones. They learn not only relevant facts but the arguments for and against particular alternatives and what other people unlike themselves, their circumstances, and their ways of thinking are like. This is helpful, both at the individual level, for discerning what policies most closely accord with one's values and interests and voting accordingly, and at the aggregate level, for forming "authentic" majorities---majorities that "know what they are doing" [@luskin2003]. How much citizens know shapes what they prefer and how they vote [@converse1964; @bartels1996; @althaus1998; @dellicarpini1996], even on foreign policy [@aldrich1989; @foyle2004], though information shortcuts can partly stand in for knowledge [@popkin1991; @lupia1994].

The expectation of learning follows directly from the nature of "deliberation," conventionally construed as discussion, but not just any discussion [@fishkin2005]. While definitions vary, deliberation is generally viewed as normatively elevated discussion---in which, among other things, the participants clarify and refine their preferences by open-mindedly weighing competing arguments. As part of that weighing, they absorb factual information and sift out misinformation.

Or so the theory runs. But how much do participants actually learn, is it the deliberation that teaches them, and who learns the most? We examine learning of relevant facts. This is not the only sort of learning involved, but it is the easiest to measure, and it is both important in itself and positively correlated with the rest.

We compare the knowledge participants bring to deliberation with what they know afterward, using one sample throughout the main analysis. Two of these polls also interviewed people who were not invited, allowing us to compare changes among attendees and controls. Earlier studies document learning in Deliberative Polls and other forums [@luskin2002b; @barabas2004; @sturgis2005; @esterling2011; @theuwis2025]. We report gains in both absolute and relative terms, adjust for guessing, and examine whether participants learn more alongside more knowledgeable groupmates.

# Deliberative Polling and Learning

Deliberative Polling [@fishkin1991; @fishkin1995; @luskin2002b] aims to capture fuller-information, more thoughtful opinions than those expressed in ordinary polls. In very brief, a Deliberative Poll interviews and assembles, as many as are willing of, a random sample to deliberate some policy or electoral choice, in alternating small group discussions and plenary question-and-answer sessions with panels of competing experts, and then re-measures the sample's opinions at the end. These latter are frequently different, both at the individual level and in the aggregate, from what they were before deliberating. The absolute net change is statistically significant nearly three-quarters of the time and averages nearly .1 on a 0--1 scale [@luskin2007b].

This design, admittedly a particular embodiment of deliberation, affords additional opportunities and incentives to learn. The participants are provided with carefully balanced briefing documents, rehearsing the arguments each side would make. There is usually a section of basic background information, and the arguments embed each side's pet facts---those most congenial to its arguments. In addition, the participants know from the moment they are first recruited that they will be attending the deliberation and that the deliberation (typically) will be an "Event," capital "e"---away from home, usually televised or taped for later broadcast, often attended by high-profile policy experts or policy-makers in the role of observers or panelists. Thus, motivated by a desire to "do well," many participants use the weeks between their recruitment and the event itself to prepare---not only by reading the briefing materials but by paying heightened attention to stories about the topic in the media, by talking more with friends, family, and coworkers about it, and even, in some cases, by researching it in the library or online.

So if knowledge, like its close relative "sophistication," depends on the confluence of cognitive ability, motivation, and exposure to relevant information [@luskin1990; @luskin2002a], the deliberation in Deliberative Polling stands to increase two of those three. There is not much to be done, in adulthood, about ability---and in fact we should want the participants to be representative in this, as in all other respects. But the treatment can and does affect motivation and exposure. Judging from past polls, that is enough to produce considerable, sometimes dramatic learning [@luskin2002b; @list2013].

The online design is intended to be as similar as possible to the face-to-face one, given the difference in mode. The deliberation is by voice rather than text, so as to capture as much as possible of deliberation's social elements. The participants still "meet" in small groups, still led by trained moderators, and still question panels of policy experts and policy makers. They still receive briefing materials. They are still a representative sample---sometimes randomly sampled offline, with "nonusers" receiving free access, including free equipment, to get them online, sometimes drawn from online panels and matched to approximate a random sample. On the other hand, deliberating by voice, even by video, online can never capture the full richness and complexity of in-person interactions; the discussions are interrupted, spread over a period of weeks, with much greater possibility of real-world intrusion, and the assignment to small groups is perforce constrained by possible meeting times. All these differences may act to weaken the online treatment's effects [@luskin2006; @iyengar2005].

# Data

The main sample comprises `r size("core", "polls")` polls and `r count(size("core", "n"))` attendees with knowledge answers before and after deliberation and recorded discussion-group membership. It supports both the learning estimates and the analysis of discussion groups. We use the versioned Deliberative Poll dataset [@dpdata], selecting one respondent record per poll and averaging item correctness into knowledge scores. Appendix \ref{app:items} gives the questions and scoring rules.

America in One Room and the climate poll interviewed separate samples who were never invited to deliberate [@fishkin2021]. We compare changes in their scores with changes among attendees in the main sample, using the same questions in both arms. These comparisons may reflect selection into attendance as well as deliberation. Additional control comparisons and follow-ups appear in Appendix \ref{app:controls}.

# Measuring Learning

We measure knowledge by closed-ended knowledge questions, mostly multiple choice, in some cases extracted from party or candidate placements, and in a few cases true-false [@luskin1987; @luskin2005; @price1999; @mondak1999; @mondak2001]. Appendix \ref{app:items} lists the items. Correct responses are scored as 1, incorrect responses and "don't knows" as 0 [@luskin2011], and these scores are then averaged across items to give the proportion answered correctly.

Table \ref{tab:models} models the change in proportion correct, $K_2-K_1$, conditional on initial knowledge $K_1$. It compares observed gains among participants with the same measured starting score. Including $K_1$ makes this algebraically equivalent to modeling $K_2$ conditional on $K_1$, as Section \ref{sec:model} shows. An unadjusted change-score model instead constrains the baseline coefficient to zero in the gain equation. These specifications can disagree when groups differ at baseline [@allison1990; @vanbreukelen2006]. Score ceilings, guessing, and measurement error complicate the relationship between observed gains and learning of the underlying facts [@luskin2016]. We therefore treat the T1 coefficient as an association, not an estimate of how prior knowledge causes learning.

A correct answer can reflect knowledge or guessing. We apply the item model of @cor2016 separately to each poll's paired answers. Under the assumption that knowledge is not lost between the two interviews, it estimates the share who learned each fact. We average across questions to obtain learning on a 0--1 scale. In each control study, we fit this model separately to attendees and controls and subtract their estimated learning. Both arms use the same items; their guessing probabilities are estimated separately. Table \ref{tab:models} uses observed individual gains, since the item model estimates learning rates rather than an individual latent outcome.

We estimate uncertainty using a hierarchical bootstrap. For estimates across polls, we sample polls with replacement separately within online and in-person modes, retaining their observed counts, and then sample whole discussion groups within each selected poll. For an individual poll, we hold the poll fixed and sample its groups. Independently sampled controls are single-person units, sampled separately from attendees. Each sampled participant retains all items and interviews. We refit the models and recompute denominators in each of `r provenance$bootstrap_replicates` draws, using the 2.5th and 97.5th percentiles for 95\% confidence intervals. Thus the same resampling scheme propagates sampling and guessing-model estimation uncertainty. Stratifying by mode preserves support for the online coefficient in every draw. Inference treats each poll's question battery as fixed.

# How Much Do Participants Learn?

An absolute gain is the change in the percentage of items answered correctly. A relative gain divides that change by the poll's initial mean score: an increase from 40\% to 50\% correct is 10 percentage points, or 25\% of the initial score. We calculate relative gains from poll means rather than averaging individual ratios, which are undefined for people with zero initial scores. The main pooled estimates give each poll equal weight; the relative result is the average of the poll-specific ratios, not the ratio of two pooled means.

Across the main sample, the average observed gain is `r pp(pooled("raw, pooled")$estimate)` percentage points (95\% interval `r pp(pooled("raw, pooled")$lower)` to `r pp(pooled("raw, pooled")$upper)`). The average relative increase is `r pp(pooled("relative, pooled")$estimate)`\% (95\% interval `r pp(pooled("relative, pooled")$lower)` to `r pp(pooled("relative, pooled")$upper)`\%). Estimated learning after adjustment for guessing averages `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points (`r pp(pooled("guessing adjusted, pooled")$lower)` to `r pp(pooled("guessing adjusted, pooled")$upper)`). Appendix Table \ref{tab:meta} also reports precision-weighted averages and separate estimates by mode.

A before-and-after gain can include learning from repeated questions or events outside deliberation. Figure \ref{fig:learning} therefore also compares attendees' gains with controls' gains. The observed differences are `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` percentage points in America in One Room and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` percentage points in the climate poll. Their equally weighted average is `r pp(control_average("raw")$estimate)` percentage points (95\% interval `r pp(control_average("raw")$lower)` to `r pp(control_average("raw")$upper)`), rising to `r pp(control_average("adjusted")$estimate)` percentage points after guessing adjustment (`r pp(control_average("adjusted")$lower)` to `r pp(control_average("adjusted")$upper)`). Both estimates use the same attendees, controls, and questions. These comparisons do not eliminate selection into attendance.

```{r learning, fig.cap = "Learning in percentage points, with 95\\% hierarchical-bootstrap confidence intervals. A: observed attendee gains in the main-sample polls. B: observed attendee gain minus control gain in the two polls with common baseline and follow-up interviews. Each panel ends with equally weighted observed and guessing-adjusted averages. Whole discussion groups are resampled within polls; independent controls are resampled individually. Pooled intervals resample polls within mode. With one control study per mode, the pooled control interval is conditional on these two studies. Each draw refits the guessing model, treating question batteries as fixed.", fig.pos = "tbp", out.width = "100%"}
figure("learning")
```

# Who Learns?

## Regressors and hypotheses

Table \ref{tab:models} begins with initial knowledge, groupmates' initial knowledge, group size, mode, and the poll's mean initial score. Its second column adds education, age, gender, and the gender composition of the group. These variables are available in all `r size("core", "polls")` polls. Education may improve the acquisition of new information [@luskin1990]; we allow its association with gain to vary with initial knowledge. The interaction between gender and the proportion of female groupmates tests whether that association varies with group composition. Appendix \ref{app:further} develops the additional hypotheses about attitudes and briefing reading.

*Initial knowledge* ($K_1$). The information-rich tend to get information-richer. The more one knows going in, the more one tends to learn and thus know going out. Absent ceiling effects, this predicts a positive association with gain; bounded scores can instead produce a negative association. Prior knowledge on a particular topic improves people's ability to comprehend new information, enabling them to extract the central elements of a message and draw appropriate inferences efficiently [@eckhardt1991; @recht1988]. Prior knowledge also enhances people's ability to store new information on that topic and retrieve it later [@cooke1993; @fiske1990; @hambrick2003; @mcgraw1990; @recht1988; @schneider1993]. So the less political information individuals have stored in memory, the more difficult it is for them to acquire new information.

*Mode*. Online deliberation, with its thinner exposure to other participants, may be, either intrinsically or at least in its current instantiations, a somewhat etiolated version of the face-to-face treatment [@luskin2006]. It is therefore plausible that people learn less from online deliberations.

*Group size*. We count group members with both baseline and follow-up answers. This is a lower bound on actual group size when a member lacks one of the interviews. Larger groups may offer more perspectives but give each participant less time to speak [@latane1979].

*Group knowledge*. Participants may learn more from the other members of their small group when the latter know more. The other members' mean T1 score measures their overall knowledge. Their mean correctness on the particular items a participant missed at T1 measures knowledge that could fill that participant's gaps. We add this targeted measure alongside the overall group mean in a supplementary model. Participants who missed no T1 items receive zero on the targeted measure and a separate indicator for having no missed items.

*Initial knowledge in the poll*. We include the poll's mean T1 score to account for differences in baseline knowledge across events. It reflects both attendees' knowledge and the difficulty of the question battery. It is a contextual control; we do not interpret it as a direct measure of issue salience.

## Model {#sec:model}

Let $G_{igd}=K_{igd2}-K_{igd1}$ denote the change in proportion correct for participant $i$ in discussion group $g$ and poll $d$. We fit a hierarchical linear model [@raudenbush2002]:
$$G_{igd}=\beta_0+\beta_1K_{igd1}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd},$$
where $\mathbf{z}_{igd}$ contains groupmates' mean initial knowledge, group size, mode, and the poll's mean initial score. The second specification adds education, age, gender, the proportion of female groupmates, and the education-by-T1 and gender-by-group-composition interactions. The $\beta$ coefficients are fixed regression parameters. For fitting, we assume
$$\alpha_{gd}\sim N(0,\sigma_g^2),\qquad \gamma_d\sim N(0,\sigma_p^2),\qquad \varepsilon_{igd}\sim N(0,\sigma_e^2),$$
with these components mutually independent conditional on the regressors and independent across distinct groups, polls, and individuals, respectively. The group and poll random intercepts allow outcomes to covary within discussion groups and across groups in the same poll. Their variances, $\sigma_g^2$ and $\sigma_p^2$, measure residual variation at those levels; $\sigma_e^2$ measures individual residual variation. We estimate the coefficients and variance components by maximum likelihood [@bates2015] and obtain confidence intervals from the hierarchical bootstrap.

Adding $K_{igd1}$ to both sides gives the equivalent post-deliberation equation:
$$K_{igd2}=\beta_0+(1+\beta_1)K_{igd1}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd}.$$
Thus using gain instead of T2 shifts the coefficient on T1 down by one and leaves the other coefficients and fitted residuals unchanged. We report gain because it directly expresses the change the paper seeks to explain. Conditioning on T1 compares that change among participants with the same measured starting knowledge.

The analysis includes `r count(size("core", "n"))` participants in `r size("core", "groups")` groups and `r size("core", "polls")` polls. The demographic model retains `r count(size("demographic", "n"))` participants and all `r size("demographic", "polls")` polls; `r count(size("core", "n") - size("demographic", "n"))` participants have missing covariates. Appendix Table \ref{tab:sample-comparison} distinguishes changes due to sample restrictions from changes due to added controls.

## Who gains more?

```{r models, results="asis"}
table_1 <- model_table(models, c("core", "demographic"), c("Baseline", "Demographic controls"))
separators <- rep("", nrow(table_1))
summary_start <- match("Participants", table_1[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(table_1) - 1L)] <- "\\addlinespace[0.25em]"
table_tex <- knitr::kable(
  table_1, format = "latex", booktabs = TRUE, position = "!htb", align = "lrr",
  linesep = separators,
  caption = sprintf("Change in proportion correct (T2 minus T1), conditional on T1, in %s polls. Coefficients with 95 percent hierarchical-bootstrap confidence intervals.", size("core", "polls"))
)
cat(sub("\\end{table}", paste0("\\par\\medskip\\noindent{\\footnotesize\\raggedright\\emph{Notes.} Groupmates' mean T1 excludes the respondent herself; both that mean and group size use members with baseline and follow-up answers. Both models have poll and small-group random intercepts. Education is coded as below secondary completion, secondary/some college, or degree or higher; the first category is the reference. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts. The models with additional attitude measures and reading reports appear in Appendix Table \\ref{tab:historical-models}.\\par}", "\n\\end{table}"), table_tex, fixed = TRUE))
```

Table \ref{tab:models} compares gains conditional on measured initial knowledge. The T1 coefficient in the baseline model is `r term("core", "k1")`. Because scores have a ceiling and contain measurement error, this should not be read as a causal effect of prior knowledge. The second column brings the education, age, and gender hypotheses into the same set of polls. Appendix Table \ref{tab:historical-models} adds attitude extremity and opinion heterogeneity in `r size("historical", "polls")` polls and examines briefing-material reading in `r size("briefing", "polls")` polls. Appendix Table \ref{tab:sample-comparison} fits the baseline specification again on each restricted sample before adding covariates, making the source of any coefficient change explicit. Appendix Table \ref{tab:hetero} examines whether attendee--control differences vary with initial knowledge or education.

At the reference initial score of zero, both education categories have larger estimated gains than the below-secondary group; the education-by-T1 interactions allow these gaps to vary with starting knowledge. The age estimate is `r term("demographic", "age_decades")` per decade. The intervals for age, gender, and the gender-by-group-composition interaction include zero. The gender coefficient compares women with men when the share of female groupmates is zero; the interaction shows how that difference changes as the share increases.

Reported reading of the briefing materials is also associated with larger gains. Across `r size("briefing", "polls")` polls, the difference between the lowest and highest reading categories is `r pp(reading$estimate)` percentage points (95\% interval `r pp(reading$lower)` to `r pp(reading$upper)`), conditional on initial knowledge, demographics, and poll fixed effects. These reports were collected after deliberation, so the association need not reflect the effect of reading alone.

## Group composition and learning

Table \ref{tab:models} addresses the peer-knowledge hypothesis directly. A ten-percentage-point difference in groupmates' mean initial knowledge is associated with `r num(10 * models$estimate[models$model == "core" & models$term == "group_k1"], 2)` percentage points of additional gain in the baseline model and `r num(10 * models$estimate[models$model == "demographic" & models$term == "group_k1"], 2)` after demographic adjustment. The intervals in the table quantify uncertainty about these pooled associations.

Appendix Figure \ref{fig:peers} estimates the association separately within each poll. It shows whether the pooled result conceals different patterns across settings. Its average weights those separate slopes equally; Table \ref{tab:models} instead fits a common slope in a hierarchical model. Its contribution is to display variation across polls.

Appendix Table \ref{tab:baseline-composition} examines whether participants with similar baseline characteristics were grouped together. For initial knowledge, the association with groupmates' initial knowledge is `r interval(balance_knowledge$estimate, balance_knowledge$lower, balance_knowledge$upper)`. Its interval includes zero. This diagnostic concerns characteristics measured before discussion; T2 and gain can themselves respond to group composition. A causal interpretation of peer coefficients also requires evidence about the assignment process. Group knowledge after deliberation is difficult to interpret because groupmates share the same experience [@manski1993].

# Conclusion

Participants' factual knowledge increases across the polls we study. Reporting absolute gains in percentage points and increases relative to initial scores makes the magnitude interpretable, while adjustment for guessing addresses one source of measurement error. In the two polls with common before-and-after control interviews, attendees show larger gains than controls. The comparisons remain subject to selection into attendance, and the peer regressions are descriptive. The knowledge questions concern facts selected by organizers, so these estimates describe learning about the topics the events emphasized.

\FloatBarrier
# References {-}

<div id="refs"></div>

\clearpage
\appendix
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

# The Deliberative Polls {#app:polls}

These `r nrow(polls)` polls supply the attendee sample throughout the paper. Three also provide control comparisons, including Northern Ireland's T3-only contrast. Tables \ref{tab:polls} and \ref{tab:poll-gains} list polls in the same chronological order.

```{r polls, results = "asis"}
polls |>
  dplyr::transmute(
    poll, year,
    mode = dplyr::if_else(mode == "Online", "Online", "In person"),
    control = dplyr::if_else(control_group, "Yes", "No")
  ) |>
  purrr::set_names(c("Poll", "Year", "Mode", "Control")) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lclc",
    caption = "Polls with respondent-level knowledge-item answers at both waves."
  ) |>
  print()
```

\clearpage
# Further results {#app:further}

```{r poll-gains}
gains |>
  dplyr::arrange(match(poll_id, polls$poll_id)) |>
  dplyr::transmute(
    Poll = pollname, `Initial (%)` = pp(k1_mean),
    `Gain (pp)` = pp(raw),
    `Adjusted (pp)` = pp(adjusted)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = c(">{\\raggedright\\arraybackslash}p{2.8in}", "r", "r", "r"),
    caption = "Initial knowledge (percent correct), observed gains, and guessing-adjusted learning (percentage points) across the analyzed polls.")
```

\begin{minipage}{\textwidth}

```{r meta}
meta |>
  dplyr::filter(!grepl("SD|relative", model)) |>
  dplyr::mutate(
    Weighting = dplyr::if_else(grepl("precision weighted", model), "Precision", "Equal"),
    Quantity = dplyr::case_when(
      parameter == "face_to_face" ~ "Observed gain: in person (pp)",
      parameter == "online" ~ "Observed gain: online (pp)",
      grepl("guessing", model) ~ "Guessing-adjusted learning (pp)",
      TRUE ~ "Observed gain (pp)"
    )
  ) |>
  dplyr::transmute(Quantity, Weighting, Polls = polls, Estimate = pp(estimate),
                   `95% interval` = paste(pp(lower), pp(upper), sep = " to ")) |>
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "",
    position = "H", caption = "Average learning across polls, with hierarchical-bootstrap confidence intervals.")
```

Observed gain is T2 minus T1; guessing-adjusted learning estimates the percentage who learned an item, averaged across questions. Equal weighting gives each poll the same weight. Precision weighting uses $w_d=1/\widehat{\mathrm{Var}}(\hat\theta_d)$, with sampling variances estimated by resampling discussion groups within each poll. The weighted average is $\sum_d w_d\hat\theta_d/\sum_d w_d$. These weights are held fixed when resampling polls and groups for its interval. The mode rows split the equal-poll average into in-person and online polls.

\end{minipage}

```{r hetero}
hetero |>
  dplyr::mutate(
    comparison = dplyr::if_else(moderator == "k1", "Initial score +10 pp", "Bachelor’s degree"),
    multiplier = dplyr::if_else(moderator == "k1", 10, 100)
  ) |>
  dplyr::transmute(Study = study, Comparison = comparison,
    `Difference (pp)` = num(multiplier * interaction, 1),
    `95% interval` = paste(num(multiplier * lower, 1), num(multiplier * upper, 1), sep = " to "),
    N = count(n)) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.8in}", ">{\\raggedright\\arraybackslash}p{1.35in}", ">{\\raggedleft\\arraybackslash}p{0.75in}",
      ">{\\raggedleft\\arraybackslash}p{0.85in}", ">{\\raggedleft\\arraybackslash}p{0.7in}"),
    caption = "How the baseline-adjusted attendee--control difference varies with initial knowledge and education.")
```

Table \ref{tab:hetero} reports changes in the attendee--control difference, not changes in everyone's knowledge. N is the regression sample size. Degree compares a bachelor’s degree with no degree. A negative entry means the attendee advantage is smaller for people with higher initial knowledge or a bachelor’s degree. For example, the America in One Room initial-knowledge estimate of `r num(10 * het("America in One Room 2019", "k1")$interaction, 1)` percentage points compares two people whose starting scores differ by ten percentage points. Each regression includes attendance, initial score, the listed characteristic, and its interaction with attendance. Centering initial scores at their sample mean does not change the reported interaction. All intervals use the hierarchical bootstrap. Selection into attendance and score ceilings limit causal interpretation.

## Additional participant characteristics

Age, education, and gender are harmonized for all `r size("core", "polls")` polls. The attitude models require additional indices, which are harmonized for `r size("historical", "polls")` polls.

*Education*. Education, qua years of schooling, is distinct from, and only moderately correlated with, political knowledge. Controlling for cognitive ability, with which it is more strongly correlated, its effect on knowledge appears negligible [@luskin1990]. But absent a separate measure of cognitive ability, it can roughly proxy the latter. Naturally, we expect it to have a positive effect. There may, however, be some mutual substitution with initial knowledge. People who know enough already may not need to be as bright to learn a lot, and people who are bright enough may not need to know as much to do so. We measure education as comparably as possible across polls in three categories: below secondary completion, secondary/some college, and a degree or higher.

*Gender and age*. These demographic variables exhibit real-world political knowledge gaps, with men knowing more than women and older people more than younger ones [@dellicarpini1996]. Age is measured in decades and gender as a dummy scored 1 for women.

*Proportion of women*. Whether from greater comfort, more widely shared cognitive or communication style, or lower stereotype threat [@steele1997], women may learn more in small groups containing more women. We calculate the proportion excluding the participant herself.

*Heterogeneity of opinion*. In the real world, people often know congenial but not uncongenial facts---those commonly cited by those sharing their views but not those commonly cited by those holding opposing views. Part of deliberation's effect on learning may consist in getting people to note and accept (if, in many cases, still argue around) uncongenial facts. This can be expected to occur mainly on site. Participants learn a great deal in anticipation of the event, but the learning at that stage is mainly of congenial facts [@hansen2004]. On site, the participants find it harder to avoid confronting the uncongenial ones. This may happen more, the more opposing points of view are represented and argued in the small group---the more attitudinally heterogeneous the group is. We gauge this as the generalized variance, the determinant of the covariance matrix, of the group's policy attitudes [@anderson2003].

*Attitude extremity*. Participants starting with more extreme views may have shuttered out more uncongenial facts, leaving them with more ground for deliberation to make up. We measure extremity as the mean, across attitude indices $j$, of $|A_{ij} - .5|$, where $A_{ij}$ is the $i$th respondent's score on the $j$th index.

*Reading the briefing materials*. One of the main information vectors is the briefing document sent to participants shortly after recruitment, normally well in advance of the on-site deliberations. It provides some background facts and lays out the major arguments for and against given policy options, which embed additional facts congenial to each argument. Respondent-linked reading reports are available for `r sum(coverage$reading > 0)` polls. They describe reading before discussion, recalled in the post-deliberation questionnaire. We score their ordered categories from zero to one using the source-specific response scales and analyze them in a separate model. The reading model uses the demographic specification with poll fixed effects and a small-group random intercept.

\clearpage
## Models with additional controls

```{r historical-models}
historical_models <- model_table(
  models, c("historical", "briefing"),
  c("Additional controls", "Reading reports")
)
separators <- rep("", nrow(historical_models))
summary_start <- match("Participants", historical_models[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(historical_models) - 1L)] <- "\\addlinespace[0.25em]"
knitr::kable(
  historical_models, format = "latex", booktabs = TRUE,
  longtable = TRUE, align = "lrr",
  linesep = separators,
  caption = "Change in proportion correct (T2 minus T1) with additional controls. Coefficients with 95 percent hierarchical-bootstrap confidence intervals."
)
```

\noindent{\footnotesize\raggedright\emph{Notes.} The model with additional controls includes poll and small-group random intercepts; the reading-reports model adds reading to the demographic specification, using poll fixed effects and small-group random intercepts. Attitude measures are omitted from that model. Adding groupmates' T1 correctness on questions the respondent missed to the model with additional controls gives `r term("items", "group_k1_items")` among `r count(size("items", "n"))` participants. Respondents with no missed T1 items receive zero on this measure and a separate indicator (`r term("items", "no_missed_items")`); none is removed for lack of an opportunity to learn. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts.\par}

## Matched-sample comparisons

Table \ref{tab:sample-comparison} separates sample restrictions from added covariates. Column 1 repeats the baseline model in all polls. Column 2 restricts it to the polls with the additional attitude measures. Columns 3 and 4 use exactly the same participants: column 3 retains the baseline specification, and column 4 adds the controls shown in Table \ref{tab:historical-models}. Thus the final comparison holds sample composition fixed.

\begin{minipage}{\textwidth}
\small
\setlength{\tabcolsep}{3pt}

```{r sample-comparison}
common_terms <- models$term[models$model == "core"]
comparison <- model_table(
  dplyr::filter(models, term %in% common_terms),
  c("core", "core_21", "core_21_complete", "historical"),
  c("All polls", "21 polls", "Same cases", "Added controls")
)
comparison[-1] <- lapply(comparison[-1], function(column) {
  vapply(column, function(cell) {
    parts <- strsplit(cell, " [", fixed = TRUE)[[1]]
    if (length(parts) == 1L) return(cell)
    paste0("\\shortstack[r]{", parts[1], " \\\\ {[", parts[2], "}}")
  }, character(1))
})
comparison_lines <- rep("\\addlinespace[0.15em]", nrow(comparison))
comparison_lines[match("Participants", comparison[[1]]) - 1L] <- "\\midrule"
knitr::kable(comparison, format = "latex", booktabs = TRUE, position = "H", escape = FALSE,
  linesep = comparison_lines, align = "lrrrr",
  caption = "Baseline coefficients under sample restrictions and additional controls. 95 percent hierarchical-bootstrap confidence intervals appear below coefficients.")
```

\end{minipage}

The peer-knowledge coefficient is `r term("core_21", "group_k1")` in the 21-poll baseline model. Among the complete cases, it is `r term("core_21_complete", "group_k1")` before adding controls and `r term("historical", "group_k1")` afterward. The change from added controls is small relative to the uncertainty in these estimates.

The corresponding check for the demographic model uses its `r count(size("demographic", "n"))` participants. On these same cases, the baseline peer-knowledge coefficient is `r term("core_demographic_sample", "group_k1")`, compared with `r term("demographic", "group_k1")` after adding demographics.

## Additional control comparisons {#app:controls}

Regressing T2 scores on attendance and T1 scores answers a different question from comparing gains. The baseline-adjusted attendee--control differences are `r pp(a1r$estimate)` percentage points in America in One Room (`r pp(a1r$lower)` to `r pp(a1r$upper)`), and `r pp(climate$estimate)` in the climate poll (`r pp(climate$lower)` to `r pp(climate$upper)`). These intervals use the same bootstrap as the main comparisons.

A year after the climate poll, the baseline-adjusted gap is `r pp(climate_t3$estimate)` percentage points (`r pp(climate_t3$lower)` to `r pp(climate_t3$upper)`). Northern Ireland has only a follow-up control interview: `r ni$n_treated` returning attendees scored `r pp(ni$estimate)` percentage points above `r ni$n_control` controls (`r pp(ni$lower)` to `r pp(ni$upper)`). This is a cross-sectional comparison without a control baseline. We do not apply the immediate-learning model's no-forgetting assumption to these later outcomes.

## Poll-specific peer associations

The following estimates fit a separate slope in each poll, conditional on own initial knowledge. Their equal-poll average is `r interval(peer("k1")$estimate, peer("k1")$lower, peer("k1")$upper)`.

```{r peers, fig.cap = "Association between the other group members' mean T1 knowledge and own knowledge gain, by poll, with 95\\% intervals. Each estimate controls for own T1 knowledge. Intervals use the hierarchical bootstrap. The pooled estimate weights polls equally and resamples polls within mode and then their groups. These are descriptive associations.", fig.pos = "tbp", out.width = "100%"}
figure("peer_effects")
```

## Baseline composition of discussion groups

We regress each participant's baseline characteristic on its mean among the other group members, controlling for the mean among other members of the poll and poll fixed effects [@guryan2009]. This is a descriptive sorting diagnostic; it does not reconstruct the original assignment mechanism. A randomization test would require the original assignment pools and restrictions, including any scheduling or language constraints. Balance on observed covariates alone does not establish random assignment.

```{r baseline-composition}
assignment |>
  dplyr::transmute(
    Characteristic = dplyr::recode(covariate, k1 = "Initial knowledge", female = "Female", age = "Age", education_ba = "Degree or higher"),
    Association = interval(estimate, lower, upper), Participants = count(n)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "", caption = "Baseline composition of discussion groups. Associations with 95 percent hierarchical-bootstrap intervals.")
```

\FloatBarrier
# Knowledge items {#app:items}

The catalog lists `r nrow(item_catalog)` questions for the same `r nrow(gains)` polls used in the learning analysis throughout the paper [@dpdata]. Wording is paraphrased where only source labels survive; capitalization and unambiguous abbreviations are normalized for readability. Original labels and scoring rules are preserved in the cited dataset. Choices are transcribed where recoverable; truncated or unavailable options are identified. Notes explain open-response coding.

```{r items, results='asis'}
cat(item_appendix_markdown(items = item_catalog))
```
