---
title: 'Deliberation and Learning: Evidence From Deliberative Polls\thanks{Data and replication materials: \url{https://github.com/soodoku/dp-learning}.}'
author: "Robert C. Luskin, Gaurav Sood, and James S. Fishkin"
date: "Working draft"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    fig_caption: true
bibliography: [references.bib, data-version.bib]
link-citations: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{placeins}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H")
options(knitr.graphics.rel_path = FALSE)
source("R/paper.R")
figure <- \(name) knitr::include_graphics(normalizePath(file.path("figs", paste0(name, ".pdf"))))
count <- \(x) prettyNum(x, big.mark = ",")
prose_count <- \(n) if (n %in% 0:9) {
  c("zero", "one", "two", "three", "four", "five", "six", "seven", "eight", "nine")[n + 1L]
} else count(n)
gains <- read_out("poll_gains.csv")
gaps <- read_out("knowledge_gaps.csv")
retention <- read_out("retention.csv")
meta <- read_out("meta.csv")
controlled_learning <- read_out("control_learning.csv")
controlled_pooled <- read_out("control_learning_pooled.csv")
provenance <- jsonlite::read_json("tabs/provenance.json", simplifyVector = TRUE)
models <- read_out("models.csv")
sensitivity <- read_out("knowledge_sensitivity.csv")
effects <- read_out("control_effects.csv")
hetero <- read_out("control_heterogeneity.csv")
peers <- read_out("peer_effects_pooled.csv")
selected <- read_out("selection.csv")
assignment <- read_out("assignment_check.csv")
coverage <- read_out("covariate_coverage.csv")
attrition_changes <- read_out("attrition_contrasts.csv")
attrition_flow <- read_out("attrition_flow.csv")
attrition_means <- read_out("attrition_means.csv")
phase_changes <- read_out("phase_contrasts.csv")
phase_coverage <- read_out("phase_coverage.csv")
phase_selection <- read_out("pre_arrival_selection.csv")
phase_provenance <- jsonlite::read_json("tabs/phase_provenance.json", simplifyVector = TRUE)
phase_balanced <- dplyr::filter(phase_changes, sample == "all_three_observed_attendees", n_people > 0)
phase_pairs <- dplyr::filter(phase_changes, sample == "available_paired_attendees", n_people > 0,
                             study_id != "btp-primaries-2004" | poll_id == "btp-online-primaries-2004")
phase_baseline_exit <- dplyr::filter(phase_pairs, contrast == "post_minus_pre_arrival")
phase_row <- function(poll, contrast) {
  row <- dplyr::filter(phase_balanced, poll_id == poll, .data$contrast == .env$contrast)
  stopifnot(nrow(row) == 1L)
  row
}
phase_selection_row <- function(poll, reference) {
  row <- dplyr::filter(phase_selection, source_dataset == "control", poll_id == poll,
                      analysis == "baseline_selection_difference", category == reference)
  stopifnot(nrow(row) == 1L)
  row
}
marousi_exit <- phase_row("marousi-2006", "post_minus_arrival")
marousi_before <- phase_row("marousi-2006", "arrival_minus_pre_arrival")
marousi_total <- phase_row("marousi-2006", "post_minus_pre_arrival")
europe_exit <- phase_row("tomorrows-europe-2007", "post_minus_arrival")
europe_before <- phase_row("tomorrows-europe-2007", "arrival_minus_pre_arrival")
europe_total <- phase_row("tomorrows-europe-2007", "post_minus_pre_arrival")
a1r_recruitment <- phase_selection_row("america-in-one-room-2019", "attended_minus_recruitment_nonattender")
climate_invited <- phase_selection_row("a1r-climate-2021", "completed_minus_invited_noncompleter")
a1r_baseline_control <- phase_selection_row("america-in-one-room-2019", "attended_minus_control")
climate_baseline_control <- phase_selection_row("a1r-climate-2021", "completed_minus_control")
amr_baseline_control <- phase_selection_row("amr-2024", "attended_minus_control")
balance_knowledge <- assignment[assignment$covariate == "k1", ]
polls <- read_out("polls.csv")
main_interviews <- read_out("main_interviews.csv")
item_catalog <- read_item_catalog() |> dplyr::filter(poll_id %in% gains$poll_id)

a1r <- control_row(effects, "America in One Room 2019", "Attended vs uninvited control")
climate <- control_row(effects, "America in One Room: Climate 2021", "Completed vs uninvited control")
climate_t3 <- control_row(effects, "America in One Room: Climate 2021", "Completed vs control, one year later")
ni <- control_row(effects, "Northern Ireland 2007", "T3 attendee vs control, no baseline")
pp <- function(x) num(100 * x, 1)
pooled <- function(model) meta[meta$model == model & meta$parameter == "mu", ]
control_gain <- function(id, metric) controlled_learning[controlled_learning$poll_id == id & controlled_learning$term == metric, ]
control_average <- function(metric) controlled_pooled[controlled_pooled$term == metric, ]
persist <- round(100 * climate_t3$estimate / climate$estimate)
het <- \(study, moderator) hetero[hetero$study == study & hetero$moderator == moderator, ]
peer <- \(p) peers[peers$peer == p, ]
term <- \(model, t) model_term(models, model, t)
size <- \(model, what) models[[what]][models$model == model][1]
reading <- models[models$model == "briefing" & models$term == "read_briefing", ]
gap <- function(comparison, term) {
  row <- dplyr::filter(gaps, .data$comparison == .env$comparison, .data$term == .env$term)
  stopifnot(nrow(row) == 1L)
  row
}
initial_correct <- mean(gains$k1_mean)
later_correct <- mean(gains$k2_mean)
initial_headroom <- 1 - initial_correct
headroom_closed <- (later_correct - initial_correct) / initial_headroom
stopifnot(
  initial_headroom > 0,
  isTRUE(all.equal(later_correct - initial_correct, pooled("raw, pooled")$estimate))
)
```

\begin{abstract}
We study factual learning among `r count(size("core", "n"))` participants in `r size("core", "groups")` discussion groups across `r size("core", "polls")` Deliberative Polls held between `r min(polls$year)` and `r max(polls$year)`. The polls address elections, energy, health, crime, climate, and other public issues in Europe, Australia, China, and the United States. Across polls, absolute gains averaged `r pp(pooled("raw, pooled")$estimate)` percentage points and relative gains `r pp(pooled("relative, pooled")$estimate)`\%. An item model assuming no forgetting yields estimated learning of `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points. In two polls with control groups, attendees gained an average of `r pp(control_average("raw")$estimate)` percentage points more than controls. In `r prose_count(dplyr::n_distinct(phase_balanced$poll_id))` studies, we compare knowledge before arrival, on arrival, and after deliberation. Substantial gains occur before participants arrive.
\end{abstract}

# Introduction

Among the great virtues commonly ascribed to deliberation is learning. The participants learn about the topics under discussion and, to a lesser degree, incidentally, about related ones. They learn not only relevant facts but the arguments for and against particular alternatives and what other people unlike themselves, their circumstances, and their ways of thinking are like. This is helpful, both at the individual level, for discerning what policies most closely accord with one's values and interests and voting accordingly, and at the aggregate level, for forming "authentic" majorities---majorities that "know what they are doing" [@luskin2003]. How much citizens know shapes what they prefer and how they vote [@converse1964; @bartels1996; @althaus1998; @dellicarpini1996], even on foreign policy [@aldrich1989; @foyle2004], though information shortcuts can partly stand in for knowledge [@popkin1991; @lupia1994].

The expectation of learning follows directly from the nature of "deliberation," conventionally construed as discussion, but not just any discussion [@fishkin2005]. While definitions vary, deliberation is generally viewed as normatively elevated discussion---in which, among other things, the participants clarify and refine their preferences by open-mindedly weighing competing arguments. As part of that weighing, they absorb factual information and sift out misinformation.

Or so the theory runs. But how much do participants actually learn, is it the deliberation that teaches them, and who learns the most? We examine learning of relevant facts. This is not the only sort of learning involved, but it is the easiest to measure, and it is both important in itself and positively correlated with the rest.

We compare participants' initial and later knowledge, using one sample for the main learning estimates and discussion-group models. Interview timing varies across studies. Where measurements before arrival, on arrival, and after deliberation are available, we separately measure changes before arrival and during the event among the same participants. We also compare pre-arrival knowledge by invitation and attendance to assess selection into the observed participant sample. In America in One Room, we compare attendees with nonattenders in the recruitment sample and with a separate control group not invited to deliberate. In the climate study, the corresponding participant group consists of people classified as completing the deliberations; other invitees may have participated in some sessions. Before-and-after knowledge scores are available for these participants and controls; the remaining recruitment or invitation sample contributes only initial scores. Earlier studies document learning in Deliberative Polls and other forums [@luskin2002b; @barabas2004; @sturgis2005; @esterling2011; @theuwis2025]. We report gains in both absolute and relative terms, adjust for guessing, and examine whether participants learn more alongside more knowledgeable groupmates.

# Deliberative Polling and Learning

Deliberative Polling [@fishkin1991; @fishkin1995; @luskin2002b] aims to capture fuller-information, more thoughtful opinions than those expressed in ordinary polls. In very brief, a Deliberative Poll interviews and assembles, as many as are willing of, a random sample to deliberate some policy or electoral choice, in alternating small group discussions and plenary question-and-answer sessions with panels of competing experts, and then re-measures the sample's opinions at the end. These latter are frequently different, both at the individual level and in the aggregate, from what they were before deliberating. The absolute net change is statistically significant nearly three-quarters of the time and averages nearly .1 on a 0--1 scale [@luskin2007b].

This design, admittedly a particular embodiment of deliberation, affords additional opportunities and incentives to learn. The participants are provided with carefully balanced briefing documents, rehearsing the arguments each side would make. There is usually a section of basic background information, and the arguments embed each side's pet facts---those most congenial to its arguments. In addition, the participants know from the moment they are first recruited that they will be attending the deliberation and that the deliberation (typically) will be an "Event," capital "e"---away from home, usually televised or taped for later broadcast, often attended by high-profile policy experts or policy-makers in the role of observers or panelists. Thus, motivated by a desire to "do well," many participants use the weeks between their recruitment and the event itself to prepare---not only by reading the briefing materials but by paying heightened attention to stories about the topic in the media, by talking more with friends, family, and coworkers about it, and even, in some cases, by researching it in the library or online.

So if knowledge, like its close relative "sophistication," depends on the confluence of cognitive ability, motivation, and exposure to relevant information [@luskin1990; @luskin2002a], the deliberation in Deliberative Polling stands to increase two of those three. There is not much to be done, in adulthood, about ability---and in fact we should want the participants to be representative in this, as in all other respects. But the treatment can and does affect motivation and exposure. Judging from past polls, that is enough to produce considerable, sometimes dramatic learning [@luskin2002b; @list2013].

The online design is intended to be as similar as possible to the face-to-face one, given the difference in mode. The earlier online polls used voice discussion to capture as much as possible of deliberation's social elements.^[More recent online Deliberative Polls use live video discussion. The 2021 America in One Room climate poll used the Stanford Online Deliberation Platform, with an automated moderator managing speaking turns and the discussion agenda [@fishkin2025scaling, pp. 436--437].] In these earlier polls, participants still "met" in small groups, led by trained moderators, and questioned panels of policy experts and policy makers. They still receive briefing materials. They are still a representative sample---sometimes randomly sampled offline, with "nonusers" receiving free access, including free equipment, to get them online, sometimes drawn from online panels and matched to approximate a random sample. On the other hand, deliberating by voice, even by video, online can never capture the full richness and complexity of in-person interactions; the discussions are interrupted, spread over a period of weeks, with much greater possibility of real-world intrusion, and the assignment to small groups is perforce constrained by possible meeting times. All these differences may act to weaken the online treatment's effects [@luskin2006; @iyengar2005].

# Theory and hypotheses

Learning requires exposure to information and the ability and motivation to absorb it. These mechanisms suggest when participants should learn as well as who should learn most.

*When learning occurs*. We expect some learning before participants arrive. The briefing materials provide information, and the prospect of discussing the issues gives participants a reason to absorb it. Invitation can also prompt them to seek information elsewhere. Thus preparation should raise knowledge between recruitment and arrival, even among some invitees who ultimately do not attend.

We expect further learning during deliberation. Discussion gives participants opportunities to ask questions, encounter facts they missed, and correct misunderstandings. Expert answers and explanations from other participants can make difficult material easier to grasp. Preparation may help participants learn from these exchanges by giving them the background needed to understand them. But it also leaves less room for further gains on a bounded knowledge test. We therefore expect gains both before arrival and during the event, without predicting which will be larger. Smaller gains during the event need not mean that discussion offers little: they may partly reflect what participants have already learned.

After the event, the question becomes how much of the gain lasts. If participants stop encountering or using the information, we expect some of it to be forgotten. Continued attention and discussion may instead sustain knowledge or add to it. Later interviews therefore address the durability of the gains. Changes since exit reflect both forgetting and subsequent learning, rather than forgetting alone.

The same mechanisms motivate our expectations about individual and group differences.

*Initial knowledge*. The information-rich tend to get information-richer. The more one knows going in, the more one tends to learn and thus know going out. Absent ceiling effects, this predicts a positive association with gain; bounded scores can instead produce a negative association. Prior knowledge on a particular topic improves people's ability to comprehend new information, enabling them to extract the central elements of a message and draw appropriate inferences efficiently [@eckhardt1991; @recht1988]. Prior knowledge also enhances people's ability to store new information on that topic and retrieve it later [@cooke1993; @fiske1990; @hambrick2003; @mcgraw1990; @recht1988; @schneider1993]. So the less political information individuals have stored in memory, the more difficult it is for them to acquire new information.

*Education*. Education, qua years of schooling, is distinct from, and only moderately correlated with, political knowledge. Controlling for cognitive ability, with which it is more strongly correlated, its effect on knowledge appears negligible [@luskin1990]. But absent a separate measure of cognitive ability, it can roughly proxy the latter. Naturally, we expect it to have a positive effect. There may, however, be some mutual substitution with initial knowledge. People who know enough already may not need to be as bright to learn a lot, and people who are bright enough may not need to know as much to do so.

*Gender and age*. These demographic variables exhibit real-world political knowledge gaps, with men knowing more than women and older people more than younger ones [@dellicarpini1996]. These differences can reflect prior opportunities to acquire information. Conditional on initial knowledge, their association with learning during deliberation need not have the same direction. We include both as participant characteristics without imposing a directional hypothesis.

*Gender composition*. Whether from greater comfort, more widely shared cognitive or communication style, or lower stereotype threat [@steele1997], women may learn more in groups containing more women. We examine this possibility by interacting gender with the proportion of female groupmates.

*Reading the briefing materials*. Reading provides exposure to the facts and competing arguments discussed at the event. We therefore expect participants who read more of the materials to learn more, although motivation may influence both reading and learning.

*Group knowledge*. More knowledgeable groupmates can supply facts, correct errors, and explain unfamiliar arguments. We therefore expect their initial knowledge to be positively associated with a participant's gain, conditional on that participant's own initial knowledge.

*Group size and mode*. Larger groups may offer more perspectives but give each participant less time to speak [@latane1979], leaving the expected association with learning ambiguous. Online deliberation may offer thinner exposure to other participants and weaker opportunities to learn than in-person discussion [@luskin2006].

*Attitudes*. Strong policy preferences may motivate attention to relevant information or make conflicting information harder to accept. Disagreement within a group may expose participants to unfamiliar arguments, but it can also impede communication. We examine attitude extremity and group disagreement without imposing a directional prediction.

# Research design and model

## Interview timing and the quantities of interest

A Deliberative Poll begins by recruiting citizens, usually through a survey conducted before they receive briefing materials. Those who participate can prepare before arriving, then deliberate in small groups and question experts. Knowledge may be measured at four stages: before arrival ($t_0$), on arrival before discussion ($t_1$), at the end of deliberation ($t_2$), and at a later follow-up ($t_3$). For online polls, arrival means the start of the discussion sessions. The labels refer to these stages in every poll; a missing arrival questionnaire does not turn a pre-arrival interview into $t_1$. A questionnaire administered partway through deliberation, as in New Haven, is an interim measurement.

These stages distinguish three quantities. The change $K_{t_1}-K_{t_0}$ measures gains before arrival, when participants may read the briefing materials or prepare in other ways. The change $K_{t_2}-K_{t_1}$ measures gains during the event. The change $K_{t_2}-K_{t_0}$ includes both periods. Follow-up measurements at $t_3$ address how much knowledge remains later. Each is a change among the same people answering the same questions. None by itself identifies a causal effect: repeated testing and learning outside the study can also change scores.

Our main pooled estimate is the average within-poll change among attendees, giving each poll equal weight. It compares pre-arrival and exit ($t_0$ to $t_2$), covering both preparation and the event. The separate three-wave comparisons distinguish gains before arrival from gains during deliberation. Later follow-ups address how much knowledge persists and are analyzed separately.

## Invitation, attendance, and causal effects

The two studies with controls, America in One Room and its climate successor, allow us to subtract controls' pre-arrival-to-exit change from attendees' change. This difference removes changes common to both groups. It identifies the effect of participation on attendees only if, without participation, attendees and controls would have had the same average change. Initial differences in knowledge alone do not violate that assumption, but differences in how the two groups would otherwise learn could do so.

An intention-to-treat (ITT) effect concerns invitation, including any preparation it induces before the event. With randomized invitations and follow-up outcomes for all invitees and controls, it could be estimated without assuming that nonattenders learned nothing. Our participant-control comparisons do not estimate that effect: follow-up knowledge scores are unavailable for the remaining recruitment or invitation sample, and the 2019 recruitment indicator does not establish individual invitation status. Assigning nonattenders zero gains would be an assumption about missing outcomes. Assuming instead that invitation had no effect on their learning would still allow the gains they would have made without it.

Under random assignment, an invitation that increases attendance, no interference between participants, and the exclusion and monotonicity assumptions, dividing the ITT effect by the invitation-induced change in attendance identifies a local average treatment effect (LATE) for people whose attendance is induced by invitation [@angrist1996]. The exclusion assumption requires invitation to affect knowledge only through attendance. That is questionable here. The climate study sent electronic briefing materials to registered delegates before deliberation, so recipients could read them without eventually attending [@norc2021, pp. 5--6]. An invitation could also prompt independent preparation. Such learning belongs in the effect of invitation but prevents interpreting the usual instrumental-variable ratio as the effect of attendance alone.

## Model {#sec:model}

Let $K_{igd,t}$ denote the proportion of knowledge questions answered correctly by participant $i$ in group $g$ and poll $d$ at stage $t$. Our outcome is the gain from pre-arrival to exit, $G_{igd}=K_{igd,t_2}-K_{igd,t_0}$. We fit a hierarchical linear model [@raudenbush2002]:
$$G_{igd}=\beta_0+\beta_1K_{igd,t_0}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd},$$
where $\mathbf{z}_{igd}$ contains groupmates' mean initial knowledge, group size, and mode. The second specification adds education, age, gender, the proportion of female groupmates, and the education-by-initial-knowledge and gender-by-group-composition interactions. The third and fourth specifications separately extend the demographic model: the third adds attitude extremity and group disagreement, and the fourth adds reported briefing reading. The $\beta$ coefficients are fixed regression parameters. For fitting, we assume
$$\alpha_{gd}\sim N(0,\sigma_g^2),\qquad \gamma_d\sim N(0,\sigma_p^2),\qquad \varepsilon_{igd}\sim N(0,\sigma_e^2),$$
with these components mutually independent conditional on the regressors and independent across distinct groups, polls, and individuals, respectively. The group and poll random intercepts allow outcomes to covary within discussion groups and across groups in the same poll. Their variances, $\sigma_g^2$ and $\sigma_p^2$, measure residual variation at those levels; $\sigma_e^2$ measures individual residual variation. We estimate the coefficients and variance components by maximum likelihood [@bates2015] and obtain confidence intervals from the hierarchical bootstrap.

Adding $K_{igd,t_0}$ to both sides gives the equivalent equation for later knowledge:
$$K_{igd,t_2}=\beta_0+(1+\beta_1)K_{igd,t_0}+\mathbf{z}_{igd}'\boldsymbol{\beta}_z+\alpha_{gd}+\gamma_d+\varepsilon_{igd}.$$
Thus using gain instead of the later score shifts the coefficient on initial knowledge down by one and leaves the other coefficients and fitted residuals unchanged. We report gain because it directly expresses the change the paper seeks to explain. Conditioning on initial knowledge compares that change among participants with the same measured starting knowledge.

An unadjusted change-score model instead constrains the baseline coefficient to zero in the gain equation. These specifications can disagree when groups differ at baseline [@allison1990; @vanbreukelen2006]. Score ceilings, guessing, and measurement error complicate the relationship between observed gains and learning of the underlying facts [@luskin2016]. We therefore treat the initial-knowledge coefficient as an association, not an estimate of how prior knowledge causes learning.

The comparisons follow these distinctions. We first describe who attends and how attendees differ from the comparison groups before the event. We then report absolute and relative knowledge gains, compare attendees with controls, distinguish gains before arrival from gains during the event, and examine how much knowledge persists at later follow-ups. Finally, we use the model to assess the hypotheses about participants' initial knowledge, education, and other characteristics; the knowledge and composition of their discussion groups; and their attitudes and reading of the briefing materials.

# Data and measures

The main sample comprises `r size("core", "polls")` polls and `r count(size("core", "n"))` attendees with initial and later knowledge scores and recorded discussion-group membership. It supports both the learning estimates and the analysis of discussion groups. We use the Deliberative Poll dataset [@dpdata], linking each participant's answers across interviews. Each person contributes one paired observation within a poll; overlapping datasets for the same study do not supply additional participants. Appendix \ref{app:items} gives the questions and scoring rules.

The two datasets for the 2004 online Primaries describe the same experiment and include overlapping respondents. We include that study once, using attendees with at least one documented discussion session and observed initial and exit questionnaires. This gives 239 paired attendees, of whom 238 have known groups and enter the main models. The 217-person historical sample is an overlapping subset of that study.

Comparisons by interview timing require documented attendance, evidence that respondents answered each questionnaire, and the same knowledge questions and scoring denominator at both interviews. The main analysis requires observed questionnaires at every stage through immediate exit that the poll collected; a stage never collected does not exclude a respondent. An unanswered knowledge item within a completed questionnaire receives zero credit, but an absent or unavailable questionnaire does not receive a zero score. For the National Issues Convention, both the main analysis and the timing comparisons require observed pre-arrival and exit questionnaires. Pair-specific comparisons include attendees without known discussion groups when their attendance and questionnaire responses are documented; the common three-wave comparisons require recorded group membership. We compare pre-arrival and exit knowledge in `r dplyr::n_distinct(phase_baseline_exit$study_id)` polls and all three measurements in `r prose_count(dplyr::n_distinct(phase_balanced$poll_id))`. Appendix \ref{app:phases} reports the available pairs and their sample sizes.

America in One Room and the climate poll interviewed separate samples who were never invited to deliberate [@fishkin2021]. We compare changes in their scores with changes among participants in the main sample, using the same questions in both arms. The climate sample identifies completion of the deliberations rather than attendance at any session. These comparisons may reflect selection into attendance or completion as well as deliberation. Additional control comparisons and follow-ups appear in Appendix \ref{app:controls}.

## Knowledge

We measure knowledge with factual questions, mostly multiple choice or true-false, along with some numerical estimates and questions about party or candidate positions [@luskin1987; @luskin2005; @price1999; @mondak1999; @mondak2001]. Appendix \ref{app:items} lists the items. Correct responses are scored as 1, incorrect responses and "don't knows" as 0 [@luskin2011], and these scores are then averaged across items to give the proportion answered correctly.

The analysis includes `r count(size("core", "n"))` participants in `r size("core", "groups")` groups and `r size("core", "polls")` polls. The demographic model retains `r count(size("demographic", "n"))` participants and all `r size("demographic", "polls")` polls; `r count(size("core", "n") - size("demographic", "n"))` participants have missing covariates. The reading model includes `r count(size("briefing", "n"))` participants in `r size("briefing", "polls")` polls. The attitude model retains `r count(size("attitudes", "n"))` participants in all `r size("attitudes", "polls")` polls; `r count(size("demographic", "n") - size("attitudes", "n"))` additional participants lack attitude measures.

## Who attends? {#sec:selection}

Participation is voluntary, so participants may differ from the people invited or recruited to attend. Their initial knowledge helps characterize whom the gain estimates describe. Table \ref{tab:pre-arrival-selection} compares initial scores for attendees in America in One Room and completers in the climate study with the remaining recruitment or invitation sample and with people not invited to deliberate. A positive gap means that the analyzed participants started with more knowledge; it does not measure how much they learned.

Attendees in America in One Room and completers in the climate study had higher initial scores than both comparison groups. In America in One Room, the pre-arrival gaps are `r pp(a1r_recruitment$estimate)` and `r pp(a1r_baseline_control$estimate)` percentage points, respectively. In the climate poll they are `r pp(climate_invited$estimate)` and `r pp(climate_baseline_control$estimate)` points (Table \ref{tab:pre-arrival-selection}). In 2019, the recruitment sample was subsequently subsampled for invitations, so we cannot identify every nonattender as someone who declined an invitation. In the climate study, baseline respondents in the treatment sample were invited to register [@norc2019, p. 6; @norc2021, p. 4]. These summaries use everyone with an observed baseline and the relevant attendance or completion classification, including people outside the grouped-model sample. Other climate invitees include partial attendees, people with documented zero sessions, and people whose nonattendance is inferred from an absent post questionnaire. The completion comparison retains all other invitees. These comparisons therefore have different denominators from the paired gain and group-model samples.

The six-country antimicrobial-resistance study supplies an additional baseline comparison [@mendelson2026]. Its 1,280 participants scored `r pp(amr_baseline_control$estimate)` percentage points above 1,139 controls before invitations. This unweighted comparison describes the survey respondents across six countries. We lack observations for the 1,847 invitees who did not attend. Discussion-group identifiers are unavailable, so this study does not enter the group models or the group-bootstrap learning comparisons.

```{r pre-arrival-selection, results="asis"}
phase_selection_table(phase_selection) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, position = "H", linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.65in}",
              ">{\\raggedright\\arraybackslash}p{1.2in}", "r", "r", "r"),
    caption = "Pre-arrival knowledge differences: participants minus each comparison group. The climate study identifies completers; other invitees may have attended some sessions. The antimicrobial-resistance row pools six country samples without weights. N counts observed baseline scores; gaps are percentage points. Descriptive summaries have no estimated sampling intervals."
  ) |>
  print()
```

The baseline gaps show that attendees began with greater knowledge; they do not by themselves establish bias in the gain estimates. Comparing changes removes fixed differences in initial knowledge. An attendee-control comparison identifies the effect of participation if, without participation, the groups would have experienced the same average change. That assumption can fail if attendance is associated with preparation or learning outside the event.

## Participant and group characteristics

*Education*. We use the source-specific ordered categories of lower, middle, and higher education. These retain the reviewed groupings within each poll; they do not establish equivalent qualifications across countries. The lower category is the regression reference. We interact education with initial knowledge to allow their associations with gain to depend on one another.

*Gender and age*. Gender is coded 1 for women and 0 for men. Age enters in decades. Missing or refused responses remain missing.

*Group composition and mode*. Groupmates' initial knowledge and the proportion of women exclude the participant herself. Group size counts eligible participants with both pre-arrival and exit answers, so it is a lower bound on actual attendance when members lack an interview. Mode distinguishes online from in-person deliberation.

*Reading the briefing materials*. Respondent-linked reading reports are available for `r sum(coverage$reading > 0)` polls. They describe reading before discussion, recalled in the post-deliberation questionnaire. We score their ordered categories from zero to one using the source-specific response scales.

*Attitude extremity and disagreement*. We scale baseline policy measures to 0--1 using their response endpoints. The earlier polls retain their existing policy indices; the additional polls use individual policy preferences. Extremity is the participant's mean absolute distance from the midpoint, $E_i=J_i^{-1}\sum_j |A_{ij}-.5|$, over observed measures. Group disagreement is the average absolute difference between group members' responses:
$$H_g=\frac{1}{J_g}\sum_{j=1}^{J_g}\frac{2}{n_{gj}(n_{gj}-1)}\sum_{i<k}|A_{ij}-A_{kj}|.$$
Here $n_{gj}$ counts group members with an observed response to measure $j$, and $J_g$ counts measures with at least two such responses. A difference of .10 in $H_g$ means the average pairwise gap is ten percent of the response scale larger. Reversing a scale leaves both measures unchanged. Missing responses are excluded. Policy batteries differ in content and aggregation across polls, so these measures summarize attitudes toward each poll's own issues.

## Guessing and uncertainty

A correct answer can reflect knowledge or guessing. We apply the item model of @cor2016 separately to each poll's paired answers. Under the assumption that knowledge is not lost between interviews, it estimates the share who learned each fact. We use this adjustment only for the pre-arrival-to-exit comparisons, not for the later follow-ups. We average across questions to obtain learning on a 0--1 scale. In each control study, we fit this model separately to attendees and controls and subtract their estimated learning. Both arms use the same items; their guessing probabilities are estimated separately. Table \ref{tab:models} uses observed individual gains, since the item model estimates learning rates rather than an individual latent outcome.

We estimate uncertainty using a hierarchical bootstrap. For estimates across polls, we sample polls with replacement separately within online and in-person modes, retaining their observed counts, and then sample whole discussion groups within each selected poll. For an individual poll, we hold the poll fixed and sample its groups. Independently sampled controls are single-person units, sampled separately from attendees. Each sampled participant retains all items and interviews. We refit the models and recompute denominators in each of `r provenance$bootstrap_replicates` draws, using the 2.5th and 97.5th percentiles for 95\% confidence intervals. Thus the same resampling scheme propagates sampling and guessing-model estimation uncertainty. Stratifying by mode preserves support for the online coefficient in every draw. Inference treats each poll's question battery as fixed. The demographic gap comparisons use whole-poll resampling, described with Appendix Table \ref{tab:knowledge-gaps}.

# How Much Do Participants Learn?

An absolute gain is the change in the percentage of items answered correctly. A relative gain divides that change by the poll's initial mean score: an increase from 40\% to 50\% correct is 10 percentage points, or 25\% of the initial score. We calculate relative gains from poll means rather than averaging individual ratios, which are undefined for people with zero initial scores. The main pooled estimates give each poll equal weight; the relative result is the average of the poll-specific ratios, not the ratio of two pooled means.

Across the main sample, the average absolute gain is `r pp(pooled("raw, pooled")$estimate)` percentage points (95\% interval `r pp(pooled("raw, pooled")$lower)` to `r pp(pooled("raw, pooled")$upper)`). The average relative gain is `r pp(pooled("relative, pooled")$estimate)`\% (95\% interval `r pp(pooled("relative, pooled")$lower)` to `r pp(pooled("relative, pooled")$upper)`\%). Estimated learning after adjustment for guessing averages `r pp(pooled("guessing adjusted, pooled")$estimate)` percentage points (`r pp(pooled("guessing adjusted, pooled")$lower)` to `r pp(pooled("guessing adjusted, pooled")$upper)`). Precision weighting lowers the observed average to `r pp(pooled("raw, precision weighted")$estimate)` points. Gains are larger in person than online, averaging `r pp(meta$estimate[meta$model == "raw, by mode" & meta$parameter == "face_to_face"])` and `r pp(meta$estimate[meta$model == "raw, by mode" & meta$parameter == "online"])` points, respectively (Appendix Table \ref{tab:meta}). These mode differences may also reflect differences in participants, topics, and timing; they do not identify an effect of meeting in person.

A before-and-after gain can include learning from repeated questions or events outside deliberation. Figure \ref{fig:learning} therefore also compares attendees' gains with controls' gains. The observed differences are `r pp(control_gain("america-in-one-room-2019", "raw")$estimate)` percentage points in America in One Room and `r pp(control_gain("a1r-climate-2021", "raw")$estimate)` percentage points in the climate poll. Their equally weighted average is `r pp(control_average("raw")$estimate)` percentage points (95\% interval `r pp(control_average("raw")$lower)` to `r pp(control_average("raw")$upper)`), rising to `r pp(control_average("adjusted")$estimate)` percentage points after guessing adjustment (`r pp(control_average("adjusted")$lower)` to `r pp(control_average("adjusted")$upper)`). Both estimates use the same attendees, controls, and questions. These comparisons do not eliminate selection into attendance.

```{r learning, fig.cap = "Knowledge-score changes in percentage points, with 95\\% hierarchical-bootstrap confidence intervals. A: pre-arrival to exit changes. For online polls, pre-arrival is before the discussion sessions begin. The main sample requires observed questionnaires at the stages each poll collected through immediate exit. B: observed attendee gain minus control gain in the two polls with pre-arrival and exit interviews. Each panel ends with equally weighted observed and guessing-adjusted averages. Whole discussion groups are resampled within polls; independent controls are resampled individually. Pooled intervals resample polls within mode. With one control study per mode, the pooled control interval is conditional on these two studies. Each draw refits the guessing model, treating question batteries as fixed.", fig.pos = "tbp", out.width = "100%"}
figure("learning")
```

\FloatBarrier

## Changes before arrival and during the event {#sec:phases}

California, Europolis, Marousi, Michigan, and Tomorrow's Europe have comparable knowledge measures before arrival, on arrival, and after deliberation. All five show gains before arrival, ranging from `r pp(min(phase_balanced$estimate[phase_balanced$contrast == "arrival_minus_pre_arrival"]))` to `r pp(max(phase_balanced$estimate[phase_balanced$contrast == "arrival_minus_pre_arrival"]))` percentage points. Gains during the event range from `r pp(min(phase_balanced$estimate[phase_balanced$contrast == "post_minus_arrival"]))` to `r pp(max(phase_balanced$estimate[phase_balanced$contrast == "post_minus_arrival"]))` points; the intervals for Marousi and Michigan's four-item party-placement battery include zero. Table \ref{tab:phase-changes} uses the same respondents for all three changes within each poll, so the total equals the sum of its two component changes before rounding. Michigan is shown for both its full nine-item battery and the four party-placement questions common to all three interviews. Pair-specific samples, reported in Appendix \ref{app:phases}, answer a different question when questionnaire availability differs.

```{r phase-changes, results="asis"}
cat("\n```{=latex}\n")
phase_comparison_table(phase_balanced) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, escape = FALSE, position = "H",
    align = c(">{\\raggedright\\arraybackslash}p{1.4in}", "r", "r", "r", "r", "r"),
    col.names = c("Poll", "Items", "N", "\\shortstack[r]{Pre-arrival \\\\ to arrival}",
                  "\\shortstack[r]{Arrival \\\\ to exit}", "\\shortstack[r]{Pre-arrival \\\\ to exit}"),
    caption = "Knowledge changes among attendees with all three questionnaires and recorded discussion groups. Entries are percentage points; brackets contain 95 percent group-bootstrap intervals. Items reports the battery size where documented; a dash marks an unavailable count."
  ) |>
  print()
cat("\n```\n")
```

These estimates are unweighted changes in proportion correct, without adjustment for guessing. Intervals resample discussion groups in `r phase_provenance$bootstrap_replicates` draws, retaining each respondent's three scores together. The Marousi comparisons exclude `r prose_count(marousi_exit$n_excluded_unknown_group)` otherwise eligible respondents whose discussion groups are unknown. Its original merge also contains conflicting telephone and exit identifiers; these estimates therefore depend on the original matching of respondents across interviews, which remains unresolved.

Changes before arrival could reflect briefing, preparation, repeated testing, or other events between interviews. Arrival-to-exit changes isolate the event's time window, but do not identify deliberation's causal effect without a credible comparison. Even on a common sample, these estimates describe the attendees with observed questionnaires at all three phases; they do not describe all recruited citizens or all attendees.

## How much knowledge persists?

Later interviews allow us to ask how much of the gain remains after the event. We compare pre-arrival, exit, and follow-up scores among the same attendees answering the same questions at all three interviews. Table \ref{tab:retention} reports the National Issues Convention's follow-up about ten months after the event and the climate poll's follow-up about a year later [@luskinfishkin2005nic, p. 18; @fishkin2025scaling]. Neither follow-up enters the main pre-arrival-to-exit estimate.

```{r retention, results="asis"}
retention |>
  dplyr::mutate(
    Poll = dplyr::recode(poll_id,
      `nic-1996` = "National Issues Convention (ten months)",
      `a1r-climate-2021` = "America in One Room: Climate (one year)"
    ),
    value = ifelse(is.finite(lower),
      paste0(pp(estimate), " [", pp(lower), ", ", pp(upper), "]"), pp(estimate)
    )
  ) |>
  dplyr::select(Poll, N = n_people, contrast, value) |>
  tidyr::pivot_wider(names_from = contrast, values_from = value) |>
  dplyr::select(Poll, N, exit_minus_pre_arrival, followup_minus_exit, followup_minus_pre_arrival) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, escape = FALSE, position = "H",
    align = c(">{\\raggedright\\arraybackslash}p{1.65in}", "r", "r", "r", "r"),
    col.names = c("Poll and follow-up", "N",
      "\\shortstack[r]{Pre-arrival \\\\ to exit}",
      "\\shortstack[r]{Exit to \\\\ follow-up}",
      "\\shortstack[r]{Pre-arrival \\\\ to follow-up}"),
    caption = "Knowledge changes among attendees observed at all three interviews. Entries are percentage points; brackets contain 95 percent discussion-group-bootstrap intervals."
  ) |>
  print()
```

Knowledge remains above baseline at follow-up in both studies. Climate participants lose some of their initial gain after exit, whereas the National Issues Convention shows no clear decline. These comparisons describe returning attendees. Using the same people at each interview keeps changes in sample composition from being mistaken for changes in knowledge, but does not establish that those lost to follow-up had the same experience. Adjusting for observed predictors of follow-up response yields similar point estimates (Appendix \ref{app:attrition-adjustment}). Changes after exit can reflect forgetting, further learning, or both. We therefore use observed scores without the guessing model's no-forgetting assumption. In the climate poll, the baseline-adjusted attendee--control gap also remains positive a year later (Appendix \ref{app:controls}).

# Who Learns?

## Who gains more?

```{r models, results="asis"}
cat("\n```{=latex}\n")
table_1 <- model_table(models, c("core", "demographic", "attitudes", "briefing"), c("Baseline", "Demographics", "+ Attitudes", "+ Reading"))
table_1 <- stack_intervals(table_1)
separators <- rep("\\addlinespace[0.15em]", nrow(table_1))
summary_start <- match("Participants", table_1[[1]])
separators[summary_start - 1L] <- "\\midrule"
separators[summary_start:(nrow(table_1) - 1L)] <- "\\addlinespace[0.25em]"
table_tex <- knitr::kable(
  table_1, format = "latex", booktabs = TRUE, position = "!htb", align = "lrrrr", escape = FALSE,
  linesep = separators,
  caption = "Change in proportion correct (exit minus pre-arrival), conditional on initial knowledge. Coefficients with 95 percent hierarchical-bootstrap confidence intervals below."
)
table_tex <- sub("\\centering", "\\centering\\small\\setlength{\\tabcolsep}{3pt}", table_tex, fixed = TRUE)
cat(sub("\\end{table}", paste0("\\par\\medskip\\noindent{\\footnotesize\\raggedright\\emph{Notes.} Groupmates' mean initial knowledge excludes the respondent herself; both that mean and group size use members with initial and later scores. All models have poll and small-group random intercepts. Education uses each source's lower, middle, and higher ordered categories; the lower category is the reference. These categories do not represent identical qualifications across polls. Marginal $R^2$ covers fixed effects; conditional $R^2$ includes random intercepts. Columns 3 and 4 separately add attitudes and reading (0--1) to column 2. Reading reports are available for fewer polls. Replacing pairwise disagreement with the mean within-group SD gives a dispersion coefficient of ", term("attitude_sd", "attitude_sd"), "; the peer-knowledge coefficient is ", term("attitude_sd", "group_k1"), ".\\par}", "\n\\end{table}"), table_tex, fixed = TRUE))
cat("\n```\n")
```

Table \ref{tab:models} compares gains conditional on measured initial knowledge. The initial-knowledge coefficient in the baseline model is `r term("core", "k1")`. Because scores have a ceiling and contain measurement error, this should not be read as a causal effect of prior knowledge. The second column brings the education, age, and gender hypotheses into the same set of polls. The third column adds attitudes, and the fourth adds briefing reading. In both control studies, the attendee--control difference is larger among respondents who start with less knowledge. The climate poll also shows a larger difference among respondents without a bachelor's degree, while the corresponding education estimate in the 2019 poll is imprecise (Appendix Table \ref{tab:hetero}).

At the reference initial score of zero, both education categories have larger estimated gains than the lower-education group; the education-by-initial-knowledge interactions allow these gaps to vary with starting knowledge. The age estimate is `r term("demographic", "age_decades")` per decade. The intervals for age, gender, and the gender-by-group-composition interaction include zero. The gender coefficient compares women with men when the share of female groupmates is zero; the interaction shows how that difference changes as the share increases. The third column adds baseline attitude extremity and policy disagreement. On the same participants, adding these measures changes the peer-knowledge coefficient from `r coef_text(models$estimate[models$model == "demographic_attitude_sample" & models$term == "group_k1"])` to `r coef_text(models$estimate[models$model == "attitudes" & models$term == "group_k1"])`; the intervals for both added attitude measures include zero.

Reported reading of the briefing materials is also associated with larger gains. Across `r size("briefing", "polls")` polls, the difference between the lowest and highest reading categories is `r pp(reading$estimate)` percentage points (95\% interval `r pp(reading$lower)` to `r pp(reading$upper)`), conditional on initial knowledge, demographics, group characteristics, and mode, with poll and small-group random intercepts. These reports were collected after deliberation, so the association need not reflect the effect of reading alone.

## Do knowledge gaps narrow?

The gender gap narrows, but the educational gap does not show comparable convergence. On the same respondents before and after, the average difference between men and women declines from `r pp(gap("gender", "pre")$estimate)` to `r pp(gap("gender", "post")$estimate)` percentage points. The reduction is `r pp(-gap("gender", "change")$estimate)` points (95\% interval `r pp(-gap("gender", "change")$upper)` to `r pp(-gap("gender", "change")$lower)`). Women gain more on average, although a gap remains at exit.

For education, relative standing within a poll is also relevant: the same qualification can occupy different positions across countries and cohorts. We therefore compare participants above their poll's median education with those at or below it in the `r gap("relative_education", "change")$polls` polls with that definition. The gap increases from `r pp(gap("relative_education", "pre")$estimate)` to `r pp(gap("relative_education", "post")$estimate)` points, widening by `r pp(gap("relative_education", "change")$estimate)` points (95\% interval `r pp(gap("relative_education", "change")$lower)` to `r pp(gap("relative_education", "change")$upper)`). Using the source-specific ordered categories, the higher-versus-lower gap also widens. The higher-versus-middle gap changes little (Appendix Table \ref{tab:knowledge-gaps}). Each comparison gives eligible polls equal weight.

These are changes in observed gaps, not differences conditional on equal starting knowledge. Women begin with lower scores, so their larger average gain can coexist with the imprecise conditional gender association in Table \ref{tab:models}. The comparisons span pre-arrival to exit and include preparation as well as the event; they do not establish that deliberation caused the gaps to narrow or widen.

## Group composition and learning

Table \ref{tab:models} addresses the peer-knowledge hypothesis directly. A ten-percentage-point difference in groupmates' mean initial knowledge is associated with `r num(10 * models$estimate[models$model == "core" & models$term == "group_k1"], 2)` percentage points of additional gain in the baseline model and `r num(10 * models$estimate[models$model == "demographic" & models$term == "group_k1"], 2)` after demographic adjustment. The intervals in the table quantify uncertainty about these pooled associations.

The average of the separate within-poll peer-knowledge associations is also positive but imprecise: a ten-percentage-point difference in groupmates' initial knowledge is associated with `r num(10 * peer("k1")$estimate, 1)` points of additional gain (95\% interval `r num(10 * peer("k1")$lower, 1)` to `r num(10 * peer("k1")$upper, 1)`). Appendix Figure \ref{fig:peers} displays the estimates across polls. Its average gives each poll equal weight; Table \ref{tab:models} instead fits a common slope in a hierarchical model.

Appendix Table \ref{tab:baseline-composition} examines whether participants with similar baseline characteristics were grouped together. For initial knowledge, the association with groupmates' initial knowledge is `r interval(balance_knowledge$estimate, balance_knowledge$lower, balance_knowledge$upper)`. Its interval includes zero. This diagnostic concerns characteristics measured before discussion; later knowledge and gain can themselves respond to group composition. A causal interpretation of peer coefficients also requires evidence about the assignment process. Group knowledge after deliberation is difficult to interpret because groupmates share the same experience [@manski1993].

# Conclusion

Participants' factual knowledge increases between recruitment and the end of deliberation. Reporting absolute and relative gains makes the magnitude interpretable, while adjustment for guessing addresses one source of measurement error. In the two polls with common before-and-after control interviews, attendees show larger gains than controls. Interpreting the gain differences causally requires attendees and controls to have comparable changes without participation; the peer regressions are descriptive. The three-wave comparisons show that knowledge can increase before arrival as well as during the event. Generalizing beyond attendees requires assumptions about those who did not attend or complete later questionnaires. The knowledge questions concern facts selected by organizers, so these estimates describe learning about the topics the events emphasized.

Higher average knowledge does not imply that knowledge becomes more equally distributed. On average, the gender gap narrows without disappearing, while the gap between respondents classified as more and less educated within their poll widens. Deliberation and its surrounding preparation can therefore accompany both broader learning and persistent or increasing educational differences in knowledge.

The gains also need to be considered against how much room there was to learn. Giving each poll equal weight, participants answered `r pp(initial_correct)`\% of questions correctly at the initial interview and `r pp(later_correct)`\% at the later interview. The gain of `r pp(later_correct - initial_correct)` percentage points closed `r pp(headroom_closed)`\% of the initial `r pp(initial_headroom)`-point gap to a perfect score. At the later interview, `r pp(1 - later_correct)`\% of answers were still incorrect or blank. This is a comparison of average scores, not a count of previously unknown facts learned: correct answers can reflect guessing, and net gains can conceal both learning and forgetting.

Briefing materials, moderated discussions, and opportunities to question experts give participants unusually favorable conditions for learning about the issues on which they are tested. We cannot assume that every tested fact was discussed or that every participant read the relevant material. Still, substantial gaps remain despite this intensive support, suggesting that increasing factual knowledge is difficult even when citizens have time, information, and reason to attend to it. For democratic theory, these findings support the possibility of more informed public judgment while tempering expectations of comprehensive factual knowledge. A perfect test score is not a prerequisite for sound judgment, and factual recall does not exhaust what participants may learn about arguments, tradeoffs, or other people's experiences. The normative case for deliberation should therefore consider both the knowledge it adds and how well it helps citizens make decisions with the knowledge they have.

\FloatBarrier
# References {-}

\begingroup
\small

<div id="refs"></div>

\endgroup

\clearpage
\appendix
\counterwithin{table}{section}
\counterwithin{figure}{section}
\renewcommand{\thetable}{\thesection\arabic{table}}
\renewcommand{\thefigure}{\thesection\arabic{figure}}

# The Deliberative Polls {#app:polls}

These `r nrow(polls)` polls supply the main learning and discussion-group sample. The comparisons by interview timing use separate samples, described in Appendix \ref{app:phases}. Three also provide control comparisons, including Northern Ireland's T3-only contrast. Tables \ref{tab:polls} and \ref{tab:poll-gains} list polls in the same chronological order.

```{r polls, results = "asis"}
polls |>
  dplyr::left_join(main_interviews, by = "poll_id", relationship = "one-to-one") |>
  dplyr::transmute(
    poll, year,
    mode = dplyr::if_else(mode == "Online", "Online", "In person"),
    control = dplyr::if_else(control_group, "Yes", "No"),
    interviews = dplyr::if_else(is.na(phase_t1) | is.na(phase_t2), "Unverified",
                               paste(phase_t1, phase_t2, sep = " to "))
  ) |>
  purrr::set_names(c("Poll", "Year", "Mode", "Control", "Interviews")) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = "lclcc",
    caption = "Polls and interview pairs used in the main analysis. t0 is pre-arrival and t2 exit."
  ) |>
  print()
```

\clearpage
# Further results {#app:further}

```{r poll-gains}
gains |>
  dplyr::arrange(match(poll_id, polls$poll_id)) |>
  dplyr::transmute(
    Poll = pollname, `Initial (%)` = pp(k1_mean),
    `Gain (pp)` = pp(raw),
    `Adjusted (pp)` = pp(adjusted)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE,
    linesep = "", align = c(">{\\raggedright\\arraybackslash}p{2.8in}", "r", "r", "r"),
    caption = "Initial knowledge (percent correct), observed gains, and guessing-adjusted learning (percentage points) in the main sample. Interview timing and rules for including respondents with uncertain questionnaire completion vary by poll.")
```

\begin{minipage}{\textwidth}

```{r meta}
meta |>
  dplyr::filter(!grepl("SD|relative", model)) |>
  dplyr::mutate(
    Weighting = dplyr::if_else(grepl("precision weighted", model), "Precision", "Equal"),
    Quantity = dplyr::case_when(
      parameter == "face_to_face" ~ "Observed gain: in person (pp)",
      parameter == "online" ~ "Observed gain: online (pp)",
      grepl("guessing", model) ~ "Guessing-adjusted learning (pp)",
      TRUE ~ "Observed gain (pp)"
    )
  ) |>
  dplyr::transmute(Quantity, Weighting, Polls = polls, Estimate = pp(estimate),
                   `95% interval` = paste(pp(lower), pp(upper), sep = " to ")) |>
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "",
    position = "H", caption = "Average learning across polls, with hierarchical-bootstrap confidence intervals.")
```

Observed gain is the exit score minus the pre-arrival score; guessing-adjusted learning estimates the percentage who learned an item, averaged across questions. Equal weighting gives each poll the same weight. Precision weighting uses $w_d=1/\widehat{\mathrm{Var}}(\hat\theta_d)$, with sampling variances estimated by resampling discussion groups within each poll. The weighted average is $\sum_d w_d\hat\theta_d/\sum_d w_d$. These weights are held fixed when resampling polls and groups for its interval. The mode rows split the equal-poll average into in-person and online polls.

\end{minipage}

## Sensitivity to zero knowledge scores {#app:zero-scores}

A zero score can reflect incorrect answers, explicit “don't know” responses, or an entirely blank knowledge section on an otherwise completed questionnaire. We retain all of these cases in the main analysis. Table \ref{tab:zero-scores} separately removes people with a zero score at either the pre-arrival or exit interview, and people whose entire knowledge section is blank at either interview. Explicit “don't know” responses do not count as blanks; uncertain blankness remains in the sample. These exclusions change the population being described; they are sensitivity checks, not alternative diagnoses of nonparticipation.

```{r zero-scores, results="asis"}
sensitivity |>
  dplyr::filter(term %in% c("k1", "group_k1", "online")) |>
  dplyr::mutate(
    Sample = dplyr::recode(sample, main = "Main", zero_score = "Exclude zero scores", all_blank = "Exclude blank batteries"),
    Term = dplyr::recode(term, k1 = "Initial knowledge", group_k1 = "Groupmates' mean initial", online = "Online"),
    cell = interval(estimate, lower, upper)
  ) |>
  dplyr::mutate(N = count(n)) |>
  dplyr::select(Sample, N, Term, cell) |>
  tidyr::pivot_wider(names_from = Term, values_from = cell) |>
  stack_intervals() |>
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "", position = "H", escape = FALSE,
    caption = "Baseline model under alternative treatments of zero scores. Entries show coefficients and 95 percent intervals.",
    label = "zero-scores") |>
  cat()
```

Excluding zero scores lowers the estimated peer-knowledge association; excluding wholly blank batteries changes it little. All three intervals include zero.

All three specifications use the baseline model in Table \ref{tab:models}. Group size and groupmates' mean knowledge are recalculated for each sample. N counts respondents in the grouped model. Intervals use `r provenance$bootstrap_replicates` hierarchical bootstrap draws with the same procedure as the main analysis.

```{r knowledge-gaps, results="asis"}
gaps |>
  dplyr::mutate(value = ifelse(term == "change",
                              paste0(pp(estimate), " [", pp(lower), ", ", pp(upper), "]"), pp(estimate))) |>
  dplyr::select(comparison, polls, term, value) |>
  tidyr::pivot_wider(names_from = term, values_from = value) |>
  dplyr::transmute(
    Comparison = dplyr::recode(comparison,
      gender = "Men minus women",
      relative_education = "Poll-specific higher minus lower education",
      higher_vs_lower = "Higher minus lower education",
      higher_vs_middle = "Higher minus middle education"),
    Polls = polls, `Pre-arrival` = pre, Exit = post, `Change [95% interval]` = change
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "",
               align = c(">{\\raggedright\\arraybackslash}p{2.1in}", "r", "r", "r", "r"),
               caption = "Knowledge gaps before and after, in percentage points. Negative changes indicate narrowing. Intervals resample whole polls within mode, preserving paired interviews.") |>
  print()
```

These comparisons use the demographic-model sample and calculate each gap within a poll before averaging equally across polls. A poll enters a comparison only if both categories are observed. The relative-education comparison uses each poll's reviewed ordered education measure. The median is calculated among unique historical participants in each of the `r gap("relative_education", "change")$polls` polls with this measure; we compare participants above the median with those at or below it. Tied categories remain together, so the two groups need not be equally large. Table \ref{tab:models} uses three education categories. Europolis has no respondents in the lower category in this sample, so the higher-versus-lower comparison covers `r gap("higher_vs_lower", "change")$polls` polls; the gender and higher-versus-middle comparisons cover `r gap("gender", "change")$polls`. Some education categories contain few respondents within individual polls. For these gap comparisons, we resample whole polls within mode in `r gap("gender", "change")$bootstrap_replicates` draws, retaining all respondents and both interviews together. The interval for the change is calculated from paired changes in each draw, not by subtracting the endpoints of separate intervals. These comparisons use the same scoring and questionnaire-completion rules as the main analysis.

\clearpage

```{r hetero}
hetero |>
  dplyr::mutate(
    comparison = dplyr::if_else(moderator == "k1", "Initial score +10 pp", "Bachelor’s degree"),
    multiplier = dplyr::if_else(moderator == "k1", 10, 100)
  ) |>
  dplyr::transmute(Study = study, Comparison = comparison,
    `Difference (pp)` = num(multiplier * interaction, 1),
    `95% interval` = paste(num(multiplier * lower, 1), num(multiplier * upper, 1), sep = " to "),
    N = count(n)) |>
  knitr::kable(format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.8in}", ">{\\raggedright\\arraybackslash}p{1.35in}", ">{\\raggedleft\\arraybackslash}p{0.75in}",
      ">{\\raggedleft\\arraybackslash}p{0.85in}", ">{\\raggedleft\\arraybackslash}p{0.7in}"),
    caption = "How the baseline-adjusted attendee--control difference varies with initial knowledge and education.")
```

Table \ref{tab:hetero} reports changes in the attendee--control difference, not changes in everyone's knowledge. N is the regression sample size. Degree compares a bachelor’s degree with no degree. A negative entry means the attendee advantage is smaller for people with higher initial knowledge or a bachelor’s degree. For example, the America in One Room initial-knowledge estimate of `r num(10 * het("America in One Room 2019", "k1")$interaction, 1)` percentage points compares two people whose starting scores differ by ten percentage points. Each regression includes attendance, initial score, the listed characteristic, and its interaction with attendance. Centering initial scores at their sample mean does not change the reported interaction. All intervals use the hierarchical bootstrap. Selection into attendance and score ceilings limit causal interpretation.

## Groupmates' knowledge of missed questions {#app:item-peers}

The demographic specification can also include groupmates' initial correctness on the particular questions the participant missed at the initial interview. For each missed question, we calculate the proportion of other group members who answered correctly, then average across those questions. This targeted measure enters alongside groupmates' overall initial knowledge. Respondents who answered every initial question correctly receive zero on the targeted measure and a separate indicator for having no missed questions.

The targeted peer coefficient is `r term("items", "group_k1_items")` among `r count(size("items", "n"))` participants in `r size("items", "polls")` polls. The coefficient on overall peer knowledge is `r term("items", "group_k1")`, and the no-missed-questions indicator is `r term("items", "no_missed_items")`. All other regressors, random intercepts, and the bootstrap follow Table \ref{tab:models}.

## Additional control comparisons {#app:controls}

Regressing later scores on attendance and initial scores answers a different question from comparing gains. The baseline-adjusted attendee--control differences are `r pp(a1r$estimate)` percentage points in America in One Room (`r pp(a1r$lower)` to `r pp(a1r$upper)`), and `r pp(climate$estimate)` in the climate poll (`r pp(climate$lower)` to `r pp(climate$upper)`). These intervals use the same bootstrap as the main comparisons.

A year after the climate poll, the baseline-adjusted gap is `r pp(climate_t3$estimate)` percentage points (`r pp(climate_t3$lower)` to `r pp(climate_t3$upper)`). Northern Ireland has only a follow-up control interview: `r ni$n_treated` returning attendees scored `r pp(ni$estimate)` percentage points above `r ni$n_control` controls (`r pp(ni$lower)` to `r pp(ni$upper)`). This is a cross-sectional comparison without a control baseline. We do not apply the main guessing model's no-forgetting assumption to these later outcomes.

## Poll-specific peer associations

The following estimates fit a separate slope in each poll, conditional on own initial knowledge. Their equal-poll average is `r interval(peer("k1")$estimate, peer("k1")$lower, peer("k1")$upper)`.

```{r peers, fig.cap = "Association between the other group members' mean initial knowledge and own knowledge gain, by poll, with 95\\% intervals. Each estimate controls for own initial knowledge. Intervals use the hierarchical bootstrap. The pooled estimate weights polls equally and resamples polls within mode and then their groups. These are descriptive associations.", fig.pos = "tbp", out.width = "100%"}
figure("peer_effects")
```

## Baseline composition of discussion groups

We regress each participant's baseline characteristic on its mean among the other group members, controlling for the mean among other members of the poll and poll fixed effects [@guryan2009]. This is a descriptive sorting diagnostic; it does not reconstruct the original assignment mechanism. A randomization test would require the original assignment pools and restrictions, including any scheduling or language constraints. Balance on observed covariates alone does not establish random assignment.

```{r baseline-composition}
assignment |>
  dplyr::transmute(
    Characteristic = dplyr::recode(covariate, k1 = "Initial knowledge", female = "Female", age = "Age", education_higher = "Higher education category"),
    Association = interval(estimate, lower, upper), Participants = count(n)
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "", caption = "Baseline composition of discussion groups. Associations with 95 percent hierarchical-bootstrap intervals.")
```

\clearpage
# Interview timing and analysis samples {#app:phases}

Questionnaire completion creates an additional selection stage. Appendix Table \ref{tab:phase-attrition} compares baseline knowledge among known attendees with observed, absent, or uncertain exit questionnaires. A blank quiz within an observed questionnaire scores zero; a wholly absent questionnaire remains unmeasured. Respondents whose questionnaire completion is uncertain are excluded from the paired comparisons. Those whose attendance is unknown remain in the baseline summaries and sample counts, with attendance recorded as unknown. Attendance classifications use source indicators, session records, onsite questionnaires, and verified historical identity links. For the harmonized participation measure, a wholly unanswered immediate post questionnaire implies nonattendance even when a roster or session record indicates attendance; the original indicator remains available. Main analyses also require observed questionnaires at earlier stages that the poll collected. A stage that was never collected does not exclude anyone. Cases without sufficient source evidence remain unknown. Uncertain attendance does not establish that a respondent was absent, and missing scores do not establish that a poll lacked an arrival survey.

Table \ref{tab:phase-pairs} compares knowledge among attendees who answered both questionnaires in each pair of interviews. These estimates are unweighted changes in proportion correct without guessing adjustment. The overlapping historical Primaries subset is omitted in favor of the attendee sample used in the main analysis. We use one source copy where original respondent identities, answers, scores, attendance, and discussion groups agree. We retain comparisons with different question batteries or respondent samples. For Tomorrow's Europe, Archive and Cor–Sood replication identify overlapping source versions whose discussion-group assignments differ; their estimates are not independent polls. Each row requires observed questionnaires and the same battery and denominator at both phases; it does not require the third phase or discussion-group membership. Thus the three rows for a poll need not use the same people and need not add up. Table \ref{tab:phase-changes} instead restricts all three changes to the common three-wave sample with recorded discussion groups. The Items column distinguishes separate batteries: California has five common questions and an expanded eight-question arrival/exit battery; Europolis has six and nine, respectively. Michigan has nine questions across all three interviews, including four party-placement questions; a separate battery has six party placements at arrival and exit. We report descriptive estimates without intervals when discussion-group membership is unknown or too few groups are available to estimate uncertainty.

```{r phase-pairs, results="asis"}
phase_pair_table(phase_pairs) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{1.65in}",
              ">{\\raggedright\\arraybackslash}p{1.45in}", "r", "r", "r", "r"),
    caption = "Knowledge changes among attendees who answered both questionnaires. Changes and 95 percent group-bootstrap intervals are percentage points. Items reports battery size; a dash means an unavailable count. A missing interval means groups are unknown or too few."
  ) |>
  print()
```

Table \ref{tab:phase-attrition} reports exit-questionnaire availability and baseline knowledge for Marousi, Tomorrow's Europe, and the two control studies. An absent exit requires evidence of a wholly absent questionnaire; an uncertain exit is not recoded as absent. Baseline means use the scored records counted in the last-but-one column. The groups shown exhaust known attendees in these source frames, including people excluded from paired-change estimates.

```{r phase-attrition, results="asis"}
phase_attrition_table(phase_selection) |>
  knitr::kable(
    format = "latex", booktabs = TRUE, longtable = TRUE, linesep = "",
    align = c(">{\\raggedright\\arraybackslash}p{2.0in}",
              ">{\\raggedright\\arraybackslash}p{1.05in}", "r", "r", "r"),
    caption = "Exit availability among known attendees: sample counts and pre-arrival knowledge (percent correct)."
  ) |>
  print()
```

Marousi's full telephone frame contains `r count(phase_coverage$n_people[phase_coverage$poll_id == "marousi-2006" & phase_coverage$phase == "t0"])` records. The source attendance indicators and questionnaire records are retained to study selection and attrition; the harmonized attendance measure follows the completion rule above. An absent arrival or exit form does not establish whether someone was invited or declined an invitation. Some Cor–Sood cohorts retain unresolved attendance, recorded separately from questionnaire presence. Matching respondent numbers in separate datasets are not sufficient to establish that records describe the same person.

## Adjusting for missing interviews {#app:attrition-adjustment}

The comparisons above describe people who answered the relevant questionnaires. If returning respondents learned more or less than other members of the original source cohort, their gains need not describe that larger cohort. We examine this possibility in Marousi, the National Issues Convention, and the climate poll. In Marousi, the target is the original source cohort with positive attendance evidence and a pre-arrival score. This includes `r with(subset(attrition_flow, poll_id == "marousi-2006" & phase == "t2"), n_target - n_target_observed)` people without an exit questionnaire who remain nonparticipants under the main analysis's completion rule. Defining this attrition target before requiring an exit questionnaire avoids excluding the missing outcomes we seek to assess. For the two follow-up studies, the target is attendees with both pre-arrival and exit scores, including those missing at follow-up. These latter comparisons address retention among attendees whose initial gain is observed; they do not adjust for missing baseline or exit interviews.

```{r attrition-cohorts, results="asis"}
attrition_flow |>
  dplyr::select(pollname, phase, n_target, n_target_observed) |>
  tidyr::pivot_wider(names_from = phase, values_from = n_target_observed) |>
  dplyr::transmute(Poll = pollname, `Target N` = n_target,
                   `Pre-arrival` = t0, Arrival = t1, Exit = t2, `Follow-up` = t3) |>
  dplyr::mutate(dplyr::across(-Poll, ~ifelse(is.na(.x), "--", as.character(.x)))) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "",
               align = c(">{\\raggedright\\arraybackslash}p{1.7in}", rep("r", 5)),
               caption = "Observed knowledge scores within each target cohort. A dash denotes a phase outside that comparison.") |>
  print()
```

Missing interviews are treated separately at each stage. A missed arrival interview does not imply that the exit interview is missing. People with unknown attendance are not classified as nonattenders, and an absent questionnaire is not assigned a knowledge score of zero.

Within each poll, we estimate the probability of observing a score using logistic regression and its conditional mean using fractional logistic regression. We combine the two predictions in an augmented inverse-probability-weighted estimate [@seaman2018]. Let $R_{it}$ indicate an observed score at stage $t$, $\hat p_t(X_i)$ its estimated response probability, and $\hat m_t(X_i)$ its predicted value. The estimated mean for the target cohort is

$$
\widehat{\mu}_t = \frac{1}{N}\sum_{i=1}^{N}\left[\hat m_t(X_i) +
\frac{R_{it}}{\hat p_t(X_i)}\{K_{it}-\hat m_t(X_i)\}\right],
$$

where the residual term is zero when the score is unobserved. We calculate changes by subtracting stage means for the same target cohort, so the component changes sum to the total. Table \ref{tab:attrition-adjusted} compares these estimates with complete-case changes and normalized inverse-probability-weighted changes.

The Marousi models use pre-arrival knowledge. The National Issues Convention models use pre-arrival and exit knowledge. The climate models also include age, a bachelor's-degree indicator, and gender. We use demographics only where they are recorded for the entire target cohort, avoiding further exclusions or demographic imputation. All models are fitted separately by poll, with the same predictors in the response and outcome models.

```{r attrition-adjusted, results="asis"}
attrition_changes |>
  dplyr::mutate(Comparison = dplyr::case_when(
    from == "t0" & to == "t1" ~ "Pre-arrival to arrival",
    from == "t1" & to == "t2" ~ "Arrival to exit",
    from == "t0" & to == "t2" ~ "Pre-arrival to exit",
    from == "t2" & to == "t3" ~ "Exit to follow-up",
    TRUE ~ "Pre-arrival to follow-up"
  )) |>
  dplyr::transmute(Poll = pollname, Comparison,
                   `Complete cases` = pp(complete_case), Weighted = pp(ipw), Augmented = pp(aipw)) |>
  knitr::kable(format = "latex", booktabs = TRUE, position = "H", linesep = "",
               align = c(">{\\raggedright\\arraybackslash}p{1.6in}",
                         ">{\\raggedright\\arraybackslash}p{1.35in}", "r", "r", "r"),
               caption = "Knowledge changes before and after adjustment for interview nonresponse, in percentage points. Entries are point estimates; this comparison does not estimate sampling intervals.") |>
  print()
```

The Marousi complete-case column includes `r attrition_changes$n_complete[attrition_changes$poll_id == "marousi-2006"][1]` attendees, including the `r prose_count(marousi_exit$n_excluded_unknown_group)` with unknown groups excluded from Table \ref{tab:phase-changes}. The adjusted estimates target all `r attrition_changes$n_target[attrition_changes$poll_id == "marousi-2006"][1]` baseline-observed members of the original source cohort. In the follow-up studies, the complete-case samples match Table \ref{tab:retention}, while adjustment targets the larger cohorts in Table \ref{tab:attrition-cohorts}.

The fitted response probabilities are no lower than `r pp(min(attrition_means$min_probability))`\%, and no respondent receives an inverse-probability weight above `r num(max(attrition_means$max_weight), 2)`. We do not trim weights. Adjustment changes these point estimates by at most `r pp(max(abs(attrition_changes$aipw - attrition_changes$complete_case)))` percentage points. This similarity concerns selection on the measured predictors. Identification still requires that, conditional on those predictors, respondents and nonrespondents have the same expected knowledge and a positive chance of response. Under that assumption, the augmented estimator is consistent if either the response model or the outcome model is correctly specified. It does not remove selection on unmeasured knowledge, establish a causal effect of attendance, or estimate what nonattenders would have learned.

\clearpage
# Knowledge items {#app:items}

The catalog lists `r nrow(item_catalog)` questions for the same `r nrow(gains)` polls used in the main learning and group-model analysis [@dpdata]. Wording is paraphrased where only source labels survive; capitalization and unambiguous abbreviations are normalized for readability. Original labels and scoring rules are preserved in the cited dataset. Choices are transcribed where recoverable; truncated or unavailable options are identified. Notes explain open-response coding.

```{r items, results='asis'}
cat(item_appendix_markdown(items = item_catalog))
```
